Syntaxin11 Deficiency Inhibits CRAC Channel Priming to Suppress Cytotoxicity and Gene Expression in T Lymphocytes

Version 1

Abstract

Mutations in Syntaxin11, a Q-SNARE, result in a fatal immune disorder known as familial hemophagocytic lymphohistiocytosis 4 (FHLH4) in human patients. A key diagnostic feature of FHLH4 is defective T and natural killer (NK) cell cytotoxicity. Here we show that Syntaxin11 directly binds and regulates Orai1, the pore forming subunit of calcium release activated calcium (CRAC) channels. CRAC channels enable store-operated calcium entry (SOCE) from the extracellular space and are crucial for granule exocytosis and nuclear factor of activated T cells (NFAT) dependent gene expression in activated lymphocytes. Syntaxin11 depletion strongly inhibited SOCE, CRAC currents, NFAT activation, interleukin-2 gene expression and degranulation in FHLH4 patient T lymphocytes and cell lines. Remarkably, defects of granule exocytosis as well as interleukin-2 expression could be reversed by ionomycin in patient T lymphocytes and a constitutively active, H134S, mutant of Orai1 rescued calcium entry in Syntaxin11 depleted cells. Further analyses showed that Syntaxin11 primes Orai1 for optimal on-site multimeric assembly which was Stim independent but required for gating. Priming of ion channel pore subunits is, therefore, a primary function of specific SNAREs which may have preceded their role in vesicle fusion.

Introduction

Familial hemophagocytic lymphohistiocytosis 4 (FHLH4) is a life-threatening immune disease caused due to mutations of STX11 [1]. The underlying cause of FHLH is reduced cytolytic activity of T and NK cells [2], which renders patients susceptible to recurrent infections. Patients suffer from high fever, severe lymphopenia in early infancy and succumb to disease by adolescence, unless given bone marrow transplants. STX11 is highly expressed in almost all cells of the immune system and, being a Q-SNARE, is, by default, associated with exocytosis of vesicles. Intriguingly, FHLH4 patients also manifest cytokine storm, which cannot be reconciled with a presumed general defect in membrane trafficking and vesicle fusion.

Most eukaryotic cells have a limited amount of calcium sequestered inside intracellular stores, the largest of which is the endoplasmic reticulum (ER). Signaling from cell surface receptors induces the release of stored calcium which, in turn, activates store-operated calcium entry (SOCE) to replenish stores, sustain signaling and drive other calcium dependent cellular processes [3]. Calcium release activated calcium (CRAC) channels are the major conductors of SOCE [4]. In lymphocytes and mast cells, CRAC currents have been shown to be crucial for cytotoxicity, degranulation as well as gene expression [5, 6]. Working in concert with a number of accessory proteins, Orai (CRACM) multimers conduct SOCE upon functional clustering with Stim proteins in ER-PM junctions [7–9]. Yet, the prevailing thought attributes all structural transitions that result in the activation of CRAC currents to Stim proteins [10].

Genome-wide RNAi screens, which initially identified Orai (CRACM), have also yielded a wealth of information regarding additional crucial players in SOCE [8, 11, 12]. For instance, we have previously shown that alpha-soluble N-ethylmaleimide sensitive factor (NSF) attachment protein (α-SNAP), a well-known synaptic family adaptor protein that forms a part of the 20S SNARE super-complex [13], is a crucial component of the CRAC channel supramolecular complex [8, 9]. α-SNAP independently binds Stim as well as Orai and is required for the correct assembly and calcium selectivity of CRAC channels [14–16]. α-SNAP depletion in cells revealed that Stim:Orai coupling is necessary but not sufficient for SOCE, a process that requires distinct time-resolved molecular steps in vivo.

In addition to α-SNAP, SNAREs also came up as candidates in two genome wide RNAi screens for regulators of SOCE but were not characterized [8, 12]. Furthermore, specific SNAREs have been previously found to form inhibitory associations with specific ion channels, but their exact role was not defined [17]. Therefore, we hypothesized a more fundamental function for the association of SNAREs with ion channel pore subunits and set out to test this in the context of SOCE. Using a targeted RNAi screen, biochemical and biophysical analyses, we have found a direct role of STX11 in facilitating assembly of CRAC channel pore subunit, Orai1, into optimal multimers. Assembly of Orai1 was independent of, and preceded, trapping and gating by Stim1. Analyses of T cells derived from symptomatic STX11 deficient, FHLH4 patient showed that the previously reported [2], as well as newly identified, defects in degranulation and interleukin-2 gene expression predominantly stem from suppressed SOCE. Based on the findings presented here, we demonstrate a novel SNARE dependent priming step for CRAC channel pore in the molecular sequence of SOCE and suggest preventive pharmacological interventions for FHLH4 patients.

Results

RNAi dependent ablation of a Q-SNARE, Syntaxin 5 (STX5), has earlier shown inhibition of SOCE in a genome-wide screen performed in Drosophila S2 cells [12] and STX1A has been previously shown to bind a variety of ion channels [17]. Therefore, we first knocked down STX5 (Supplementary Figure 1A) and STX1A (Supplementary Figure 1B) using different sequences of shRNA targeting each gene in HEK293 cells and measured SOCE in response to Thapsigargin (TG). In both cases, we did not see a significant defect in SOCE. Since several proteins involved in synaptic function are not ubiquitously expressed and many show redundancy in function, we searched online databases to target SNAREs that are expressed in HEK293 and/or T cells (Supplementary Figure 1C-H). We found that Syntaxin11 knockdown showed strong and consistent inhibition of SOCE by day 3 to 4 (Supplementary Figure 2A-B) (Figure 1A-B). Quantitative PCR (qPCR) analysis of total RNA extracted from scramble (scr) and STX11 shRNA treated cells showed nearly 70% depletion of STX11 mRNA (Supplementary Figure 2C) and ectopic expression of STX11 in STX11 depleted Jurkat cells rescued SOCE (Figure 1C,D). The efficiency of STX11 depletion was assessed for each experiment reported in this study. We restricted our analysis to day 3 to 4, post RNAi treatment, as longer incubations with RNAi or CRISPR knockout of STX11 affected cell viability.

STX11 is required for SOCE.

Measurement of SOCE in STX11 depleted Jurkat and FHLH4 patient T cells. (A-B) Representative Fura-2 calcium imaging assay (A) and quantification of independent repeats (B) measuring thapsigargin (TG) induced SOCE in Jurkat T cells treated with scr (black), STX11#1 (red) or STX11#2 (blue) shRNAs. (n=30-40 cells/group, N=3). (C-D) Representative Fura-2 calcium imaging assay (C) and quantification of repeats (D) showing reconstitution of SOCE in STX11 depleted Jurkat T cells by ectopic expression of STX11. scr shRNA with empty vector (EV) expression (black), STX11 shRNA with EV expression (red), STX11 shRNA with STX11 expression (green). (E) Sanger sequencing of FHLH4 patient DNA showing deletion of a single Adenine at the 752nd position and the resulting frameshift. (F) Schematic showing the predicted domain distribution of the wildtype versus FHLH4 mutant STX11 protein. (G) Western blot on WCLs of healthy donor (HD) and FHLH4 patient PBMCs showing the relative molecular weight and abundance of the wildtype and mutant STX11 bands. (H) Western blot on WCLs of HEK293 cells over-expressing wildtype or FHLH4 mutant STX11 and empty vector. (I) Fura-2 calcium imaging assay measuring anti-CD3 induced SOCE in HD (black) and FHLH4 (red) T cells. (n=30-40 cells/group). (J) Fura-2 calcium imaging assay measuring SOCE in HD expressing empty vector (EV) (black) and FHLH4 T cells expressing human STX11 (red), respectively. (n=30-40 cells/group). (K-L) Representative Fura-2 calcium imaging assay (K) and quantification of repeats (L) measuring SOCE in Jurkat T cells treated either with scr shRNA expressing empty vector (EV) (black), STX11 shRNA, expressing EV (red), STX11 shRNA, expressing wildtype STX11 (green), STX11 shRNA, expressing FHLH4 mutant STX11 (blue).

Given that STX11 is highly expressed in most immune cells and is the only known SNARE that, when mutated, causes defects in T and NK cell cytotoxicity in humans and mice [2], we hypothesized that reduced SOCE contributes to the cytotoxicity defects of FHLH4 patient T cells. To test this hypothesis, we isolated PBMCs from an FHLH4 patient with a homozygous deletion frameshift mutation in the STX11 coding sequence, c.752delA:p.Lys251fs. The patient presented with all symptoms of FHLH4 disease by 4 years of age. Given the limited sample amount, we expanded PBMC numbers by periodically stimulating patient and healthy donor cells with PHA and IL-2, prior to analysis. Isolation and Sanger sequencing of DNA from patient PBMCs showed deletion of a single Adenine at position 752 in the STX11 gene which lead to frameshift following Lysine 251 (Figure 1E). The absence of any contaminating wildtype peaks showed that the mutation was present in all blood cells. Figure 1F shows schematic of the predicted mutant STX11 protein with an estimated molecular weight of ∼39.5 KDa compared to ∼33 KDa of wildtype STX11. The frameshift mutation resulted in disruption of the SNARE domain, ablation of the terminal cysteines and elongation of the resulting transcript. We subjected whole cell lysates prepared from the FHLH4 patient and healthy donor (HD) PBMCs to SDS-PAGE and western blot and found that while the wildtype STX11 band was absent, a faint but distinct band running at ∼37 KDa was present in the FHLH4 patient sample lysate (Figure 1G). Therefore, the predicted and observed molecular weight of the FHLH4 mutant protein was higher but its levels appeared significantly reduced compared to STX11 from HD. This is likely due to protein instability and degradation because ectopic expression of the FHLH4 STX11 mutant in HEK293 cells showed at least two clear bands (Figure 1H), although reduced reactivity to anti-STX11 antibody is also a possibility. In accordance with our findings in the STX11 depleted cell lines (Figure 1A-B and Supplementary Figure 2A-B), SOCE was found to be significantly defective in FHLH4 T cells when compared to HD (Figure 1I). Further, expression of wildtype (Figure 1J), but not FHLH4 mutant STX11 (Figure 1K-L), reversed defective SOCE in STX11 deficient patient (Figure 1J) and Jurkat T cells (Figure 1K-L), respectively. These data suggest that FHLH4 mutant STX11 was functionally compromised in addition to being unstable and no additional unknown mutations resulted in reduced SOCE.

To determine whether the inhibition of SOCE in Figure 1 resulted from changes in the activity of CRAC channels or their expression level, we quantified the level of Orai1 expression in plasma membrane. To do this, we labelled a previously characterized U2OS cell line stably expressing Orai1 tagged with a bungarotoxin binding site (BBS) in the extracellular loop and YFP in the C-terminus (Orai-BBS-YFP) with Alexa-647 conjugated alpha-bungarotoxin (BTX-A647) (Supplementary Figures 3A) [9]. There was no change in the expression of total Orai1 in the PM of STX11 depleted U2OS cells (Supplementary Figure 3B), even though SOCE was significantly suppressed (Supplementary Figures 3C-D). The ER-targeted CEPIA expressing control and STX11 depleted cells showed comparable ER calcium content (Supplementary Figure 3E). Labelling STX11 depleted cells with ER, Golgi organelle specific markers also showed no abnormalities (Supplementary Figures 3F-G) suggesting that the overall health, function and calcium content of Golgi and ER were not adversely affected by STX11 depletion.

Previous studies have shown that SOCE in T cells is mediated by CRAC channels formed by Orai proteins. To test whether CRAC current (ICRAC) was affected by knockdown of STX11 in T cells, we performed whole cell patch clamp recordings on scramble and STX11 shRNA treated Jurkat T cells (Figure 2A-B). Leak-subtracted ICRAC was recorded in 20 mM Ca2+ and a Na+-based divalent cation free (DVF) solution. These recordings showed that both the Ca2+ and DVF CRAC currents were significantly decreased in the STX11 shRNA treated cells with the normalized currents shown in Figure 2C. The electrophysiological properties of the residual current in STX11 shRNA treated cells was indistinguishable from ICRAC in scramble shRNA treated cells in terms of blockade of the Ca2+ current by La3+ and depotentiation of the DVF current over tens of seconds. Moreover, the reversal potential of the Ca2+ and DVF currents were similar between scramble and STX11 shRNA treated Jurkat T cells, suggesting that STX11 depletion does not affect the calcium selectivity of CRAC channels (Figure 2D). Together, these results demonstrate that STX11 works by specifically modulating ICRAC in T cells.

STX11 regulates ICRAC.

(A-D) Measurement of ICRAC in STX11 depleted Jurkat T-cells in the whole-cell recording configuration in 20 mM extracellular Ca2+ Ringer’s solution. ICRAC was induced by passive depletion of intracellular Ca2+ stores by dialyzing 8 mM BAPTA into the cell via the patch-pipette. (A) Representative current at −100 mV in Jurkat T cells treated with scr shRNA construct. The current is blocked by extracellular La3+ (10 µM) and replacing the 20 mM Ca2+ Ringer’s solution with a divalent free solution (DVF) evokes a large Na+ current which depotentiates over tens of seconds. The current-voltage (I-V) relationship of the Ca2+ and DVF currents are shown on the right. (B) ICRAC from a Jurkat T cell treated with STX11 shRNA. Both Ca2+ and Na+ current amplitudes are reduced relative to control cells. The I-V relationships (right plots) show no change in ion selectivity. (C-D) Summary of the current amplitudes of Ca2+ and Na+ currents and current reversal potentials in scr and STX11 knockdown cells.

CRAC channel mediated sustained calcium influx is crucial for the activation of a key calcium dependent transcription factor known as nuclear factor of activated T cells (NFAT) [4, 18]. We stimulated control and STX11 depleted Jurkat T cells with Thapsigargin (TG)+PMA, prepared nuclear, cytosolic extracts and subjected them to western blot using anti-NFAT1 antibody (Figure 3A). Nuclear translocation of NFAT was severely compromised in STX11 depleted T cells. In addition, immunofluorescence imaging and quantification of total versus nuclear NFAT from anti-CD3 stimulated Jurkat cells showed a defect in nuclear translocation of NFAT1 in STX11 depleted cells (Figure 3B-C). To determine whether NFAT induced gene expression was affected [4, 18], we quantified IL-2 mRNA by qPCR and secreted IL-2 levels from culture supernatants of anti-CD3 stimulated Jurkat T cells. STX11 depleted cells showed a significant decrease in IL-2 gene expression as well as secretion (Figure 3D, 3E). We stimulated HD and FHLH4 patient T cells either through T cell receptor using anti-CD3 or with phorbol myristate acetate (PMA) + Ionomycin and assessed IL-2 gene expression. Importantly, induction of IL-2 expression was significantly defective in FHLH4 patient T cells but could be largely restored with Ionomycin (Figure 3F). We next performed granule release assay on the in vitro cultured HD and FHLH4 patient CD8 T cells in response to anti-CD3 mediated stimulation and observed a significant defect in the FHLH4 patient CD8 T cell degranulation (Figure 3G, 3H). To determine whether the reduced granule release results from a direct defect in vesicle fusion or SOCE, we stimulated cells with Ionomycin along with (PMA). Remarkably, CD8 T cell degranulation was fully restored in FHLH4 T cells suggesting that defective SOCE primarily causes cytolytic defects in FHLH4 T cells. Of note, the patient exhibited all symptoms of FHLH4 disease and we observed strong suppression of SOCE (Figure 1I), IL-2 gene expression and secretion (Figure 3D-F). However, in vitro culture of FHLH4 patient CD8 T cells with IL-2 has been previously shown to partially or fully restore cytolytic defects by other groups [19, 20]. Therefore, considering previous reports, a relatively lower (50%) defect in CD8 T cells degranulation was not surprising [19, 20]. Collectively, these data suggest that defects in T cell degranulation as well as interleukin-2 gene expression originate from suppressed SOCE and significantly contribute to the immune dysregulation of FHLH4 patients.

STX11 regulates NFAT activation, IL-2 gene expression and degranulation in FHLH4 patient T cells.

(A-C) Estimation of nuclear translocation of NFAT. (A) Western blot showing nuclear translocation of NFAT in Jurkat T cells treated with scr or STX11 shRNA and stimulated with PMA+TG. N=3. (B) Representative confocal images of control and STX11 depleted Jurkat T cells stimulated with anti-CD3, immunolabelled with anti-NFAT1 antibody and counter-stained with DAPI. (N=3) (C) Box and whisker plot showing quantification of nuclear NFAT from cells populating 10 randomly chosen fields per group in (B). (D) Quantification of IL-2 mRNA in anti-CD3 stimulated Jurkat T cells treated with scr or STX11 shRNA using qPCR. The bars show mean ± SE of relative IL-2 mRNA. N=3. (E) Quantification of IL-2 EIA on supernatants of scr (black) or STX11 (red) shRNA treated Jurkat T cells stimulated with anti-CD3. (F) qPCR analysis of IL-2 mRNA expression in various stimulated HD (black) and FHLH4 T cells (red). (G-H) Granule release assay performed on HD (black) and FHLH4 CD8 T (red) cells (G) and its quantification (H).

Unlike most Q-SNAREs, STX11 lacks a transmembrane domain but harbors a stretch of cysteines proposed to be essential for its membrane localization close to the C-terminus (Supplementary Figure 4A) [21]. The cellular localization of STX11 is not established because, unlike primary T cells, native STX11 is undetectable in Jurkat and HEK293 cell lines using available commercial antibodies. Tagging STX11 on either N- or C-terminus results in partial degradation (Supplementary Figure 4B) and complete mis-localization of the protein (Supplementary Figure 4C). Therefore, we inserted an HA tag inside the N-terminal unstructured region of human STX11. The internal HA tagged STX11 as well as un-tagged STX11 localized in the plasma membrane of most HEK 293 cells (Figure 4A) and co-localized with Orai1-YFP in cells co-expressing Orai1 (Figure 4B). To test whether STX11 and Orai1 formed a complex, we co-expressed STX11 with either Flag- or Myc-tagged Orai1 in HEK293, prepared whole cell lysates (WCLs) and subjected them to co-immunoprecipitation (co-IP) followed by western blot using either anti-Flag, anti-Myc or anti-STX11 antibodies. Both Orai1 and STX11 could co-IP each other (Figure 4C-D). To further assess whether STX11 directly bound Orai1, we expressed and purified from E. coli, MBP-tagged N- and C-terminal cytosolic tails of Orai1 (Figure 4E). In vitro pull-down assay performed by incubating MBP-tagged Orai1 cytoplasmic tails with full length soluble His-tagged STX11 showed that the C-terminus of Orai1 directly bound STX11 (Figure 4F). SNAREs typically utilize their SNARE domain for associating with other SNAREs [22]. To identify the domain of STX11 involved in the regulation of SOCE, we expressed His-tagged Habc and SNARE domains of STX11 in E. coli, purified and assessed their binding to the MBP-tagged Orai1 tails using pull-down assay. We found that the Habc domain of STX11 bound to the C-terminus of Orai1 (Figure 4G). The SNARE domain of STX11 did not show any binding to Orai1 and only a faint binding to SNAP23 (Supplementary Figure 4D). We next used AlphaFold3 (AF3) to obtain deeper insights into the domain interactions of STX11 and Orai1. Of all the possible combinations, the complex of STX11 Habc and Orai1 C-terminus resulted in a significantly high prediction score (Supplementary Table 1). We, therefore, executed AF3 predictions with different initial seeds to generate multiple models of STX11 Habc with Orai1 C-terminus. We calculated the contact frequency of interface residues in different models and found that multiple residues showed frequency greater than 0.8 (Supplementary Figure 5A). For further analysis, we considered the predicted model with the best score (Supplementary Table 2). Next, we performed all-atom MD simulation in aqueous environment (Movies 1-3) to assess the conformation, interaction stability of the STX11 Habc and Orai1 C-terminus complex, which was stable throughout the simulation time suggested by RMSD (Supplementary Figure 5B), and binding free energy (ΔG) values (Supplementary Figure 6A) (Figure 4H). The individual trajectories were concatenated and clustered to obtain a centroid structure for this complex. There were 46 clusters, and the largest cluster, with 52 members, showed an elaborate protein-protein interface where STX11 Habc was found to be oriented in an anti-parallel orientation to the Orai1 C-terminus (Figure 4I).

STX11 directly binds resting Orai1 in plasma membrane.

Intracellular localization of STX11 and its association with Orai1. (A) Representative confocal images of HEK293 cells transfected with either HA-tagged STX11 or untagged STX11 and stained using anti-HA antibody or anti-STX11 antibody respectively. N=3 (B) Localization of HA-tagged STX11 with respect to Orai1-YFP in the basal as well as nuclear plane of HEK293 cells. N=3 (C-D) Co-IP to assess STX11 binding to Orai1. Whole cell lysates of resting and store-depleted HEK293 cells co-expressing either Flag-Orai1 and STX11 (C) or Orai1-Myc and STX11 (D) were subjected to IP and Western blot using anti-Myc, anti-Flag or anti-STX11 antibodies, as indicated. (N=3) (E) Schematic showing key domains of STX11 and Orai1 used for in vitro pull-down assays. (F) Pull-down assay showing in vitro binding of His-tagged STX11 to Orai1. (Top) Ponceau S staining showing the input of MBP alone or MBP-tagged Orai1 fragments. (Bottom) Western blot using anti-STX11 antibody. Input:1/5th of the protein. N=4. (G) Pull-down assay showing in vitro binding of His-tagged Habc domain of STX11 to Orai1 fragments. (Top) Ponceau S staining showing the input of MBP alone or MBP-tagged Orai1 fragments. Input:1/5th of the protein. (Bottom) Western blot using anti-His antibody. N=3. (H) Binding free energy distribution of STX11-Habc and Orai1 C-terminus interactions. (I) Sphere and cartoon representation of the structure of STX11 Habc (green) and Orai1 C-terminus (cyan) complex after MD simulation. The N-termini are highlighted in blue, and C-termini in red. (J) Representative confocal images showing localization of STX11 with respect to Orai1:Stim1 puncta in store-depleted cells. HEK293 cells expressing Orai1-YFP, CFP-Stim1 were transfected with HA-tagged STX11, store-depleted and stained using anti-HA antibody. (N=3) (K) Quantification of the localization of STX11 with respect to Stim1:Orai1 puncta in store-depleted cells.

Because STX11 binds the C-terminus of Orai1 and the region largely overlaps with the Stim1 binding domain, we next sought to determine whether STX11 clusters with Orai:Stim puncta. We over-expressed HA-tagged STX11 in a Orai1-YFP and CFP-Stim1 expressing stable HEK293 cell line, store-depleted with TG, and stained using anti-HA antibody. We did not observe co-clustering of STX11 in Orai:Stim puncta even 8 minutes post store-depletion (Figure 4J-K). Accordingly, STX11 did not co-IP with Stim1 in store-depleted cells (Supplementary Figure 7A). These data suggest that STX11 comes off Orai1 C-terminus prior to Stim1 dependent clustering of Orai1. Since co-IP studies suggested sustained interaction between STX11 with Orai1 (Figure 4B-C), we quantified the fraction of Orai1 that resides inside Stim:Orai puncta w.r.t total Orai1 in PM and found that post TG addition, nearly 60% of Orai1 remains outside of Stim1 puncta even in unmanipulated cells (Supplementary Figure 7B). We used AF3 to generate the complex of Stim-Orai activating region (SOAR) domain of Stim1 and Orai1 (Supplementary Figure 8A). Given that SOAR is a dimer under resting state, we used two copies each of SOAR and Orai1 [23] [24]. The prediction scores for their complex are shown in Supplementary Table 3. The best scoring model among multiple runs of AF3 was considered for MD simulations on the complex of Orai1 C-terminus (256-301) and SOAR. The ΔG values for the complex are shown in (Supplementary Figure 8B-C). Three independent MD simulations were run for the complex (Movies 4-6) which showed an initial conformational change which stabilized thereafter (Supplementary Figure 9A). All three simulation trajectories were concatenated and clustered to obtain a centroid structure for this complex (Supplementary Figure 9B). There were 43 clusters, and the largest cluster had 77 structures. Collectively, these data suggest that STX11 directly binds Orai1 with an affinity that is physiological but relatively poorer when compared to the binding of SOAR dimers to Orai1. Therefore, presence of high density of SOAR dimers in ER-PM junctions of store-depleted cells could potentially outcompete STX11:Orai1 interactions by progressively allowing formation of SOAR:Orai1 oligomers. It is also possible that another, yet to be characterized, molecular step and/or protein actively catalyzes the removal of STX11 from Orai1 prior to its entrapment in ER-PM puncta. Future studies will be able to systematically assess these possibilities.

The major interactions predicted between STX11 and Orai1 from all-atom MD simulations are highlighted in Supplementary Figure 6B-C. To assess the significance of these interactions, we performed site-directed mutagenesis of two key residues of STX11 predicted to be involved in binding to Orai1 (Figure 5A). Ectopic expression of untagged N147A_E150A STX11 showed correct localization (Figure 5A) as well as stable expression (Figure 5B) but failed to rescue SOCE in STX11 depleted Jurkat T cells (Figure 5C-D). We next expressed His-tagged wildtype and N147A_E150A mutant STX11 in E. coli., purified (Figure 5E) and assessed their ability to bind MBP-tagged Orai1 C-terminal tail in an in vitro pull-down assay (Figure 5F). Compared to wildtype STX11, N147A_E150A mutant STX11 showed reduced binding to Orai1 C-terminus. Further, we introduced four mutations, R289A_E272A_E275A_E278A, in the Orai1 C-terminus (Figure 5G), tagged with YFP and expressed in HEK293 cells. The Orai1 mutant localized to PM (Figure 5G) but failed to rescue SOCE in Orai1 deficient cells (Figure 5H-I). We expressed and purified MBP-tagged R289A_E272A_E275A_E278A mutant Orai1 C-terminus from E. coli and assessed its ability to bind purified wildtype as well as N147A_E150A mutant STX11 in a pull-down assay (Figure 5J). Binding of Orai1 mutant to both was significantly compromised. Because Stim1 binds an overlapping region of the Orai1 C-terminus, we next tested whether mutation of Orai1 residues affected Stim1 binding. We co-expressed YFP-tagged soluble CRAC activation domain (CAD) of Stim1 (YFP-CAD) (Figure 5K) with either CFP-tagged wildtype or mutant Orai1 in HEK293 cell line. YFP-CAD localizes to the cytosol in HEK293 cells but in cells overexpressing Orai1, a significant fraction of it localizes to plasma membrane due to its association with Orai1 [10]. We quantified relative YFP-CAD levels in the plasma membrane of both groups (Figure 5K-L) but found no difference in the PM recruitment of CAD suggesting that the inability of R289A_E272A_E275A_E278A mutant Orai1 to support SOCE does not result from defective binding to CAD/SOAR. Collectively, these data demonstrate a direct role of STX11 in the binding and regulation of Orai1.

Residues involved in Orai1:STX11 complex formation are crucial for SOCE.

Analysis of STX11 and Orai1 mutants. (A) Schematic showing STX11 mutations (Top). Expression of wildtype STX11 or STX11 N147A_E150A mutant in HEK293 followed by immunolabeling (A, bottom) and Western blot (B) using anti-STX11 antibody. (C-D) Representative Fura-2 calcium imaging assay (C) and quantification of repeats (D) measuring rescue of SOCE in Jurkat cells treated either with scr shRNA (black) or STX11 shRNA expressing empty vector (red), wildtype STX11 (green) or N147A_E150A mutant STX11 (blue). (E) Coomassie blue stained SDS-PAGE showing expression and purification of His-tagged wildtype or N147A_E150A mutant STX11 from E. coli. (F) Pull-down assay measuring their binding to MBP-tagged Orai1 C-terminus. (Top) Ponceau S staining showing the input of MBP-tagged fragments. (Bottom) Western blot using anti-STX11 antibody. N=4 (G) Schematic showing Orai1 mutations (Top). Confocal images of HEK293 cells expressing YFP-tagged wildtype or R289A_E272A_E275A_E278A mutant Orai1 (Bottom). (H-I) Representative Fura-2 calcium imaging assay (H) and quantification of repeats (I) measuring rescue of SOCE in HEK293 cells treated either with scr (black) or Orai1 shRNA expressing empty vector (red), wildtype Orai1 (green) or R289A_E272A_E275A_E278A mutant Orai1 (blue). (J) Pull-down assay measuring the binding of MBP-tagged wildtype or R289A_E272A_E275A_E278A mutant Orai1 C-terminus to wildtype or N147A_E150A mutant STX11. (Top) Ponceau S staining showing the input of MBP-tagged Orai1 fragments. (Bottom) Western blot using anti-STX11 antibody. N=4. (K-L) Representative confocal images (K) and quantification (L) of PM localized YFP-CAD in CFP-tagged wildtype or R289A_E272A_E275A_E278A mutant Orai1 expressing HEK293 cells.

Upon store depletion, Stim1 localizes to ER-PM junctional regions where it traps and co-clusters with Orai1. To further understand the mechanism, we first imaged HEK293 cells expressing CFP-Orai1, Stim1-YFP under resting conditions and then quantified the levels of Orai1 and Stim1 in ER-PM junctions (puncta) of store-depleted control and STX11 depleted cells. The distribution of Orai1 and Stim1 was normal in resting cells (Figure 6A). However, in store-depleted cells, clustering of CFP-Orai1 was visibly defective in STX11 RNAi treated cells (Figure 6B), even though the intensities (Figure 6C) and area of Stim1-YFP puncta were normal (Figure 6D). Quantification of Orai1 intensities confirmed a defect in entrapment (Figure 6E) with relatively higher fraction of Orai1 present outside Stim1 puncta in STX11 depleted cells (Figure 6F). To determine whether reduced clustering of Orai1 resulted from structural defects within ER-PM junctions, we reversed the direction of fluorescent tags on Orai1 as well as Stim1 and repeated the experiment in Figure 6A-F. Interestingly, the defect in Stim1 mediated entrapment of Orai1 could be rescued (Supplementary Figure 10A-B) (Figure 6G-J) but SOCE remained significantly inhibited even when Orai1 and Stim1 were co-overexpressed (Figure 6K-L). Taken together, these data show that STX11 binding to resting Orai1 has implications beyond mere co-entrapment of Orai1 with Stim1. STX11 depletion did not alter the structure or proximity of junctional ER to the PM and yet reduced SOCE even in cells co-overexpressing Orai and Stim that have been proposed to be sufficient for amplifying CRAC currents in earlier studies [25] [26] [27].

Orai1 clustering is dysfunctional in STX11 depleted cells.

Quantification of Orai1 and Stim1 intensities in the ER-PM puncta of store-depleted cells. (A-B) Representative confocal images of resting (A) and store-depleted (B) shRNA treated HEK293 cells expressing N-terminal CFP tagged Orai1 (CFP-Orai1) and C-terminal YFP tagged Stim1 (Stim1-YFP). (C-F) Mean Stim1-YFP intensity (C) and area (D) of Stim1 clusters. Mean CFP-Orai1 intensities inside (E) and outside (F) Stim1-YFP clusters of shRNA treated cells, post store-depletion. (G-J) Quantification of Orai1 and Stim1 intensities in the ER-PM puncta of C-terminal YFP-tagged Orai1 (Orai1-YFP) and N-terminal CFP-tagged Stim1 (CFP-Stim1). (G-H) Mean CFP-Stim1 intensity (G) and area (H) of Stim1 clusters. Mean Orai1-YFP intensity (I) and area (J) of Orai1-YFP clusters in control and STX11 depleted cells. (K-L) Representative Fura-2 calcium imaging assay (K) and quantification of repeats (L) measuring TG induced SOCE in Orai1-YFP and CFP-Stim1 expressing HEK293 cells treated with scr (black) or STX11 (red) shRNA.

Defects in the functional entrapment of Orai1 suggest the absence of a necessary molecular transition within Orai1 which precedes its interaction and gating by Stim1. To test whether STX11 induced structural changes within CRAC channel pore subunit, Orai1, we co-expressed Orai1-CFP along with Orai1-YFP in scramble control or STX11 depleted HEK293 cells and measured Orai:Orai FRET. We found significantly higher Orai1:Orai1 FRET in STX11 depleted resting cells even in the absence of ectopically expressed Stim1. (Figure 7A). We have previously shown that store-depletion induces on-site assembly of Orai1 dimers into oligomers [14]. Notably, store-depletion induced a similar level of increase in Orai1:Orai1 FRET in scramble control cells. To confirm these findings, we performed acceptor photobleaching analysis on Orai1-CFP (donor), Orai1-YFP (acceptor) co-expressing HEK293 resting cells where we had stably knocked down the expression of endogenous Orai1 and Stim1. Although the trends were largely similar, the differences in Orai1:Orai1 FRET were much more significant in these cells (Figure 7B-C). These data demonstrate a previously unsuspected, novel, molecular transition in Orai1 necessary for CRAC channel activation, which we define as ‘priming’. Further, these data show that priming precedes the interaction of Orai1 with Stim1 in ER-PM junctions. To determine whether ‘priming’ simply induced a structural change or facilitated correct assembly of Orai1 multimers, we expressed Flag-Orai1 in control or STX11 depleted resting cells, mechanically lysed to isolate membranes, crosslinked Orai1 oligomers using BS3 and resolved using SDS-PAGE (Figure 7D) (Supplementary Figure 11A). We found a greater percentage of higher order oligomers in STX11 depleted cells suggesting stoichiometric changes (Figure 7E) (Supplementary Figure 11A).

STX11 primes Orai1 for optimal on-site assembly.

Oligomerization of Orai1. (A) N-FRET analysis of scramble control and STX11 shRNA treated HEK293 cells co-expressing Orai1-CFP and Orai1-YFP before and after store-depletion. (B&C) Representative images (B) and quantification (C) of FRET analysis using acceptor photobleaching of control and STX11 shRNA treated Orai1 Stim1 double knockdown resting HEK293A cells co-expressing Orai1-CFP (donor) and Orai1-YFP (acceptor). (D-E) Representative western blot (D) and quantification of band intensities from repeats (E) showing distribution of BS3 crosslinked Flag-Orai1 oligomeric bands in control and STX11 depleted HEK293 cells. N=3. (F-H) Fura-2 calcium imaging assay (F) and quantification of repeats (G-H) to measure constitutive calcium influx in Orai1-CFP and YFP-CAD expressing, control (black) or STX11 depleted (red) HEK293 cells at 0 mM (G) and 2 mM (H) extracellular Ca2+. (I-J) Fura-2 calcium imaging assay (I) and quantification of repeats (J) to measure constitutive calcium influx in Orai1-H134S mutant expressing control (black) and STX11 depleted (red) HEK293 cells. (K-L) Fura-2 calcium imaging assay (K) and quantification of repeats (L) to measure constitutive calcium influx in Orai1-ANSGA mutant expressing control (black) and STX11 depleted (red) HEK293 cells.

Increasing the density of Stim1 has been earlier proposed to bypass the requirements for accessory factors in SOCE. To determine whether increasing the levels of Stim1 could compensate for STX11 depletion, we expressed Orai1 tethered to two Stim-Orai activating region (SOAR) domains of Stim1 and GFP (Orai1-S-S-GFP) (Supplementary Figure 11B-C) in HEK293 cells. Orai1-S-S-GFP has been previously shown to constitutively activate Orai1 [23]. In STX11-depleted Orai1-S-S-GFP expressing cells, the magnitude of calcium entry was significantly smaller (Supplementary Figure 11D-E). To determine whether reduced calcium entry resulted from a defect in the ability of SOAR to bind Orai1, we expressed YFP-CAD, in control or STX11 depleted Orai1-CFP expressing HEK293 cell line (Figure 7F). Quantification of PM localized YFP-CAD as a fraction of total CAD showed no defect in STX11 depleted cells (Figure 7F). Yet, as seen with Orai1-S-S-GFP, constitutive calcium entry was significantly reduced (Figure 7G-H). We next expressed a constitutively active H134S mutant of Orai1 (Supplementary Figure 11F) as it harbors an open pore and the C-terminal cytosolic tails of Orai1 are unlatched, pointing towards the cytosol [28]. Remarkably, expression of H134S Orai1 completely restored constitutive calcium influx in STX11 depleted cells (Figure 7I-J). To further establish whether binding with STX11 induces straightening of Orai1 tails or molecular transitions within transmembrane helices, we expressed in STX11 depleted cells the constitutively active ANSGA (261-265) mutant of Orai1 (Supplementary Figure 11G), which harbors 4 consecutive mutations in the Orai1 C-terminus resulting in the straightening of the C-terminal tail along with constitutive activation of CRAC currents [29]. The constitutive calcium entry in STX11 deficient Orai1 ANSGA mutant expressing cells was significantly suppressed (Figure 7K-L). Taken together, these data show that STX11 induces a molecular switch in the pore-forming subunit, Orai1, which precedes and primes it for optimal on-site assembly into multimers and subsequent gating by Stim1. Such SNARE dependent priming of Orai1 is crucial and cannot be compensated by merely increasing the density of Stim proteins bound to Orai or by introducing mutations that alter the orientation of Orai1 tails in vivo. The proposed models depicting unprimed and primed Orai1 (Figure 8A) and its role in T cell function (Figure 8B) are shown.

Models depicting STX11 mediated priming of Orai1 and its role in T cell signaling and function.

(A) STX11 unbound (unprimed) and bound (primed) Orai1 in resting and store-depleted cells. For simplicity, STX11 interaction with only two subunits of Orai1 has been shown. (B) Regulation of T cell SOCE, NFAT activation and effector function by STX11.

Q-SNAREs are often found in complex with other SNAREs such as SNAP23/25/29 in target membranes. Some previous studies have implicated the complex of SNAP23/25 and STX1A in the regulation of specific ion channels [30, 31]. The three SNAPs, SNAP23/SNAP25/SNAP29 are each capable of contributing two SNARE domains to the trans- and cis-SNARE complex. RNAi mediated depletion of SNAP23/SNAP25/SNAP29 did not show a significant reduction in SOCE of HEK 293 cells (Supplementary Figure 12A-C). To determine whether SNAP23/SNAP25/SNAP29 might still form a part of the STX11:Orai complex to regulate SOCE, we co-expressed myc-tagged Orai1 with untagged SNAPs and performed respective co-IPs. We did not find any interaction between either of the three SNAPs and Orai1 (Supplementary Figure 12D). Similar studies were performed by co-expressing YFP-Stim1 and the three myc-tagged SNAPs. Again, no interaction was found between any of the SNAPs and Stim1 (Supplementary Figure 12E). These data show that, unlike in the case of vesicle fusion, STX11 does not collaborate with SNAP23/SNAP25/SNAP29 in the regulation of SOCE.

FHLH is a heterogeneous immune disorder where diverse genes such as Prf1, STX11, UNC13D, Munc18-2, MAGT1, XIAP, LYST, AP3B1, RAB27A, ITK, amongst several others, are broadly grouped together based on disease symptoms, although the age of onset and severity varies across the spectrum [32, 33]. Congenital immunodeficiency involving defects in the CTL and NK cell cytotoxicity appear to be the common factors underlying the development of FHLH, although the individual pathways involved are highly diverse. Using shRNA, we screened 6 additional genes, Perforin1, RAB27A, UNC13D and Munc18-2, Synaptotagmin11 and 13, that have been either associated with vesicle fusion or routinely grouped together with STX11 in diagnosing FHLH (Supplementary Figure 13). SOCE was not affected by depletion of most genes. Future studies will evaluate the role of others.

Discussion

CRAC channels conduct a small, highly specific calcium current in response to the depletion of intracellular calcium stores. According to the prevailing view, Orai multimers reside in the PM and depend on the ER resident store sensor, Stim, for all structural transitions leading up to their activation [10]. Unlike other ion channels, the process of activation of CRAC currents is exceptionally slow. For instance, following break-in or store-depletion, on average, it takes ∼5 minutes for Stims to cluster in the ER-PM junctions and for measurable CRAC currents to flow from Orai1. During this time, Stims undergo intramolecular transitions and slowly trap and gate freely diffusing Orai. It is, therefore, believed that no molecular transitions take place within the pore forming subunit, Orai, until its entrapment by Stim in the ER-PM junctions.

We have shown that in STX11 deficient cells, the basal Orai:Orai FRET is significantly higher even in the absence of Stim1 or store-depletion. These data establish that Orai1 undergoes a, hitherto unrecognized, SNARE dependent molecular transition which precedes trapping and gating by Stim1. We have earlier shown that store-depletion induces dynamic on-site assembly of Orai1 dimers into oligomers [14]. In STX11 depleted resting cells, the ∼25-50% increase over basal Orai:Orai FRET largely results from a change in the oligomerization of Orai1 because the proportion of BS3 crosslinked Orai1 oligomers was also found to be shifted towards higher molecular weight. It is likely, though, that a change in the overall conformation of Orai1 protomers facilitated the change in store-dependent Orai assembly and contributed to the overall change in Orai:Orai FRET.

The change in Orai:Orai FRET induced by STX11 depletion not only persisted but increased in Orai1, Stim1 double deficient cells. Furthermore, STX11 bound a region of Orai1 C-terminus that has been previously shown to bind Stim1. Therefore, our data suggest that STX11 induced molecular transitions in Orai are independent of and precede its association and clustering by Stims. Because tagging of STX11 with fluorescent tags on either end results in mis localization and degradation, our study could not demonstrate the real-time removal of STX11 from the Orai1- C-terminus to allow for Stim1 binding in live cells. It is also unclear whether binding or removal of STX11 induced the proposed molecular transition defined as ‘priming’ of Orai1. Further, it remains to be determined whether STX11 removal is actively induced by an additional chaperone or passive displacement from the formation of relatively higher avidity oligomeric associations between Orai1:Stim1 in ER-PM junctions. Future studies will be able to systematically evaluate each of these possibilities.

We have shown that the C-terminal cytosolic tail of Orai1 directly couples with STX11 in resting cells. Although binding appeared reasonably strong in co-IPs, STX11:Orai1 interaction appeared somewhat weak in pull-downs. This is likely due, in part, to the instability of purified His-STX11 in solution which necessitated shorter co-incubation times during pull-down assays. Instability could arise from the absence of a stabilizing co-chaperone, such as one of the known binding partners of SNAREs [34] [35] or membranes to which STX11 attaches in lives cells via its C-terminal cysteines [21]. Together, these factors appear crucial for the stabilization of purified STX11 in vitro.

Stim1 binds the N- as well as the C-terminus of Orai1 but their individual relevance in the trapping versus gating of Orai1 remains unclear [36]. Though crucial for trapping and gating, Orai1 tails were missing from early structures of Drosophila Orai [28]. A previous NMR structure suggested that C-terminal tails of two adjacent Orai1 subunits, bend, pair with each other in an antiparallel fashion and sit closely apposed to PM [37]. However, in recent structures of constitutively active H134 mutant Orai, the C-terminal tails were found to orient away from the membrane [28]. In STX11 depleted cells, switching the CFP-tag from the Orai1 N- to the C-terminus could rescue its clustering but not gating by Stim1. Furthermore, STX11 depletion inhibited the constitutively active ANSGA mutant of Orai1 [29], where the tails of Orai1 are proposed to be constitutively unlatched. These data essentially reinforce our conclusions that STX11 induced molecular shifts encompass Orai1 transmembranes.

Using Fura-2 imaging assays in STX11 depleted cells, we have conclusively established that STX11 is necessary as well as limiting not only in live and unperturbed HEK, U2OS, Jurkat and primary T cells but also HEK293 cells co-overexpressing Orai1 and Stim1. Previous whole cell patch clamp analysis of HEK293 cells co-overexpressing Orai1 and Stim1 showed up to 50-100-fold amplification of CRAC currents whereas Fura-2 based imaging of live cells routinely shows only 2-fold amplification [25] [26] [27]. The patch clamp studies performed on Stim, Orai over-expressing cells are often extrapolated to conclude that all other accessory factors that have since been identified are unnecessary for native CRAC currents. A major problem with this interpretation is that even though amplification of CRAC currents was shown, none of the previous patch clamp studies established whether the higher currents resulted from a greater number of active channels or unchecked conductance per channel by performing single channel recordings. The fact that recording of ICRAC in the whole cell configuration involves use of 20mM extracellular calcium and dialysis of cells with high concentrations of BAPTA, a fast calcium chelator, the calcium dependent inactivation of CRAC channels, a key mechanism of regulation of ICRAC magnitude in live cells is disrupted. Furthermore, hyperpolarizing potentials and voltage clamping used to record CRAC currents prevents the progressive depolarization of cells due to influx of cations which would otherwise dampen the driving force for CRAC currents in intact cells. Given these limitations, Fura-2 assays offer a more realistic and physiological reflection of SOCE amplitudes in live cells as they allow for these regulatory mechanisms, which are necessary for the regulation of native currents to be still operational. While patch clamp remains a direct method of verification of CRAC currents, exercising logic in extrapolation of amplitudes of ICRAC using patch clamp is imperative for the field to move forward.

Most human mutations in the STX11 coding region result in the complete loss or severe depletion of STX11 protein levels [21]. The FHLH4 disease, therefore, results from reduced STX11 expression in most reported patients. Targeted mutagenesis of the STX11:Orai1 interface lining residues based on the AlphaFold 3 model, however, didn’t destabilize or mis-localize either N147A_E150A STX11 or R289A_E272A_E275A_E278A Orai1 and yet both mutants showed reduced binding to their respective wildtype counterparts. Importantly, the R289A_E272A_E275A_E278A mutant of Orai1 retained the ability to bind Stim1.

In addition to the established cytolytic defects in FHLH4 T cells, for the first time, we have shown that STX11 deficient, FHLH4 T cells harbor a severe defect in store-operated calcium signaling and IL-2 gene expression. Our data strongly suggest that SOCE defects significantly contribute to FHLH4 disease progression via, as well as independent of, exocytotic/ degranulation defects. We propose that design and early administration of CRAC channel agonists could potentially provide an alternative to immunosuppressants and bone marrow transplants for FHLH4 patients.

In summary, we have shown that binding to a Q-SNARE, STX11, provides an on-site molecular switch in the CRAC channel pore, Orai1, necessary for its optimal assembly and function. We suggest that ion channel priming is a novel, direct and primary role of specific SNAREs. In accordance with this hypothesis, early SNARE-like proteins have been found in bacteria, which lack an endomembrane system but express a variety of ion channels [38–40]. SNAREs are now thought to have evolved from a common archaeal precursor found in the genomes of Asgard and members of Legionella [41].

Materials and Methods

Plasmid constructs and transfection

Stim1 CAD was sub-cloned in eYFPC1 vector after amplification from YFP-Stim1 plasmid to generate YFP-CAD. pLKO.1, psPAX2 and pMD2.G were purchased from Addgene. pLKO.1 cloned shRNA sequences targeting the genes of interest, were either purchased from Horizon Discovery, UK or designed and cloned in-house using the Broad Institute portal (https://portals.broadinstitute.org/gpp/public/gene/search). Full-length human STX11 was cloned from human cDNA prepared from HEK293 cells and subcloned into pcDNA3.1(+), pcDNA/4TO/Myc-HisA (Invitrogen, Grand Island, NY), pMSCV-IRES-mcherry (Addgene), pEF1alpha-IRES (Clontech) and pET28b vector [16] with an N-terminal 6XHis-tag. The fragments of STX11 were cloned in pET28b vector with an N-terminal 6xHis tag for expression in E. coli. HA tag was inserted between H30 and G31 in STX11 cloned in pMSCV-IRES-mCherry construct using PCR. H134S and ANSGA mutants of Orai1 were generated using site directed mutagenesis and PCR, respectively. The fragments of Orai1 (1-87, 1-47, 48-103, 256-301 and 272-292) were amplified from full length constructs and cloned into pMAL-c5X vector (New England Biolabs), in-frame with MBP protein coding sequence, as described previously [16]. SNAP23, SNAP25 and SNAP29 cDNA cloned in pCMV-Sport6.1 vectors were purchased from Dharmacon and subcloned into pcDNA4/TO/Myc-HisA. SNAP23 was subcloned in pMAL-c5X vector in-frame with MBP. All plasmid DNA transfections in human cell lines and primary cells were done using Lipofectamine 2000 (Invitrogen)/ Lipofectamine 3000 (Invitrogen) or Amaxa nucleofection kit (Lonza, Basel, Switzerland) respectively, as per manufacturer’s protocol.

Cell Lines

Lentiviral shRNA transduction experiments were performed in HEK293, U2OS or Jurkat (ATCC, Manassas, VA) cell lines cultured in low glucose DMEM (Hyclone, Logan, UT) or RPMI (Hyclone) respectively, with 5-10% fetal bovine serum (Hyclone), 1X Penicillin Streptomycin, and GlutaMax (Gibco, Grand Island, NY). Stable cell lines generated using HEK293 and U2OS parent lines have been described previously [9]. HEK293-FT cell line was cultured in high glucose DMEM with 10% FBS and 10mM HEPES, 1X Penicillin Streptomycin, 1X GlutaMax and 1X non-essential amino acid and transfected with the appropriate plasmids to generate viral supernatants. All cell lines were tested for mycoplasma contamination twice every year and found to be negative.

Antibodies and Reagents

Lentiviral transductions

For the generation of lentiviral supernatants, shRNAs cloned in pLKO.1 were co-transfected with psPAX2 packaging and pMD2.G envelope plasmids into HEK293-FT cells using the calcium phosphate method of transfection. 48- and 72-hours post-transfection, viral supernatants were collected, pooled and stored at −80°C till further use. For transduction of HEK293 and Jurkat cells, 0.1 million cells were plated in 6- or 24-well plates either the day before (HEK293) or the same day (Jurkat). Viral supernatants were added to the cells along with 8ug/ml polybrene and cells were spun at 2500 RPM, 30°C for 90 mins. 24 hours or 48 hours post-spinfection, Puromycin was added at a final concentration of 1ug/ml to HEK293 and Jurkat cells, respectively. Cells were analyzed 3-5 days post transduction.

Immunocytochemistry and confocal imaging

For STX11 localization, resting or store-depleted (1μM Thapsigargin) WT HEK293 or HEK293 cells stably expressing Orai1-YFP were either plated in 6 well plates and spinfected with viral supernatants generated from pMSCV-STX11(HA)-IRES-mCherry, pMSCV-IRES-mCherry (empty vector control) or plated in carbon coated, glow discharged 35mm glass bottom dishes (IBDI) and transfected with pEF1alpha-STX11-IRES-mCherry. 24 hr post-spinfection, cells were plated in 35mm glass bottom dishes. For immunolabelling, cells were washed with Ringer’s buffer, fixed with 4% PFA and blocked with 3% BSA containing 0.1% NP40 for 1.5 hrs. Post-fixation and permeabilization, cells were incubated either with anti-HA (CST) or rabbit anti-Stx11 (SySy/ Proteintech) primary antibodies O/N at 4°C, washed and incubated with either Alpaca VHH anti-rabbit AF488, donkey anti-rabbit A647 or goat anti-rabbit A647 secondary antibodies for 1 hour at room temperature (RT). Cells were counter stained with DAPI for 10 min at RT and imaged using Olympus FV3000 laser scanning confocal microscope. Images were acquired sequentially using the following parameters: DAPI (DM405/488 dichroic, 405nm excitation, 430-470nm emission); anti-HA Alexa488 (DM405/488, 488nm excitation, 500-590nm emission); STX11 (DM405/488/561/640, 640nm excitation, 650-750nm emission); Orai1-YFP (DM405/488/561/640, 488nm excitation, 521-591nm emission); CFP-Stim1 (DM405-445/514/594, 405nm excitation, 448-510nm emission). For STX11-N147A-E150A mutant, HEK293 cells were spinfected with viral supernatants and imaged as described above where WGA (Wheat germ agglutinin) conjugated with Alexa 488 was used to mark plasma membrane.

Quantification of Orai1 and Stim1 intensities inside/outside ER-PM puncta

HEK293 cells stably expressing Orai1-YFP with CFP-Stim1 or CFP-Orai1 with Stim1-YFP were plated in 6 well plates and transduced with scramble or STX11 shRNA. 48hr post transduction, cells were trypsinized and plated in carbon coated 35mm dishes. Images were acquired using a TIRF microscope setup described before 14. Briefly, resting cells were imaged first and positions were marked. Store-depleted images were acquired after incubation with 1 μM Thapsigargin (TG) and 10mM EGTA for ∼6 mins. Analysis was done using Otsu local thresholding in ImageJ software. Stim1 and Orai1 intensity (AU) and area µm2 values obtained from the analysis were plotted using Origin software.

Orai:Orai FRET

Scramble and STX11 shRNA transduced HEK293 cells were plated on freshly carbon coated, glow discharged 35mm glass bottom dishes (IBDI) and transfected with 0.5 μg each of Orai1-CFP and Orai1-YPF plasmid DNA using Lipofectamine 3000. 16 hours post transfection, co-transfected cells were imaged using PlanApoN 60XO/1.42 NA objective with Olympus FV3000 laser scanning microscope. Orai1-CFP, Orai1-YFP and FRET images were acquired with the following excitation laser and emission bandwidth parameters using DM405-445/514 dichroic mirror: CFP Ex-405nm Em-460-630nm; YFP Ex-514nm Em- 530-630nm; FRET Ex-405nm, Em-530-630nm. ROIs representing the plasma membrane rim were drawn around each cell, followed by local thresholding using Bernsen’s algorithm in Fiji software to generate an image mask which was applied to CFP, YFP, FRET images to extract mean intensity values. The background subtracted mean intensities of FRET (IFRET), CFP (IDonor) and YFP(Iacceptor) images were measured for each cell. Cells singly expressing Orai1-CFP or Orai1-YFP only were used as CFP donor or YFP acceptor channel bleed-through (BT) controls. The BT was calculated using the following formula: FRET and NFRET percentage was calculated by subtracting the contributions of CFP and YFP bleed through from the FRET signal using the following: FRET=IFRET−(BTDonor∗IDonor)−(BTAcceptor∗IAcceptor) NFRET%=(FRET(IAcceptor∗IDonor))∗100

Acceptor Photobleaching

Scramble and STX11 shRNA transduced, Orai1, Stim1 double knockdown HEK293 cells where were plated on Poly-D-Lysine coated glass bottom dishes (Cellvis) and transfected with 0.5 ug DNA each of Orai1-CFP and Orai1-YPF using Lipofectamine 2000. 16 hours post transfection, co-transfected cells were imaged using PlanApoN 60XO/1.42 NA objective with Olympus FV3000 laser scanning microscope. Images were acquired with the following excitation laser and emission bandwidth parameters using DM405-445/514 dichroic mirror: CFP Ex-405nm Em-460-508nm; YFP Ex-514nm Em- 530-630nm, with 4% offset. The pre bleach images of Orai1-CFP (the donor) and Orai1-YFP (the acceptor) were captured, sequentially. This was followed by bleaching of the acceptor, Orai1-YFP, using 100% 514 laser power for 2 mins and then acquiring the donor and acceptor images post bleach. ROIs were drawn around the cell periphery using the Orai1-CFP image. Local thresholding was performed using Bernsen’s algorithm to generate binary masks, which were then applied to the CFP images to measure mean intensities. The background subtracted mean intensities of CFP images were measured for each cell and the FRET efficiencies were calculated as follows: E (%) = (1 -Ibefore/Iafter) *100. Ibefore and Iafter denotes the average fluorescence intensity of the donor in the plasma membrane before and after acceptor photobleaching.

Alpha-bungarotoxin binding assay

U2OS cell line stably expressing Orai1-BBS-YFP construct has been described previously14. Briefly, Orai1-BBS-YFP expressing U2OS cells were either transduced with scramble shRNA/ STX11 shRNA/ pMSCV-IRES-mcherry (EV) or pMSCV-STX11-IRES-mcherry. Transduced cells were trypsinized either 3-4 days post-transduction or 48 hour post-transduction (in the case of pMSCV), centrifuged and labelled on ice with AF647 conjugated alpha-bungarotoxin (1μg/ml) containing 2% FBS in 1X HBSS for 30 mins. Following this, cells were washed with HBSS and fixed with 4% PFA (∼20 min at RT). Samples were run using Cytoflex FACS analyzer (Beckman Coulter) and data were analyzed using Flow Jo software.

Validation of STX11 knockdown using qPCR

Scramble and STX11 shRNA treated HEK293 cells were trypsinized, counted and equal number of cells were used for lysis. The lysates were homogenized using QIA shredder columns (Qiagen) and RNA isolation was done using RNeasy Plus Mini kit (Qiagen). Total RNA isolated from both the groups was used for cDNA synthesis using oligo dT primers and Superscript IV as per manufacturer’s guidelines. The concentration of synthesized cDNA was estimated using Qubit ssDNA assay kit and STX11 TaqMan probes were used to perform the qPCR using LightCycler 96 (Roche). Beta-actin was used as housekeeping control.

Single cell Ca2+ imaging

Cells were plated in carbon-coated glass bottom dishes (one day prior or 30 mins before the assay for HEK293 and Jurkat T cells respectively) and loaded with 1μM Fura-2 AM dye in HBSS (CaCl2 1.8mM, KCl 5.36mM, MgSO4 0.81mM, NaCl 136.89 mM, Na2HPO4 0.335mM, D-Glucose 5.55mM) for 30 mins, at 37°C in the dark. After incubation, cells were washed and incubated in Ringer’s buffer (135mM NaCl, 5mM KCl, 1.8mM CaCl2, 1mM MgCl2, 5.6mM Glucose, 10mM HEPES) for an additional 10 mins, washed and imaged in Ringer’s buffer or Calcium-free Ringer’s buffer (135mM NaCl, 5mM KCl, 1mM MgCl2, 5.6mM Glucose, 10mM HEPES at pH7.5), as indicated. Olympus IX-71 inverted microscope equipped with a Lamda-LS illuminator (Sutter Instrument, Novato, CA), Fura-2 (340/380) filter set (Chroma, Bellows Falls, VT), a 10X 0.3NA objective lens (Olympus, UPLFLN, Japan), and a Photometrics Coolsnap HQ2 CCD camera was used to capture images at a frequency of ∼1 image pair every 2 or 4 seconds interval. Data were acquired, analyzed and plotted using MetaFluor (Molecular Devices, Sunnyvale, CA), Microsoft Excel, and Origin softwares. Approximately 30-40 cells were imaged per group in each experiment unless otherwise stated. Mean SOCE ± SE for each experiment was calculated from three independent experiments where the maximum value of the scramble shRNA group (amongst the three experiments) was set at 100% and the relative responses of other scr and STX11 shRNA groups were normalized w.r.t it.

For constitutive calcium influx assay, shRNA treated HEK293 cells were either transfected or nucleofected with Orai1-S-S-EGFP, Orai1(H134S)-YFP or Orai1(ANSGA)-YFP and ∼12-14 hour post transfection (∼4-5 hour post nucleofection), cells were loaded with Fura-2 AM, washed and imaged in Ringer’s buffer with 0 mM calcium to acquire baselines and 2 mM calcium thereafter. To measure CAD induced constitutive calcium influx, Orai1-CFP expressing stable HEK293 cells were nucleofected with YFP-CAD and analyzed ∼4 hour later as described above. To identify cells expressing Orai1 mutants and fusion proteins in the plasma membrane, cells were imaged using a 20X 0.7NA water objective lens (Olympus, UApoN340, Japan). Images were acquired at a frequency of ∼1 image pair every 10 seconds interval to avoid photobleaching and analyzed as described above.

Measurement of NFAT nuclear localization by Western blot

Jurkat T cells were transduced with scramble or STX11 shRNA to knock-down STX11 as described above. On day 4 post-transduction, cells were collected, spun down and resuspended in plain RPMI media and rested for 1hr at 37°C. Following this, cells were counted and divided into two equal groups. One group was resuspended in RPMI (unstimulated) and the other in RPMI media containing 1 μM TG + 10 ng/ml PMA (Phorbol Myristate Acetate) (stimulated) and incubated at 37°C for 30 mins. Following incubation, cells were pelleted and the nuclear and cytosolic protein fractions were separated using NE-PER kit according to the manufacturer’s guidelines and subjected to SDS-PAGE and western blot using the mouse anti-NFATc2 primary antibody (Santa Cruz) followed by Donkey anti-mouse secondary antibody.

Estimation of nuclear translocation of NFAT by immunolabeling

Control and STX11 depleted Jurkat T cells were rested in plain RPMI for ∼1 hour prior to the assay. ∼100,000 cells per group were plated on freshly carbon-coated coverslips for 40 minutes and stimulated for 1 hour at 37°C with 5 μg/ml anti-CD3 antibody, diluted in plain RPMI. Following this, cells were washed and fixed using 4% PFA diluted in 1X PBS for 20 minutes at RT, washed twice and incubated with 30 mM Glycine for 10 minutes at RT. For permeabilization and blocking, cells were incubated with 0.1% Saponin, 3% BSA diluted in 1X PBS for 1 hour at RT, washed and incubated with anti-NFAT primary antibody (anti-NFAT1, CST) at 4°C, overnight. Following primary antibody application, cells were washed with 1X PBS and incubated with anti-Rabbit AF647 secondary antibody for 1 hour, washed and stained with DAPI for 5 minutes followed by additional washes. Images were acquired in the FV3000 confocal microscope using a 100X objective lens. DAPI was used to identify the nuclear area. Nuclear versus whole cell (total) NFAT mean intensity ratio was plotted across different groups.

Whole cell lysates (WCLs), Western blot and Co-immunoprecipitation (Co-IP)

HEK293 cells transfected with the desired plasmids were lysed using buffer containing 50 mM Tris-Cl (pH 8.0), 150 mM NaCl, 1% NP-40, 1 mM PMSF, and protease inhibitor cocktail. The whole cell lysates were centrifuged at 21000 g for 15 minutes and supernatants were subjected to SDS-PAGE, and western blot. For immunoprecipitations, lysates were divided into two equal parts. To the first tube, the appropriate anti-mouse or anti-rabbit primary antibody was added and to the second, same amount of the respective IgG control antibody was added. The antibody-lysate mixtures were incubated overnight at 4°C. Following this, Protein A/G Mag Sepharose beads were added and incubated with the antibody-lysate mixtures for 4 hours, washed with the lysis buffer containing 0.1% NP-40, boiled with 1X Laemmli buffer + 120 mM DTT and subjected to SDS-PAGE and western blot analysis. Typically, 1/10th of the whole cell lysate (WCL) was loaded in the input lane of the co-IP blots. The full-length blots pertaining to each figure have been deposited in Zenodo https://doi.org/10.5281/zenodo.19045452.

E. coli expression and in-vitro binding assays

Full-length His6-tagged Stx11 and truncation mutants were cloned in pET28b, expressed in Lemo21 (DE3) E. coli cells and induced with 1 mM IPTG (Isopropyl β- d-1-thiogalactopyranoside) for 14-18 hours at 18°C. The cell pellets were lysed in buffer containing 50 mM Tris-Cl (pH 8.0), 150 mM NaCl, 10% Glycerol, 1 mM PMSF, 1% Sarkosyl, 0.1 mg/ml Lysozyme, protease inhibitor cocktail and sonicated on ice. DNase I was added after sonication and the lysates were further incubated for ∼60 minutes before centrifugation at 21000 g for 40 mins. The supernatants were subjected to SDS-PAGE to confirm expression by Coomassie staining and subsequently used for pull-down assays. MBP-tagged Orai1 constructs were expressed in Lemo21 (DE3) E. coli cells, induced with 0.3mM IPTG (Isopropyl β- d-1-thiogalactopyranoside) for 14-18 hours at 18°C. The cell pellets were lysed in buffer containing 50 mM Tris-Cl (pH 8.0), 150 mM NaCl, 5% Glycerol, 1 mM PMSF, 0.1 mg/mL Lysozyme and protease inhibitor cocktail (1:200) and sonicated on ice. DNase I was added after sonication and the lysates were further incubated for ∼60 minutes before centrifugation at 21000 g for 40 mins. The supernatants were collected and subjected to SDS-PAGE to confirm expression by Coomassie staining and used for pull-down assays. Following lysis, the His6-tagged Stx11 proteins were purified using Talon Beads and MBP-tagged Orai1 proteins using Dextrin Sepharose/Amylose resin according to the manufacturer’s guidelines. For in-vitro binding assays, lysates prepared from cells expressing MBP or MBP-tagged Orai1 fragments were incubated with 25 μL Dextrin Sepharose Beads and incubated for 2 hours at 4°C. After incubation, beads were washed thrice and ∼50-125 ng of purified His-tagged full length Stx11, His-tagged SNARE or His-tagged Habc domains were diluted in buffer containing 50mM Tris-Cl (pH8), 150 mM NaCl, 5% Glycerol and 0.1% NP-40 and added to the beads. Following 1 hour of incubation at 4°C, the beads were washed thrice, re-suspended and boiled in the binding buffer containing 1X Laemmli and 120 mM DTT. The eluate was subjected to SDS-PAGE and Western blot. MBP and MBP-tagged proteins were detected with Ponceau staining, Full length STX11 using Rabbit anti-STX11 primary antibody (Thermo) followed by Donkey anti-rabbit HRP and His-tagged STX11 fragments were detected or mouse anti-6XHis primary antibody (Invitrogen) followed by Donkey anti-mouse HRP secondary antibody.

BS3 Crosslinking of Flag-Orai Oligomers

Scramble and STX11 shRNA transduced HEK 293 cells were transfected with Flag-Orai1, harvested, washed, pelleted and subjected to one cycle of snap freeze-thaw. The pellet was resuspended in ice cold lysis buffer (1X HBSS, 1mM PMSF and protease inhibitor cocktail), sonicated for 30 secs (Branson 2510) and mechanically homogenized using dounce homogenizer. The homogenized supernatant was spun at 1500 x g for 10 mins at 4° C, to pellet the cell debris. The supernatant was ultracentrifuged at 105,000 x g for 1 hour at 4° C (Optima MAX-XP ultracentrifuge, Beckman Coulter). The membrane fraction pellet was resuspended in lysis buffer and re-homogenized using dounce homogenizer. The protein concentration was estimated using BCA assay and BS3 cross-linker was added to the resuspended membranes, at a final concentration of 1mM, and incubated for 30 mins at room temperature. BS3 was quenched using 50mM Tris-HCl pH 7.5 for 15 mins at room temperature, followed by addition of 1X Laemmli and 100 mM DTT and incubation at 37° C for 10 mins. The samples were resolved using 4-15% gradient SDS-PAGE and subjected to Western blot using anti-Flag antibody.

Quantification of Flag-Orai oligomers using Western blot

Equal size ROIs were drawn over corresponding protein bands of same size, using ImageJ. The total intensity of each band was measured by calculating the area under the intensity curve and its relative proportion was calculated by dividing the raw intensity of each band by the sum of intensities of all the bands in the respective lanes.

qPCR to estimate gene expression in Jurkat T cells

Scramble (scr) and STX11 shRNA treated Jurkat T cells were rested for ∼1 hour prior to the assay, stimulated with soluble anti-CD3 (5-10 μg/ml) for 3 hours and washed with HBSS. Total RNA was isolated using RNeasy Mini kit as per manufacturer’s instructions. cDNA was synthesized using random hexamers and Invitrogen Superscript IV kit and quantified using Qubit ssDNA assay. Taqman probes for IL-2 and beta-actin housekeeping control were used for performing the qPCR in triplicates. Data analysis was done by calculating the double delta Ct values.

EIA to measure IL-2 levels in supernatants of Jurkat T cells

Scramble (scr) and STX11 shRNA treated Jurkat T cells were stimulated with plate-coated anti-CD3 (10 μg/ml) and the culture supernatants were collected at 12 and 24 hour post-stimulation. EIA plates were coated with purified mouse anti-human IL-2 capture antibody (1 μg/ml) overnight at 4°C, washed with 0.05% PBST and blocked with 2% BSA for 2 hours. After blocking, the plates were incubated with the culture supernatants for 2 hours at RT, washed and incubated with the detection antibody (Biotin mouse anti-human IL-2, 1 μg/ml) for 2 hours. After washing, Streptavidin HRP (0.1 μg/ml) followed by TMB reagent were added. 250 mM HCl was added to stop the reaction and absorbance was measured at 450 nm.

Isolation and culture of human PBMCs

Whole blood freshly collected in Heparin or Citrate Phosphate Dextrose Adenine (CPDA) solution was subjected to density gradient centrifugation using Ficoll-Paque PLUS media. The buffy coat was separated, washed twice with HBSS and cultured with IL-2 (50 ng/ml). Cells were stimulated once per week with PHA (2 μg/ml) for 72 hours and rested in IL-2 for the remaining 72-96 hours. Unless specified, all the assays were performed following 24-48 hours of rest, post stimulation.

Isolation of genomic DNA and sequencing

Genomic DNA was extracted from PBMCs using Phenol-Chloroform-Isoamyl alcohol. STX11 genomic region flanking the mutation was PCR amplified using primers: 5’ Forward -CATGCACGACTACAACCAGGC and 3’ Reverse - GGGACAGCAGAAGCAGCAGAGGG. The resulting PCR products were separated on 2% agarose gel, excised and extracted using the Macherey-Nagel Nucleospin columns and subjected to Sanger sequencing using the 5’ Forward PCR primer to confirm the mutation.

Measurement of SOCE in human PBMCs

PBMCs in culture were washed, rested and allowed to adhere to freshly carbon-coated (IBDI) dishes for 1 hour in plain RPMI at 37°C. The cells were washed with HBSS and incubated with 1 ml of 1 μM Fura-2-AM diluted in Ringer’s buffer for 40 minutes, washed and incubated for an additional 10 minutes to allow de-esterification of the dye. The assay was started with 1 ml of Calcium free Ringer’s buffer in the imaging dish and images were captured every 4 seconds. After capturing the baseline for ∼60 seconds, stores were depleted by the addition of 1 μM Thapsigargin. In other assays, 10 μg/ml of anti-CD3 and 5 μg/ml of the secondary antibody were used to cross-link the TCRs and thereby induce store depletion. ∼5 minutes post-store-depletion, Calcium Chloride (CaCl2) was added back to the dish at a final concentration of 2 mM to estimate the magnitude of store operated calcium entry.

Degranulation assay

PBMCs were cultured in RPMI containing 10% FBS and 50 ng/ml IL-2 and stimulated with PHA (2 μg /ml) 48hrs before the assay. 24hrs prior to the assay, IL-2 was washed off but PHA was re-added. On the day of the assay cells were washed twice to remove PHA and any growth factors. To measure degranulation, cells were either left unstimulated, stimulated with a combination of anti-CD3 (10 μg/ml plate-coated), anti-CD28 (2 μg/ml soluble) and anti-CD49d (2 μg/ml soluble) or a combination of Ionomycin (1 μM) and PMA (50 ng/ml) for 3.5 hours. CD107a-PE antibody (1:50 dilution) was added to each of the three groups at the start of stimulation. Following stimulations, cells were transferred to ice, washed with cold HBSS containing 2% FBS and incubated with anti-CD8 APC for 20 minutes, washed twice with cold HBSS and analyzed using Cytoflex (Beckman Coulter) flow cytometer.

qPCR to estimate gene expression in human PBMCs

PBMCs were taken off IL-2 48 hours prior to the assay. On the day of the assay, cells were washed twice, left unstimulated and either stimulated with anti-CD3 (10 μg/ml, plate-coated), anti-CD28 (2 μg/ml) and anti-CD49d (2 μg/ml) or with Ionomycin (1 μM) and PMA (50 ng/ml) for 6 hours. To end the stimulation, cells were washed with cold HBSS, pelleted and used for RNA extraction. Total RNA was isolated using Qiagen RNeasy mini kit and cDNA was synthesized using random hexamers and Superscript IV (Invitrogen). cDNA was quantified using Qubit and subjected to QPCR analysis using Taqman probes for IL-2 and beta-actin housekeeping gene control in triplicates. Analysis was performed by calculating the double delta Ct values.

Transduction of human PBMCs with pMSCV-STX11

For viral transduction of PBMCs, polystyrene non-TC treated 24-well plates were coated with retronection (20 μg/ml) overnight at 4°C, blocked with 2% BSA for 30 minutes at room temperature and washed twice with HBSS. The retroviral supernatants were added onto the coated wells and the plates were spun at 1800g for 2 hours at 30°C. Following spin, the wells were washed with the blocking solution. PBMCs cultured in RPMI containing 10% FBS were stimulated with PHA (2 μg/ml) and IL-2 (50 ng/ml) for 48 hours prior to transduction. Stimulated PBMCs were transferred to coated plates at a density of 0.25 million cells/well centrifuged at 400g for 40 minutes at 30°C. The cells were analyzed 48-72 hours post transduction.

Patch clamp measurements

Patch clamp recordings were performed by using an Axopatch 200B amplifier (Axon Instruments, Foster City, CA) interfaced to an ITC-18 input/output board (Instrutech, Port Washington, NY) and an iMac G5 computer. Currents were filtered at 1 kHz with a four-pole Bessel filter and sampled at 5 kHz. Recording electrodes were pulled from 100-µl pipettes (VWR), coated with Sylgard, and fire-polished to a final resistance of 2 to 5 MΩ. Stimulation and data acquisition and analysis were performed by using in-house routines developed by RS Lewis (Stanford University) on the Igor Pro platform (Wavemetrics, Lake Oswego, OR). All data were corrected for the liquid junction potential of the pipette solution relative to Ringer’s in the bath (−10 mV) and for leak currents collected in 20 mM extracellular free calcium plus 10µM La3+. The holding potential was +30 mV. The standard voltage stimulus consisted of a 100-ms step to –100 mV followed by a 100-ms ramp from –100 to +100 mV applied at 1 s intervals. ICRAC was typically activated by passive depletion of ER Ca2+ stores by intracellular dialysis of 8 mM BAPTA.

Solutions and chemicals

The standard extracellular Ringer solution contained 130 mM NaCl, 4.5 mM KCl, 20 mM CaCl2, 10 mM tetraethylammoniumchloride (TEA-Cl), 10 mM D -glucose and 5 mM HEPES (pH 7.4 with NaOH). The standard divalent-free (DVF) Ringer solutions contained (in mM) 150 NaCl, 10 HEDTA, 1 EDTA, and 10 HEPES (pH 7.4). The internal solution contained: 135 mM Cs aspartate, 8 mM MgCl2, 8 mM Cs-BAPTA and 10 mM HEPES (pH 7.2 with CsOH).

Data analysis

Analysis of current amplitudes was typically performed by measuring the peak currents during the −100 mV pulse. Averaged results are presented as the mean value standard error of the mean (SEM). Reversal potentials were measured from the average of several leak-subtracted sweeps (4–6) in each cell. For datasets with two groups, statistical analysis was performed with two- tailed t test to compare between control and test conditions with a confidence level of 95%, and results with p<0.05 were considered statistically significant.

STX11:Orai1 and SOAR:Orai complex prediction

Full length STX11 and its two domains, (Habc 41-167 and SNARE 183-267) and Orai1 (N-terminus 1-87 and C-terminus 256-301) were used to generate the complex of STX11 and Orai1 using AlphaFold3 (AF3) in all combinations. For Stim1, SOAR, residue 344-444 were used along with Orai C-terminus 256-301. The resultant models were assessed using ipTM and pTM scores. The best scoring combinations were further considered to generate more models by changing the seed values. The best model in terms of highest ipTM and pTM values was considered for further analysis. Custom script was used to analyze the contact frequency of interacting residues of STX11-Habc and Orai1 C-terminus complex as well as SOAR and Orai1 C-terminus, across all predicted models.

Molecular dynamics (MD) simulation

The STX11 Habc:Orai1 C-terminus and SOAR:Orai1 C-terminus complexes were subjected to all-atom MD simulations. Initially, the complexes were prepared using protein-preparation module of Schrodinger which involves H-bond network optimization and restrained minimization of the initial structure. The prepared structures were solvated using TIP3P water model in an orthorhombic box and neutralized by adding counter ions. OPLS4 force field was used, and simulation system was generated by specifying 150 mM salt (NaCl). The solvated system was subjected to default relaxation protocol of Desmond followed by production run for 500 ns at 300K and 1 atm pressure in NPT ensemble. The default relaxation protocol includes several short simulation steps. (1) Brownian dynamics simulation for 100 ps at 10 K temperature in NVT with restraints on solute heavy atoms (2) Simulation in NVT ensemble at 10 K for 12 ps with restraints on solute heavy atoms (3) 12 ps simulation in NPT ensemble at 10 K with restraints on solute heavy atoms (4) Simulation in NPT ensemble for 12 ps with restraints on solute heavy atoms (5) Simulation in NPT ensemble for 24 ps without restraints. Three independent runs were executed with different initial seeds for both structures.

Analysis of MD simulation

The MD runs were analyzed for the stability of the complex and interactions between the subunits. The RMSD was calculated using simulation interaction diagram (SID) module of Schrodinger. The first frame was used as reference frame for RMSD calculation. Script “analyze_trajectory_ppi.py” was used to calculate the interactions between STX11:Orai1 and Stim1:Orai1. The binding energy (ΔG) at each nanosecond across simulation was calculated using “thermal_mmgbsa.py” script of Schrodinger. The representative structure was generated through trajectory clustering. The last 200 ns frames from each run were combined and clustered using scripts “trj_merge.py” and “trajectory cluster” of Schrodinger. The cluster representative of the largest cluster was used to analyze the protein-protein interactions. PDB Sum was used to calculate the interactions in the cluster representative. Figures were generated using Pymol and MD movies using Maestro.

Statistical Analysis

Statistical analysis was done using unpaired two-tailed Student’s t test where only two groups were being compared and ANOVA where more than two groups were present. Normality of distribution was checked in Origin. Mann Whitney U Test was used for datasets with non-normal (or skewed) distribution. P values for all significance tests were *P<0.05; **P<0.01; ***P<0.001.

eLife Assessment

This study reports a novel function for syntaxin 11, a specialized SNARE protein critical for the immune system whose mutations cause familial hemophagocytic lymphohistiocytosis type 4. The data convincingly show that depletion of STX11 impairs store-operated calcium entry in Jurkat T cells and that this defect is recapitulated in primary cells from a patient suffering from the disease; the authors further show that the syntaxin interacts with the pore subunit of the ORAI1 channel and propose that it primes the channel by promoting the assembly of multimers before activation by its endogenous ligand, the ER Ca2+ sensing protein STIM1. This is a conceptually important claim that challenges the prevailing view that all structural transitions in ORAI1 are STIM-driven. The data are high-quality and broadly consistent with the interpretation, but alternative mechanisms for the defects are not considered; additional work should rule out vesicular trafficking, discuss other mechanisms, and address methodological issues.

Reviewer #1 (Public review):

Summary:

Patients with STX11 mutations develop familial hemophagocytic lymphohistiocytosis Type 4, a fatal immune disorder marked by defective T and NK cell cytotoxicity and cytokine storm. The conventional explanation attributes this to impaired cytotoxic granule release, but this has never fully accounted for the broader disease picture. This study proposes an alternative mechanism. The authors show that STX11 is required for store-operated calcium entry through ORAI1 channels, which are essential for both cytotoxic killing and NFAT-driven gene expression in T cells. In STX11-deficient cells, ORAI1 currents drop, NFAT nuclear translocation fails, IL-2 expression is suppressed, and degranulation is impaired. These defects are largely rescued by ionomycin or a constitutively active ORAI1 mutant, placing the primary lesion at calcium signaling rather than the fusion machinery. Mechanistically, STX11 binds the C-terminal tail of ORAI1 via its Habc domain and maintains ORAI1 in a state competent for productive assembly prior to STIM1-dependent gating, a step the authors call "priming."

Strengths:

The paper identifies a novel and disease-relevant role for STX11 in calcium channel regulation and raises the possibility of using channel agonists as a therapeutic strategy in the disease. The biochemical and functional data are of high quality and generally consistent with the interpretation. The proposal that a non-conventional syntaxin directly interacts with ion channels to prime its activation is novel and interesting.

Weaknesses:

For readers to appreciate the value of patient experiments derived from a single individual, the authors should quote prior studies showing that STX11 protein levels are abolished in all known human STX11 mutations. The priming model, while functionally well-supported, rests on indirect structural evidence, and the precise conformational transition involved remains to be defined. These are acknowledged limitations, but alternate mechanisms have not been explored and formally excluded. More direct evidence should be provided to exclude the possibility that STX11 could act as a conventional SNARE and sustain calcium fluxes by promoting the delivery of additional ORAI1 channels from vesicles.

Reviewer #2 (Public review):

Summary:

Vig's lab delineates a critical role for STX11 in CRAC channel function, particularly in the context of the fatal immune disorder familial hemophagocytic lymphohistiocytosis type 4 (FHL4). They demonstrate that Syntaxin 11 directly binds and regulates Orai1, and that STX11 depletion abolishes CRAC currents and downstream signaling. Loss of STX11 reduces IL2 gene expression and impairs degranulation, both of which are rescued by the constitutively active Orai1 mutant H134S, whereas a gain‑of‑function mutant targeting the C‑terminus fails to restore these defects. The authors conclude that STX11 primes Orai1 for optimal local assembly that is independent of STIM1 yet required for CRAC channel gating.

Strengths:

This study is firmly grounded in disease biology and demonstrates that STX11 downregulation leads to profound functional defects. Using a comprehensive suite of methods and analyses, the authors interrogate the co-regulation of STX11 and Orai1 and present a near-complete view of STX11's modulatory role in CRAC channel function and downstream signaling pathways. The figures are clear, and the statistical analyses are rigorous and convincing.

Weaknesses:

The authors conclude that Syntaxin 11 directly binds Orai1. This conclusion is well supported by a multifaceted approach, including co-immunoprecipitation (co-IP), molecular dynamics simulations, co-localization/FRET assays, and targeted mutational analysis-all of which are thoroughly executed. While the interaction appears reasonably strong in co-IP experiments, the STX11-Orai1 interaction is comparatively weaker in pull-down assays, which the authors attribute to instability of the purified His-STX11 protein. A remaining gap is direct evidence of interaction in live cells; this is understandably challenging given that fluorescent tagging of STX11 is not feasible. Fully resolving this question lies beyond the scope of the present study and will require more advanced approaches to capture STX11 binding dynamics.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

This study reports a novel function for syntaxin 11, a specialized SNARE protein critical for the immune system whose mutations cause familial hemophagocytic lymphohistiocytosis type 4. The data convincingly show that depletion of STX11 impairs store-operated calcium entry in Jurkat T cells and that this defect is recapitulated in primary cells from a patient suffering from the disease; the authors further show that the syntaxin interacts with the pore subunit of the ORAI1 channel and propose that it primes the channel by promoting the assembly of multimers before activation by its endogenous ligand, the ER Ca2+ sensing protein STIM1. This is a conceptually important claim that challenges the prevailing view that all structural transitions in ORAI1 are STIM-driven. The data are high-quality and broadly consistent with the interpretation, but alternative mechanisms for the defects are not considered; additional work should rule out vesicular trafficking, discuss other mechanisms, and address methodological issues.

We thank the editor and reviewers for assessing our work. We have now included additional experiments in a new main Figure 2, which directly rule out any general or Orai1 plasma membrane trafficking defects in Syntaxin11-depleted cells. There are additional experiments and/or analysis in many other figures, throughout the paper. We have included new and missing methods, quantifications and calibrations, and provided response to each of the reviewer’s comments below.

Public Reviews:

Reviewer #1 (Public review):

Summary:

Patients with STX11 mutations develop familial hemophagocytic lymphohistiocytosis Type 4, a fatal immune disorder marked by defective T and NK cell cytotoxicity and cytokine storm. The conventional explanation attributes this to impaired cytotoxic granule release, but this has never fully accounted for the broader disease picture. This study proposes an alternative mechanism. The authors show that STX11 is required for store-operated calcium entry through ORAI1 channels, which are essential for both cytotoxic killing and NFAT-driven gene expression in T cells. In STX11-deficient cells, ORAI1 currents drop, NFAT nuclear translocation fails, IL-2 expression is suppressed, and degranulation is impaired. These defects are largely rescued by ionomycin or a constitutively active ORAI1 mutant, placing the primary lesion at calcium signaling rather than the fusion machinery. Mechanistically, STX11 binds the C-terminal tail of ORAI1 via its Habc domain and maintains ORAI1 in a state competent for productive assembly prior to STIM1-dependent gating, a step the authors call "priming."

Strengths:

The paper identifies a novel and disease-relevant role for STX11 in calcium channel regulation and raises the possibility of using channel agonists as a therapeutic strategy in the disease. The biochemical and functional data are of high quality and generally consistent with the interpretation. The proposal that a non-conventional syntaxin directly interacts with ion channels to prime its activation is novel and interesting.

Weaknesses:

For readers to appreciate the value of patient experiments derived from a single individual, the authors should quote prior studies showing that STX11 protein levels are abolished in all known human STX11 mutations. The priming model, while functionally well-supported, rests on indirect structural evidence, and the precise conformational transition involved remains to be defined. These are acknowledged limitations, but alternate mechanisms have not been explored and formally excluded. More direct evidence should be provided to exclude the possibility that STX11 could act as a conventional SNARE and sustain calcium fluxes by promoting the delivery of additional ORAI1 channels from vesicles.

In the revised version, we have included references for all those prior STX11 human mutations that have been biochemically characterized till date. The reviewer has correctly pointed out that STX11 protein levels were almost abolished in almost all previously reported mutations. See line 168-173. Therefore, the prior STX11 patient mutations are essentially comparable to the frameshift mutation characterized in this study, in terms of STX11 protein depletion and, therefore, the mechanisms underlying the phenotypic defects reported here as well as earlier. We, therefore, believe that our data from even a single FHLH4 patient, with severely depleted STX11 levels, and additional knockdown studies across three different cell lines, are representative of majority of STX11 mutant FHLH4 patients that have been previously characterized.

Regarding the Reviewers’ concern that absence of STX11 as a conventional SNARE could affect Orai1 channel delivery from intracellular vesicles. We would like to point out the following:

(1) In Miao et al. 2013 (1), Figure 3C-D, we showed that expression of a dominant-negative mutant of NSF, a non-redundant protein in vesicle trafficking, impaired vesicle trafficking but did not affect SOCE. This experiment had essentially ruled out a role for vesicle trafficking in SOCE. In the same paper, we had also shown that Orai1 levels in the PM do not increase post-store depletion (Figure 3-figure supplement 2).

(2) SNAP23/25 form a four helical bundle with R- and Q-SNAREs in orchestrating vesicle fusion. In this paper, we have ruled out a direct role for SNAP23, SNAP25 and SNAP29 in SOCE (Figure 2-figure supplement 3).

(3) In v1 of this manuscript, we had shown that U2OS cells stably expressing Orai1-BBS-YFP have identical levels of Orai1 in the PM with and without STX11 depletion (Supplementary Figure 3B). This showed that the biosynthesis or delivery of Orai1 to the PM is not affected by STX11 depletion. The levels were also assessed in store-depleted U2OS cells but not included because in Miao et al. 2013 we had already established that levels of PM Orai1 remain essentially equal in resting versus store-depleted cells.

In the revised version, we have included the data from store-depleted cells in U2OS and also done quantification of PM Orai1 in HEK293 and Jurkat T cells. In addition, we have added three independent membrane trafficking/ vesicle secretion assays performed in STX11-depleted cells (new Figure 2 and associated supplements). In all cases, we find no evidence of Orai1 in intracellular vesicles or a general defect in membrane trafficking/ secretion in STX11-depleted cells. Orai1 is constitutively and stably expressed in the PM in resting as well as store-depleted cells in three different cell lines.

(4) Most importantly, in Figure 7I-J of this manuscript, we showed that calcium influx from a constitutively active mutant Orai1 (Orai H134S) is identical between STX11-depleted and scramble control cells. If wildtype Orai1 was indeed stuck in vesicles in STX11-depleted cells, then how would mutant H134S Orai1 be able to rescue the defect in SOCE? We have included the quantification of PM levels of Orai1 mutants w.r.t WT Orai1 in new Figure 8-figure supplement 3B and 3D.

In summary, we have now done several new experiments to directly measure Orai1 levels in the PM and general vesicle trafficking assays in HEK293 and Jurkat T cells and have found no defects in these upon STX11 depletion.

Regarding STX11 induced precise conformational transition, we are trying to setup collaborations with scientists who might be able to visualize this in situ. Please note that while purification of isolated pore subunits of ion channels followed by crystallization or expression in synthetic membranes for cryo-EM is currently considered a gold standard in the analysis of ion channel pore subunits, we have shown that ion channels are dynamic macromolecular complexes, in vivo (2), where synaptic proteins dynamically bind to induce conformational changes and affect their stoichiometry (2). Please also see (3) and (4). More advanced approaches, therefore, need to be developed to enable visualization of the dynamics of ion channel macromolecular complexes in their native environment in situ. In the absence of such approaches, the structural insights obtained from detergent-purified isolated subunits will remain incomplete.

Reviewer #2 (Public review):

Summary:

Vig's lab delineates a critical role for STX11 in CRAC channel function, particularly in the context of the fatal immune disorder familial hemophagocytic lymphohistiocytosis type 4 (FHL4). They demonstrate that Syntaxin 11 directly binds and regulates Orai1, and that STX11 depletion abolishes CRAC currents and downstream signaling. Loss of STX11 reduces IL2 gene expression and impairs degranulation, both of which are rescued by the constitutively active Orai1 mutant H134S, whereas a gain‑of‑function mutant targeting the C‑terminus fails to restore these defects. The authors conclude that STX11 primes Orai1 for optimal local assembly that is independent of STIM1 yet required for CRAC channel gating.

Strengths:

This study is firmly grounded in disease biology and demonstrates that STX11 downregulation leads to profound functional defects. Using a comprehensive suite of methods and analyses, the authors interrogate the co-regulation of STX11 and Orai1 and present a near-complete view of STX11's modulatory role in CRAC channel function and downstream signaling pathways. The figures are clear, and the statistical analyses are rigorous and convincing.

Weaknesses:

The authors conclude that Syntaxin 11 directly binds Orai1. This conclusion is well supported by a multifaceted approach, including co-immunoprecipitation (co-IP), molecular dynamics simulations, co-localization/FRET assays, and targeted mutational analysis-all of which are thoroughly executed. While the interaction appears reasonably strong in co-IP experiments, the STX11-Orai1 interaction is comparatively weaker in pull-down assays, which the authors attribute to instability of the purified His-STX11 protein. A remaining gap is direct evidence of interaction in live cells; this is understandably challenging given that fluorescent tagging of STX11 is not feasible. Fully resolving this question lies beyond the scope of the present study and will require more advanced approaches to capture STX11 binding dynamics.

We thank the reviewer for acknowledging that analysis of the dynamic binding of STX11 will require standardization of advanced techniques which are beyond the scope of the present study. We plan to continue developing methods that will allow us to visualize the binding and unbinding of STX11 to Orai1 in vivo.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Mechanistic issues:

(1) More direct evidence should be provided to exclude the possibility that STX11 could act as a conventional SNARE and sustain calcium fluxes by promoting the delivery of additional functional channels, stored in secretory vesicles or recycling endosomes, to the membrane. A significant fraction of the ORAI1 channel is in vesicles, and the mobilization of this intracellular pool regulates the rates of calcium fluxes in HEK-293 cells (PMID 26116575) and effector T cells (PMID: 35217583). Mobilization of this pool could account for part or all of the functional effects reported here. The only evidence that STX11 depletion does not impact the plasma membrane availability of the channel relies on one single flow cytometry profile (Supplementary Figure 3). This experiment is performed in U2OS cells stably expressing a fusion protein containing an extracellular bungarotoxin site. This cellular system is only used here; the other data are obtained either in HEK-293 cells or in Jurkat T cells. The level of endogenous STX11 in U2OS cells is unknown, and the efficiency of protein depletion has not been assessed. The efficiency of depletion should be shown, and the total number of channels assessed by comparing the expression levels of permeabilized and non-permeabilized cells. A flow cytometry profile of cells treated with thapsigargin should be included to match the experimental conditions of functional recordings. It would be valuable to repeat this experiment in Jurkat T cells by expressing ectopically the channel tagged on the extracellular side. To further exclude the involvement of vesicular trafficking, the authors should also evaluate the contribution of VAMP8, as this R-SNARE has been proposed to interact with STX11 to regulate the exocytosis of specialized granules in cytotoxic T cells (PMID: 26124288).

We have not observed significant levels of intracellular Orai1, as described in the PMID 26116575 paper. Multiple technical reasons could explain the artefactual appearance of intracellular Orai1. HEK293 is an embryonic kidney cell line, with most cells showing a distinct spindle shape and filopodia as shown on the ATCC website (https://www.atcc.org/products/crl-1573). The HEK cells shown throughout the Hodeify et al. 2015 paper (PMID 26116575) lack this typical morphology that majority of the HEK cells should show. Transfecting cells with high amounts of DNA and liposomes can severely affect the health and morphology of the cells leading to artifacts where PM proteins appear to be stuck intracellularly. The plasma membrane of the cell in movie 1 of PMID 26116575, for instance, also shows membrane blebs or membrane ruffles. The authors should have used a PM marker such as WGA (wheat germ agglutinin) to distinguish PM Orai1 from any intracellular Orai1 to establish whether what appear as intracellular Orai1 vesicles are not blebs of PM Orai1. Similarly, no endo/exocytotic vesicle marker is used to distinguish them from Orai1’s presence in apoptotic vesicles of unhealthy cells. In view of the overall abnormal morphology and any PM or intracellular vesicle marker, the claim that Orai1 resides in intracellular vesicles is unfounded.

Similarly, the PMID: 35217583, quoted by reviewer #1, is about in vitro differentiated primary mouse T cells. The study lacks any detail of the generation of the HA-tagged mouse Orai1 plasmid, used in the study, or even a back reference to show how this plasmid was validated for normal expression in primary T cells earlier. Ectopic expression of CMV promoter-driven plasmids in mouse primary T cells is extremely challenging and often results in poor cell health and incomplete and selective expression in only 5-10% of cells. It is unclear if the same HA-tagged human Orai1 plasmid that was used in PMID 26116575 is being used in this study to express in primary mouse cells. In the absence of all this information, it is unclear whether there was an issue with the generation of a new HA-tagged mouse Orai1 construct, which made the protein get stuck intracellularly. Or potentially the expression of a human protein in mouse primary T cells is the problem. The functional verification of the construct by showing rescue of SOCE in Orai1-deficient primary mouse cells is an absolutely essential control but is missing. In the absence of these controls, one cannot disregard decades of robust data on the localization of Orai1 in the PM from multiple labs and papers that have established its PM localization conclusively (5) (6).

Most importantly, in both the PMID 26116575 and PMID: 35217583, HA tagged-Orai1 is detected using a bivalent antibody, followed by secondary antibody. It is well known that cross-linking of cell surface receptors or proteins using bivalent antibodies has a major caveat that involves antibody-mediated clustering and capping, especially in lymphocytes, which typically induces rapid internalization of the entire antigen-antibody complex. The paper by Sekine-Aizawa et al., in 2004, showed that tagging of receptors/ channels with bungarotoxin binding site (BBS) followed by labelling with bungarotoxin (BTX) bypasses this confounding factor and, therefore, allows accurate estimation and localization of PM versus intracellular proteins. The artefactual bivalent antibody-induced endocytosis continues even when cells are incubated on ice as endocytosis is only slowed but not stopped on ice.

Due to this challenge, we have generated BBS tagged Orai1-YFP. We use flow cytometry to show PM Orai1 because it is an unbiased and quantitative way of showing surface Orai1 expression (estimated by measuring the intensity of surface-bound BTX), simultaneously, in thousands of cells with no potential for visual bias in the selection and imaging of cells. BTX labelling is done on ice and cells are washed and fixed right after labelling with BTX-A647 to stop endocytosis. In the revised version, we have used both complementary approaches of flow cytometry and microscopy, showing representative images of cells, alongside quantifications. All new experiments were performed in HEK293 and Jurkat T cells to estimate surface versus total expression in a new main Figure 2. The post-store-depletion data have been added to the existing U2OS experiment in new Figure 2-figure supplement 2A-E.

Our BBS-tagged Orai1 construct also has a YFP tag at the C-terminus. Since the flow cytometry experiment involves gating on YFP-positive cells, the total number of channels (biosynthesis) can be compared by looking at the YFP intensities in the scramble versus STX11-depleted groups. These data have now been included in the previous and new experiments. In none of the cases could we detect any difference in YFP or BTX-A647 intensities, pre- or post-store-depletion in scr or STX11-depleted cells which would indicate defects in biosynthesis or Orai1’s presence in vesicles in any group.

Regarding a role for VAMP8, in Miao et al. eLIFE 2013 (1), Figure 3C-D, we showed that expression of a dominant negative mutant of NSF, a non-redundant protein in the vesicle trafficking pathway, impaired Transferrin receptor recycling within 20 hours but did not affect SOCE at all. This experiment had conclusively ruled out any role for vesicle trafficking in SOCE and therefore assessment of the role of each of the individual proteins involved in membrane trafficking becomes redundant. We have additionally ruled out a role for SNAP23/ SNAP25/ SNAP29 (old supplementary Figure 12, new Figure 2-figure supplement 3A-C). SNAP23/25 form a four helical bundle with most R-SNARE and Q-SNAREs to orchestrate vesicle fusion. R-SNAREs, typically, need SNAP23/25 to interact with Q-SNAREs. Since a role for these non-redundant proteins has also been ruled out by us, it is unlikely that VAMP8 plays a role in modulating the effects of STX11 in SOCE.

(2) Both ORAI1 and STX11 are S-Acylated on cysteine residues, and this post-translational modification promotes their recruitment to the immune synapse (PMID: 24910990, 34913437). One possibility that should be discussed is that STX11 could enhance the recruitment of the ORAI1 channel into lipid domains rich in cholesterol, thereby favoring its activation. This type of priming would still require direct interaction between the two proteins but involve a different mechanism than the one discussed by the authors. This could be experimentally tested by expressing a STX11 mutant lacking the cysteine residues required for its S-Acylation. It would also be interesting to test whether depletion of STX11 impairs the recruitment of ORAI1 to the immune synapse forming between Jurkat T cells and antigen-presenting cells.

In our experiments reported in this paper, we have used soluble anti-CD3 as well as plate-coated anti-CD3 in combination with soluble anti-CD28 to stimulate Jurkat and primary T cells. These antibodies are routinely used to stimulate T cells and this type of stimulus doesn’t depend on the formation of a classical immune synapse with an antigen-presenting cell (APC). Despite the absence of synapse, the T cells get fully activated and functional, as seen by NFAT translocation and secretion of cytokines, such as IL-2 in new Figure 4. Therefore, whether there is a defect in the recruitment of Orai1, or STX11, to T cell synapse formed with an APC is not within the scope of this study. Furthermore, accurate analysis of protein localization within the immune synapse requires a dedicated study employing sub-diffraction resolution microscopy approaches.

Regarding PMID: 24910990, and the mechanism of recruitment of STX11 to the membranes. We believe this remains unknown. The frameshift mutant used in our study lacked all terminal cysteines, which have been earlier proposed to be crucial for membrane targeting, as well as a terminal part of the SNARE domain and yet it localized to the PM just as well as wild-type STX11 (See new Figure 5E). We have added this result in the text line 302-305 and removed the line stating that post-translational modifications of the terminal cysteines target STX11 to the PM as previously claimed in PMID: 24910990 and mentioned in v1 of this paper. In view of these new data, it currently remains unknown whether potential attachment to PIP2 in PM via various basic residues, spread throughout the sequence (7, 8), or binding to another protein targets STX11 to the PM (PMID: 26771955). There is no obvious poly-basic stretch in STX11 sequence, therefore, a systematic and focused deletion and mutagenesis study will be needed to individually assess the above possibilities which is outside the scope of the present study.

Methodological issues

(3) Since STX11 colocalize with Orai1 already in basal conditions, independently of STIM1, it could influence basal calcium levels. This cannot be appreciated from the data presented, because all the SOCE protocols start in calcium-free conditions, preventing baseline comparison between WT and STX11-deficient cells. A potential difference in basal calcium levels should be explored, and the impact of STX11 depletion on basal calcium fluxes should be documented by calcium shifts (2 mM → 0 mM → 2 mM) or manganese quenching approaches.

The cells are typically loaded with Fura2 in 2mM calcium containing Ringer’s buffer. We switch the cells to 0mM right at the start of the SOCE protocol and start imaging within 5-10 seconds. Therefore, in our experience, the baselines of scramble versus STX11-depleted cells should show a difference even in the SOCE protocol if the basal calcium levels are affected because Fura2 is already present and bound to basal calcium present in the cytosol at the start of the protocol. Still, we have done the experiment suggested by the reviewer as shown in Author response image 1. The assay started with cells in 2 mM extracellular Ca2+ followed by addition of 10 mM EGTA, which according to the following equation quenches the 2 mM extracellular Ca2+ (https://somapp.ucdmc.ucdavis.edu/pharmacology/bers/maxchelator/CaEGTA-TS.htm).

where, [Ca2+]Free is the free/unbound Ca2+, [Ca2+]Total is the total Ca2+, [EGTA]Total is the total EGTA concentration and Kd is the dissociation constant between Ca2+ and EGTA at 37°C and pH 7.4.

To confirm complete sequestration of the extracellular Ca2+, we also repeated the assay with 20 mM EGTA but did not notice any difference between the 10 mM and 20 mM EGTA conditions. We do not see any differences in basal calcium, under any condition, between Scr and STX11 shRNA treated cells HEK or Jurkat T cells.

Representative Fura-2 traces of Scr (black) and STX11 (red) shRNA-treated HEK293 (A) and Jurkat (B) cells, where the cells were incubated with 2 mM Ca2+ followed by addition of 10 mM EGTA to quench the 2 mM Ca2+.

Since we do not have access to a perfusion system, we could not test Fura2 response after re-addition of 2mM calcium to the existing EGTA and Ca2+ mixture. However, the transition from 0mM to 2mM is already shown in Figure 8.

(4) The quantification of the calcium imaging data is problematic and requires clarification. In most figures, the data are shown normalized to the control condition. According to the method section (lines 721-724), 100% is the maximum value of the scramble shRNA group (amongst the three experiments). But what was measured here? The slope during calcium readmission? The peak amplitude after calcium readmission? Expressed as a ratio or as calcium values? The recordings are presented in micromolar calcium concentration. This implies calibration, but a calibration procedure is not mentioned. Please clarify. For the recordings of constitutive calcium entry in Figure 7, this normalization is not performed, and the data are expressed as ratio values. Here, it looks like the parameter quantified and compared is the absolute ratio value after calcium readmission. This is inappropriate. The trace in Figure 7G shows that the basal levels differ by more than two ratio units between control and STX11-depleted cells. Normalizing the data in Figure 7G to the basal ratio value would show no difference in the peak response amplitude between the two conditions. These data should be re-analyzed, and both the slope and the amplitude of the response should be presented, with statistics performed on independent recordings, not on individual cells pooled from different experiments. Cells from the same recording are not experimentally independent samples but rather replicates of the same experiment.

We have updated the relevant method section with more details in lines 823-858. The peak amplitude after calcium readmission was measured and compared to the baseline as described in the updated methods. The Fura 2 calibration method has also been added to the revised version, we apologize for this omission earlier. Separate calibration was done for Fura 2 experiments in all figures except Figure 7 from version 1 (new Figure 8). The experiments in new figure 8 were done using a different objective (20X, water) and, therefore, although a separate calibration was done for these experiments it was not applied to the data. We apologise for this omission on our part and have now applied the respective calibration to these experiments.

The trace in Figure 7G of version 1 should look different even in 0mM calcium in our opinion. The reason is the same as that explained in point 3 above; CAD-mediated constitutive activation of Orai1 is one of the strongest. The cells, even when they are being loaded with Fura2 in Ringer’s buffer with 2mM calcium, are constitutively recruiting calcium ions. This should result in a shift in Fura2 excitation due to higher levels of basal calcium. When the cells are switched to 0mM calcium and imaged within 4-5 seconds, the intracellular Fura 2 is still bound to all this extra calcium in the cytosol and therefore the baselines should show a significant difference. If the cells were imaged for several minutes in 0mM calcium, we might have seen the difference in basal ratios slowly reducing. However, we switched to 2mM calcium within 120sec. At this point any free Fura2 would be expected to bind incoming calcium again. For the same reason, the cytosol of Scr cells which would still have relatively higher levels of intracellular calcium concentration compared to STX11-depleted cells will show a smaller further increase in 2mM due to calcium-dependent inhibition of CRAC currents within the time frames we have measured. We have now added the Fura-2-calibrated response in the new Figure 8G-H. We have shown below the baseline-subtracted (normalized) response for Figure 8G. As you can see, there is still a significant difference in 2mM calcium between Scr and STX11-depleted cells but in this representation the important difference at 0mM is masked, we have therefore chosen to retain the original figure with Fura-calibrated values at 0 as well as 2mM calcium in Figure 8G-H.

The cells shown in the old Figure 7G of version 1 were not from the same recording but from three different experiments. Because the cells were imaged with 20X objective in these experiments to allow selection of Orai1-CFP or mutant Orai1-CFP and CAD-YFP double-positive cells, the number of cells analyzed per experiment was less compared to other experiments. We have now shown Fura 2 calibrated values in the new Figure 8G-L. We have also re-done the statistical analysis on three independent experiments from each. As shown in Author response image 2, the difference is still statistically significant whether we show merged cells from all three experiments or single representative experiment out of three repeats. We believe merged cells have more information to offer and therefore have retained the same figures with the original analysis in the main Figure 8G-L.

Box plots representing quantification of individual repeats of constitutive calcium influx in Scr and STX11 shRNA-treated HEK293 cells expressing YFP-CAD and Orai1-CFP without (A) and with baseline subtraction (B).

(C-D) Quantification of individual repeats of Scr and STX11 shRNA-treated HEK293 cells expressing Orai1-H134S (C) and Orai1-ANSGA (D) mutants.

(5) Quantification of pull-down experiments. The binding data in Figures 4F, 5F, and 5J are presented largely qualitatively. Densitometric quantification with statistical comparisons across wild-type and mutant conditions would make these results more convincing, particularly given that the authors themselves acknowledge the interaction appears relatively weak in vitro.

This has been done and included alongside the respective panels in the new Figure 6G and 6L (for old Figure 5F and 5J of version 1, where differences appeared relatively small in some experiments). The differences across lanes in both figures and their repeats were statistically significant.

Figure 4F showed a clear and visually significant difference in binding across lanes and repeats and therefore no quantification is needed for these experiments in our opinion.

Limitations of the study and mechanistic inferences.

(7) Interpretation of the ORAI:ORAI FRET and crosslinking data. STX11 depletion increases basal ORAI:ORAI FRET (Figure 7A-C) and shifts crosslinked species toward higher molecular weights (Figure 7D-E). The authors interpret this as ORAI1 being trapped in an unprimed state, but higher FRET and higher-order species would conventionally suggest increased rather than decreased assembly. The paper needs a clearer mechanistic explanation of what "unprimed" looks like structurally. Is this aberrant crowding, non-productive oligomerization, or something else? The distinction between a change in intermolecular distance within existing oligomers versus an increase in oligomer density matters here and should be addressed.

Higher ORAI: ORAI FRET and a shift in the size of crosslinked Orai1 oligomers, when analyzed together, suggests formation of ‘non-functional’ higher-order oligomers. Higher order does not necessarily translate to better function in the case of ion channels, it can also lead to non-selectivity or formation of ‘non-productive’ oligomers, as mentioned by the reviewer. It was shown by us earlier in Li et al. (2016) (2) that bigger oligomer size revealed by higher number of photobleaching steps of Orai1 did not translate to better function but led to non-selectivity.

In new Figure 8, crosslinking with BS3, which has a spacer arm and working distance of ~11 Å, very likely reflects a change in the number of subunits within individual oligomers and not crosslinking of independent existing oligomers. This is because we show that neither total Orai1 expression nor Orai1 expression in the PM change in any group in new Figure 2. FRET works best within 1 to 10 nm distance, and therefore, in theory, can lead to energy transfer between neighbouring Orai1 oligomers in high Orai1-expressing cells. However, because there was no change in Orai1 abundance in the PM (new Figure 2) or distribution within PM (new Figure 7E,F,J,L) of any group, FRET changes also likely reflect intra-oligomer changes rather than inter-oligomer interactions. FRET changes can also arise from a change in the respective orientation of fluorophore pairs but when analyzed together with crosslinking studies, changes in pore assembly likely coincide with conformational shifts in Orai1 protomers. Furthermore, FRET has been used earlier to show shifts in conformation of other ion channels (9). Therefore, we believe that, when used together, these two approaches strongly suggest an intermediate conformational state along with a change in number of Orai1 subunits per channel since there was no evidence of overcrowding in the PM or obvious segregation of Orai1 in specific regions of PM in new Figures 2 and 7E,F,J,L.

We could not assess whether the oligomers of Orai1 formed in the absence of STX11 possess an intact pore. The presence or absence of pore in STX11-depleted cells will require extraction of Orai1 oligomers from native membranes and performing systematic structural analysis using cryo-EM or related approaches which is outside the scope of this study.

(8) The ANSGA versus H134S discrepancy. H134S ORAI1 rescues calcium influx in STX11-depleted cells (Figure 7I-J), but the ANSGA mutant does not (Figure 7K-L). The authors conclude from this that STX11 induces molecular shifts within ORAI1 transmembrane helices, and not in its C-terminal tail. This is an important mechanistic inference that needs more discussion. What does this imply about the conformational state of primed ORAI1? And why is straightening of the tails not sufficient for full opening without the correct TM helix arrangement? This distinction has implications for how the STX11-ORAI1 interaction should be modelled and should be engaged with more thoroughly.

There is no discrepancy here, please also see our response to reviewer 2’s comment #4. The experiment implies that the conformational state of primed Orai1 involves shifts in the TM region of Orai1 and is different from unprimed state. The structural similarities between H134S and ANSGA Orai1 mutants have not been formally established. Unlike H134S, no structure exists for the ANSGA mutant. In the absence of this, it is impossible to comment on whether the two constitutively active mutants are structurally comparable or whether there are multiple ways to stabilize open states of CRAC channel pore, especially when using TM mutants of Orai1.

The goal of this experiment was to determine what kinds of structural shits STX11 potentially induces in native Orai1. Using previously characterized constitutively active mutants and fusion proteins from the CRAC field, we have ruled out a potential role for STX11 in simply changing the orientation of Orai1 C-terminal tails. A discussion on the topic of why tail straightening of Orai1 is insufficient to open Orai1 is outside the scope of this paper. As pointed by reviewer 2, it is possible that C-term tails already exist pointing towards the cytosol in native, resting Orai1, although this has not been shown in any study using structure of full-length WT Orai1 and is purely speculative at this point. We prefer to not engage in speculative structural insights.

Other points

(9) Figures 1G and 1H. The patient-derived mutant STX11 band runs at approximately 37 kDa rather than the predicted 39.5 kDa. The authors suggest instability or reduced antibody reactivity, but premature translation termination is also a possibility that should be acknowledged.

We have added this point in line 168.

(10) Figure 2B. The traces and current voltage relationships should be rescaled to show the rectification and inactivation profile of the current in cells depleted of STX11.

This has been done and modified in new Figure 3 (Figure 2 of version 1).

While preparing source data files for all figures, we noticed an error in the value of the SE in the STX11-depleted group of old Figure 2C, which has now been corrected. The SE value in the older version was erroneously pasted from an adjacent data column.

Similarly, in old Figure 3 (version 1), new Figure 4C, we noticed that some data points in the STX11 group were pasted twice in the same excel column. These cells were removed and additional cells were analyzed from the same experiment and added to this group. The overall result remains the same but the distribution of data points looks a bit different.

(11) Figure 4B: This experiment should be repeated in cells treated with thapsigargin to deplete intracellular calcium stores, and the extent of colocalization quantified by measuring the Pearson's correlation coefficient.

This has been done. Pearson’s correlation coefficient is included in new Figure 5D.

(12) Figure 4F. Why is there no detectable band in the input lane of the left blot?

Western blots show relative intensities of bands of proteins across lanes. A faint band in the input lane of old Figure 4F suggests that the IP/ co-IP/ pull down was robust. If we increase the exposure, the input band would become stronger but the pull-down band would become over-saturated and the difference in the intensities would not be linear. The faint non-specific bands in other lanes represent a fraction of soluble STX11 that tends to crash out of solution over time and gets spun down with the beads. See lines 535-541 explaining this.

(13) Figure 5. Immunofluorescence data showing the membrane staining of the mutated syntaxin and channel should be included, as well as calcium recordings of cells expressing YFP-CAD with WT and mutated ORAI1.

In version 2 Figure 6A, we have now also shown co-localization of mutant STX11 with Orai1-YFP in resting and store-depleted cells, in addition to WGA. Pearson’s correlation (not shown) did not show any significant difference in the localization of mutant synatxin 11 w.r.t Orai1. Calcium recordings of CAD-induced constitutive calcium influx from wild-type versus mutant Orai1 are now shown in new Figure 6O-P.

(14) Figure 6B. A clear colocalization of CFP-O1 and STIM1-YFP is visible on the images, yet the authors conclude from morphometric analysis that the channel is not recruited into ER-PM clusters. Please show the difference in colocalization quantified by measuring the Pearson's correlation coefficient. Pictures should also be provided with the C-terminally tagged construct.

The quantification of CFP-Orai1 localization inside Stim1-YFP puncta was already shown in old Figure 6E and F. The residence of Orai1 inside STIM1 puncta versus total Orai1 in the PM of STX11-depleted groups was clearly reduced. We have now also shown Pearson’s correlation coefficient for Stim Orai co-localization inside puncta in new Figure 7F.

TIRF microscopy images of C-terminally tagged Orai1 were already included in Supplementary Figure 10. No defect in co-clustering of C-terminally tagged Orai1-YFP and N-terminally tagged CFP-Stim1 was seen and yet SOCE was inhibited. Therefore, we never concluded from Figure 6 that Orai1 and Stim1 fail to co-localize. We said, they fail to form ‘functional’ clusters. We have now moved the representative TIRF images from supplementary figure 10 to the new main Figure 7G. The Pearson’s correlation coefficient for Stim Orai1 co-localization inside puncta is shown in new Figure 7L.

(15) Figure 6E and 6F show the same data.

Figure 6E showed fraction of Orai1 inside Stim1 puncta divided by total Orai1, and 6F showed fraction of Orai1 outside puncta divided by total Orai1. The plots are different but we agree that the data are coming from same cells. We have removed old panel 6F and replaced it with Pearson’s correlation coefficient of Stim1:Orai1 colocalization in puncta in new Figure 7F.

(16) Figure 7G-L. The difference in constitutive calcium fluxes should be confirmed by Manganese quench recordings. The surface expression of the Orai1 mutants should be shown.

We have now shown the quantification of surface expression of Orai1 mutants for each respective mutant in the new Figure 8-figure supplement 3B and 3D. The Orai1 mutants we have used in this paper are well established in the literature, they showed clear surface localization and the differences in calcium influx between Scr and STX11 treated cells upon overexpression of Orai1 mutants in HEK are robust. Therefore, we do not see any compelling reason for repeating all of the experiments from Figure 7G to 7L to also show manganese quench recordings, as suggested by the reviewer. We have applied Fura 2 calibration done for these experiments to calculate intracellular calcium. These have been shown in the revised and new Figure 8G-L, where F340/380 ratios of representative calcium assays have also been replaced with the calibrated intracellular calcium concentration.

(17) Supplementary Figure 12. The recordings show a very large variability between experiments. The different SNAREs that are depleted here could compensate for each other, accounting for this variability. It would be interesting to show the effect of the combined silencing of all the SNARES tested here. The efficiency of the protein depletion should also be documented.

Genome-wide high- or medium-throughput screens are inherently noisy. None of the genome-wide high- or medium-throughput screens show evidence of protein depletion for each gene in any of the published screens to our knowledge. We chose to only characterize the candidates that reproducibly showed > 70% inhibition of SOCE, others were ignored as noise.

Silencing of all SNAREs together will definitely lead to loss of morphology and early lethality as all membrane trafficking will be stopped. We never analyze cells that do not show normal morphology and have compromised viability for ablation of SOCE.

(18) Lines 236-238. The authors note that STX11 harbors a stretch of C-terminal cysteines proposed to be essential for its membrane localization, but do not elaborate on the underlying mechanism. It would strengthen the discussion to explicitly acknowledge that this membrane anchoring is mediated by S-acylation of these cysteines PMID: 24910990 and to connect this to the known enrichment of Orai1 in lipid rafts and the immune synapse PMID 34913437. Both observations are relevant to understanding how STX11 and Orai1 are brought into proximity at the plasma membrane, and their omission leaves an explanatory gap in the proposed interaction model.

Please see our response to point #2 above. We do not think C-terminal cysteines target STX11 to the PM. We have corrected this claim based on an earlier study, PMID: 24910990, in the revised version of this paper. Analysis of immune synapse and lipid rafts are outside the scope of this paper. The mechanism of PM targeting of STX11 is currently unestablished and will require a systematic and focused mutational analysis which is outside the scope and main focus of this paper.

(19) Line 351. The statement that syntaxin depletion does not alter the structure or proximity of junctional ER to the plasma membrane is not supported by data. Neither electron microscopy nor TIRF imaging has been performed, which would be required to back up this claim.

Because Stim1 itself can be used as a marker of ER-PM junctions, this statement was supported by data shown in Figure 6C, D, G, H of version 1 of this paper where the intensity and area of Stim1 clusters was assessed using TIRF microscopy and found to be indistinguishable between STX11 and scramble control cells. The imaging done in Figure 6G, H was TIRF imaging and this was already specified in the legend. We have now also done TIRF imaging of GFP-Mapper-expressing scr and STX11-depleted cells. Mapper is a genetically encoded fluorescent protein that was previously shown to mark ER-PM junctions (10). We found no significant difference in the area or intensity of GFP-Mapper puncta (new Figure 7O-Q), just like Stim1 puncta didn’t show any defect in STX11-depleted cells. Please see modified text from 416-423.

(20) The molecular dynamics methods need more detail: force field, simulation length, water box dimensions, and convergence criteria should all be specified to allow replication. The supplementary RMSD plots (Supplementary Figure 5B) should also show individual replicate trajectories rather than averages only.

We had already mentioned the force field (OPLS4) and simulation length (500ns) in the methods section. Also, the RMSD plots in Supplementary Figure 5B already showed individual replicates in version 1.

We have now updated the methods with following additions:

The OPLS4 force field was used for all 500 ns simulations in an orthorhombic water box with a buffer distance of 10 Å beyond the solute in each direction. Simulation stability was assessed based on the protein backbone RMSD over simulation time.

Trajectory clustering was performed using the trajectory clustering tool in Schrödinger, which applies affinity propagation to the pairwise backbone RMSD-based similarity matrix. Within each affinity propagation run, convergence was defined as no change in the set of exemplar frames for 15 consecutive iterations, with a maximum of 400 iterations per run. If convergence was not reached, the damping factor was increased from 0.5 in increments of 0.01 until convergence.

Reviewer #2 (Recommendations for the authors):

Overall, this is a timely and impactful study supported by a broad set of methods and cell types. Before publication, the manuscript should address the following points.

Major:

(1) The authors note that STX11 contains cysteine residues that enable membrane association. What is the specific mechanism of membrane attachment? Could it occur via S-acylation (palmitoylation)? Both Orai1 and STIM1 are known to undergo S-acylation, which raises the possibility that this modification might also facilitate STX11 membrane anchoring and/or co-residence with Orai1. Is STX11 constitutively membrane-associated, or does it show preferential localization to specific membrane subdomains, particularly in proximity to Orai1?

STX11 is constitutively membrane-associated and does not show any preferential localization to specific membrane subdomains in confocal images. Figure 4, panel A and B from version1 clearly showed this. In an earlier paper by Hellewell et al. 2014, PMID: 24910990, S-acylation of terminal cysteines of STX-11 was proposed to be crucial for membrane attachment of STX11 and its recruitment to the immune synapse. However, please see our response to reviewer 1’s comment #2 and a new Figure 5E for the localization of the frameshift FHLH4 mutant characterized in this paper. The frameshift mutant that we have characterized lacked all terminal cysteines as well as a short terminal part of the SNARE domain. Cloning and ectopic expression of this mutant still showed constitutive localization to PM and did not show preferential distribution to any specific regions. Therefore, we do not think that terminal cysteines of STX11 contribute to its membrane attachment, we have accordingly modified lines 291-293, 302-305, 540 in the revised version. Also see our response to your point#6 below.

(2) Is there a possibility to monitor a dynamic change in STX11 co-localization from before to after store-depletion?

We did not observe any change in the overall distribution of STX11 in cells expressing STX11 alone or co-expressing Orai1 with STX11, pre- or post-store-depletion (please see new figure 5B-C). In cells co-expressing ORAI1, STX11 and STIM1 (see new Figure 5M-N), we could not capture the dynamic segregation of STX11 into regions of PM devoid of STIM:ORAI puncta and therefore have only pre- or post-store-depletion images. Dynamic change in STX11 distribution would require live, multi-colour, high-resolution imaging of diffraction-limited ER-PM junctions and adjacent regions which is technically extremely challenging, and especially due to our inability to tag STX11 with a fluorescent tag without disrupting its localization. Also see our response to your point#6 below.

(3) The authors use CAD to prove that the interaction with the R289A_E272A_E275A_E278A mutant is normal as for the wild-type. Does this also hold for STIM1 wild-type full-length?

This is also true for full-length STIM1. The data have now been added to the new Figure 7-figure supplement1.

(4) The authors state that STIM1 binds both the N- and C-termini of Orai1. While STIM1 binding to the Orai1 C-terminus is well established, the nature of its interaction with the N-terminus remains debated. Fragment-based assays suggest direct binding to the N-terminus; however, direct interaction with full-length Orai1 has not been conclusively demonstrated. This point should be phrased more cautiously to reflect the current uncertainty.

We have re-phrased the sentence as follows in line 543: “The individual relevance of Orai1 N- versus C-terminus in the trapping versus gating of Orai1 remains unclear”

(4) In the discussion, the authors report: "Though crucial for trapping and gating, Orai1 tails were missing from early structures of Drosophila Orai [28]". A previous NMR structure suggested that the C-terminal tails of two adjacent Orai1 subunits bend and pair with each other in an antiparallel fashion, and sit closely apposed to PM [37]. However, in recent structures of constitutively active H134 mutant Orai, the C-terminal tails were found to orient away from the membrane [28]. In STX11-depleted cells, switching the CFP-tag from the Orai1 N- to the C-terminus could rescue its clustering but not gating by Stim1. Furthermore, STX11 depletion inhibited the constitutively active ANSGA mutant of Orai1 [29], where the tails of Orai1 are proposed to be constitutively unlatched. These data essentially reinforce our conclusions that STX11 induced molecular shifts encompass Orai1 transmembranes.' However, the information provided here is not fully correct. The early Drosophila Orai structure lacks the full N-terminus but retains most of the C-terminus. It was the X‑ray, not cryo‑EM, structure that suggested an antiparallel arrangement of the Orai1 C-termini. Although the "open" X‑ray structure shows unlatching and straightening of TM4-C-termini, it remains uncertain whether these features reflect physiological gating or crystallization artifacts. It is also unclear whether the Orai1 ANSGA gain‑of‑function mutant adopts a similar unlatching; however, prior work indicates that ANSGA impairs proper coupling to the C‑terminal binding interface (in contrast to Orai1 H134S, which maintains effective coupling). This raises the key question: why do H134S and ANSGA respond differently to STX11 depletion? One possibility is that these mutants stabilize distinct conformations of the TM4-C-termini ("latched" vs "unlatched" states) that differentially dictate the requirement for STX11 in channel assembly or gating. We recommend refining the discussion

We have changed the word ‘missing’ to ‘truncated’ in line 545 and 546.

We agree that there is no evidence in literature that establishes similarity between H134S and ANSGA mutation-induced conformations of Orai1. It is, however, implied in most previous studies of mutant Orai1s that there is only one possible open state/conformation. We have added the suggested point and modified the discussion in line 551-556.

(5) The authors highlight: 'A major problem with this interpretation is that even though amplification of CRAC currents was shown, none of the previous patch clamp studies established whether the higher currents resulted from a greater number of active channels or unchecked conductance per channel by performing single channel recordings.' It should be noted that CRAC channels have extremely low single‑channel conductance, making direct single‑channel recordings challenging. As a result, estimates of open probability and channel number typically rely on fluctuation (noise) analysis rather than direct measurements of single‑channel events (see https://doi.org/10.1085/jgp.200609588). We suggest acknowledging this limitation in the discussion to contextualize the interpretation of gating and channel density.

We acknowledge how challenging it is to record the single-channel conductance from CRAC channels. We have added this fact to the discussion and the reference that the reviewer has suggested in line 570-572.

(6) STX11 appears to shift Orai1 localization into puncta. Activated STIM1 is known to engage plasma membrane PIP2 to facilitate Orai1 coupling. How, if at all, is STX11 linked to PIP2 or PIP2-rich microdomains? Is there evidence for direct PIP2 binding by STX11, or for indirect recruitment via PIP2-binding partners? Any available data on STX11's lipid interactions or its enrichment within PIP2-enriched regions would help clarify this mechanism.

We have not claimed that STX11 shifts Orai1 into puncta. We already showed in old supplementary figure 10 and Figure 6G-J of version1 (v1) of this paper that the C-terminally tagged Orai1-CFP can very well form puncta and co-localize with STIM1 in STX11-depleted cells. To avoid this confusion, we have moved the old Supplementary Figure 10 from v1 to the main figure in revised version, see new Figure 7 panel G. Despite the presence of ORAI1-CFP in puncta with YFP-Stim1, the SOCE was inhibited in STX11-depleted cells. Please also see new Pearson’s correlation coefficient for Stim1 and Orai1 colocalization in Figure 7 panel L. Therefore we concluded that, Orai1 forms ‘nonfunctional’ clusters with Stim1 in STX11 depleted cells, please see modified lines 408-412, clearly explaining this.

Although syntaxin 1A has been shown to interact with cholesterol (11) as well as PIP2 (7, 8) using either a stretch of polybasic residues or basic residues spread throughout several domains. To our knowledge, these have not been proposed to recruit or segregate STX11 in membranes. STX11 doesn’t contain an obvious stretch of poly-basic residues in its sequence, either, to quickly mutate and address this question. Please also see our response to your point #1 and #2 above. Answering this question will require a systematic and dedicated mutagenesis study.

Minor:

(1) Please indicate in Figure 1 in the respective graphs in which cell type the Ca2+ imaging studies have been performed.

Done.

(2) Figure 4C, D: Why are the input bands so weak?

Please see our response to reviewer #1’s similar comment 12 above.

(3) Figure 5A: Please clarify what WGA is.

WGA is wheat germ agglutinin which is used to mark PM in imaging experiments. It binds to N-acetyl-D-glucosamine and sialic acid residues found in mammalian cell membranes and glycoproteins. We have added the explanation to the new Figure 6A legend.

The authors state that "the constitutively active ANSGA (261-265) mutant of Orai1 (Supplementary Figure 11G), which harbors 4 consecutive mutations in the Orai1 C-terminus ...". Please clearly state this is the nexus region connecting the C-terminus with TM4. The 5 aa stretch is not the C-terminus; it is just close to the C-terminus.

We have modified this, as suggested, in line 470-471.

References:

(1) Miao Y, Miner C, Zhang L, Hanson PI, Dani A, Vig M. An essential and NSF independent role for alpha-SNAP in store-operated calcium entry. Elife. 2013;2:e00802.

(2) Li P, Miao Y, Dani A, Vig M. alpha-SNAP regulates dynamic, on-site assembly and calcium selectivity of Orai1 channels. Mol Biol Cell. 2016;27(16):2542-53.

(3) Chorev DS, Baker LA, Wu D, Beilsten-Edmands V, Rouse SL, Zeev-Ben-Mordehai T, et al. Protein assemblies ejected directly from native membranes yield complexes for mass spectrometry. Science. 2018;362(6416):829-34.

(4) Dorwart MR, Wray R, Brautigam CA, Jiang Y, Blount P. S. aureus MscL is a pentamer in vivo but of variable stoichiometries in vitro: implications for detergent-solubilized membrane proteins. PLoS Biol. 2010;8(12):e1000555.

(5) Vig M, Peinelt C, Beck A, Koomoa DL, Rabah D, Koblan-Huberson M, et al. CRACM1 is a plasma membrane protein essential for store-operated Ca2+ entry. Science. 2006;312(5777):1220-3.

(6) Feske S, Gwack Y, Prakriya M, Srikanth S, Puppel SH, Tanasa B, et al. A mutation in Orai1 causes immune deficiency by abrogating CRAC channel function. Nature. 2006;441(7090):179-85.

(7) Murray DH, Tamm LK. Clustering of syntaxin-1A in model membranes is modulated by phosphatidylinositol 4,5-bisphosphate and cholesterol. Biochemistry. 2009;48(21):4617-25.

(8) van den Bogaart G, Meyenberg K, Risselada HJ, Amin H, Willig KI, Hubrich BE, et al. Membrane protein sequestering by ionic protein-lipid interactions. Nature. 2011;479(7374):552-5.

(9) Miranda P, Contreras JE, Plested AJ, Sigworth FJ, Holmgren M, Giraldez T. State-dependent FRET reports calcium- and voltage-dependent gating-ring motions in BK channels. Proc Natl Acad Sci U S A. 2013;110(13):5217-22.

(10) Chang CL, Chen YJ, Liou J. ER-plasma membrane junctions: Why and how do we study them? Biochim Biophys Acta Mol Cell Res. 2017;1864(9):1494-506.

(11) Lang T, Bruns D, Wenzel D, Riedel D, Holroyd P, Thiele C, et al. SNAREs are concentrated in cholesterol-dependent clusters that define docking and fusion sites for exocytosis. EMBO J. 2001;20(9):2202-13.

Version 2

Abstract

Mutations in Syntaxin11, a Q-SNARE, result in a fatal immune disorder known as familial hemophagocytic lymphohistiocytosis 4 (FHLH4) in human patients. A key diagnostic feature of FHLH4 is defective T and natural killer (NK) cell cytotoxicity. Here we show that Syntaxin11 directly binds and regulates Orai1, the pore forming subunit of calcium release activated calcium (CRAC) channels. CRAC channels enable store-operated calcium entry (SOCE) from the extracellular space and are crucial for granule exocytosis and nuclear factor of activated T cell (NFAT) dependent gene expression in activated lymphocytes. Syntaxin11 depletion strongly inhibited SOCE, CRAC currents, NFAT activation, interleukin-2 gene expression and degranulation in FHLH4 patient T lymphocytes and cell lines without affecting membrane trafficking. Remarkably, defects of cytolytic granule exocytosis as well as interleukin-2 expression could be reversed by ionomycin in patient T lymphocytes and a constitutively active, H134S, mutant of Orai1 rescued calcium entry in Syntaxin11 depleted cells. Further analyses showed that Syntaxin11 ‘primes’ Orai1 for optimal on-site multimeric assembly which was Stim independent but required for gating. Priming of ion channel pore subunits is, therefore, a primary function of specific SNAREs which may have preceded their role in membrane trafficking and vesicle fusion.

Introduction

Familial hemophagocytic lymphohistiocytosis 4 (FHLH4) is a life-threatening immune disease caused due to mutations of STX11 [1]. The underlying cause of FHLH is reduced cytolytic activity of T and NK cells [2], which renders patients susceptible to recurrent infections. Patients suffer from high fever, severe lymphopenia in early infancy and succumb to disease by adolescence, unless given bone marrow transplants. STX11 is highly expressed in almost all cells of the immune system and, being a Q-SNARE, is associated with exocytosis of vesicles, by default. Intriguingly, FHLH4 patients also manifest cytokine storm, which cannot be reconciled with a presumed general defect in membrane trafficking and vesicle fusion.

Most eukaryotic cells have a limited amount of calcium sequestered inside intracellular stores, the largest of which is the endoplasmic reticulum (ER). Signaling from cell surface receptors induces the release of stored calcium which, in turn, activates store-operated calcium entry (SOCE) to replenish stores, sustain signaling and drive other calcium dependent cellular processes [3]. Calcium release activated calcium (CRAC) channels are the major conductors of SOCE [4]. In lymphocytes and mast cells, CRAC currents have been shown to be crucial for cytotoxicity, degranulation as well as gene expression [5, 6]. Working in concert with a number of accessory proteins, Orai (CRACM) multimers conduct SOCE upon functional clustering with Stim proteins in ER-PM junctions [7–9]. Yet, the prevailing thought attributes all structural transitions that result in the activation of CRAC currents to Stim proteins [10].

Genome-wide RNAi screens, which initially identified Orai (CRACM), have also yielded a wealth of information regarding additional crucial players in SOCE [8, 11, 12]. For instance, we have previously shown that alpha-soluble N-ethylmaleimide sensitive factor (NSF) attachment protein (α-SNAP), a well-known synaptic family adaptor protein that forms a part of the 20S SNARE super-complex [13], is a crucial component of the CRAC channel supramolecular complex [8, 9]. α-SNAP independently binds Stim as well as Orai and is required for the correct assembly and calcium selectivity of CRAC channels [14–16]. α-SNAP depletion revealed that Stim:Orai coupling is necessary but not sufficient for SOCE, a process that requires distinct time-resolved molecular steps in vivo.

In addition to α-SNAP, SNAREs also came up as candidates in two genome-wide RNAi screens for regulators of SOCE but were not characterized [8, 12]. Furthermore, specific SNAREs have been previously found to form inhibitory associations with specific ion channels, but their exact role was not defined [17]. Therefore, we hypothesized a more fundamental function for the association of SNAREs with ion channel pore subunits and set out to test this in the context of SOCE. Using a targeted RNAi screen, biochemical and biophysical analyses, we have found a direct role of STX11 in facilitating the assembly of CRAC channel pore subunit, Orai1, into optimal multimers. Assembly of Orai1 was independent of, and preceded, trapping and gating by Stim1. Analyses of T cells derived from symptomatic STX11 deficient, FHLH4 patient showed that the previously reported [2], as well as newly identified, defects in degranulation, interleukin-2 gene expression predominantly stem from suppressed SOCE and not due to a general defect in vesicle trafficking or reduced levels of Orai1 in PM. Based on the findings presented here, we demonstrate a novel SNARE dependent priming step for CRAC channel pore in the molecular sequence of SOCE and suggest preventive pharmacological interventions for FHLH4 patients.

Results

RNAi dependent ablation of a Q-SNARE, Syntaxin 5 (STX5), has earlier shown inhibition of SOCE in a genome-wide screen performed in Drosophila S2 cells [12] and STX1A has been previously shown to bind a variety of ion channels [17]. Therefore, we first knocked down STX5 (Figure 1-figure supplement 1A) and STX1A (Figure 1-figure supplement 1B) using different sequences of shRNA targeting each gene in HEK293 cells and measured SOCE in response to Thapsigargin (TG). In both cases, we did not see a significant defect in SOCE. Since several proteins involved in synaptic function are not ubiquitously expressed and many show redundancy in function, we searched online databases to target SNAREs that are expressed in HEK293 and/or T cells (Figure 1-figure supplement 1C-H). We found that Syntaxin11 knockdown showed strong and consistent inhibition of SOCE by day 3 to 4 in HEK293 (Figure 1-figure supplement 2A-B) and Jurkat T cells (Figure 1A-B). Quantitative PCR (qPCR) analysis of total RNA extracted from scramble (scr) and STX11 shRNA treated cells showed ∼60-70% depletion of STX11 mRNA in both the cell lines (Figure 1-figure supplement 2C-D) and ectopic expression of STX11 in STX11 depleted Jurkat cells rescued SOCE (Figure 1C, D). The efficiency of STX11 depletion was assessed by qPCR for each experiment reported in this study. We restricted our analysis to day 3 to 4, post RNAi treatment, as longer incubations with RNAi or CRISPR knockout of STX11 affected cell viability.

STX11 is required for SOCE.

Measurement of SOCE in STX11 depleted Jurkat and FHLH4 patient T cells. (A-B) Representative Fura-2 calcium imaging assay (A) and quantification of independent repeats (B) measuring thapsigargin (TG) induced SOCE in Jurkat T cells treated with Scr (black), STX11#1 (red) or STX11#2 (blue) shRNAs. (n=30-40 cells/group, N=3). (C-D) Representative Fura-2 calcium imaging assay (C) and quantification of repeats (D) showing reconstitution of SOCE in STX11 depleted Jurkat T cells by ectopic expression of STX11. Scr shRNA with empty vector (EV) expression (black), STX11 shRNA with EV expression (red), STX11 shRNA with STX11 expression (green). (n=30-40 cells/group, N=3). (E) Sanger sequencing of FHLH4 patient DNA showing deletion of a single Adenine at the 752nd position and the resulting frameshift. (F) Schematic showing the predicted domain distribution of the wildtype versus FHLH4 mutant STX11 protein. (G) Western blot on WCLs of healthy donor (HD) and FHLH4 patient PBMCs showing the relative molecular weight and abundance of the wildtype and mutant STX11 bands. (H) Western blot on WCLs of HEK293 cells over-expressing wildtype or FHLH4 mutant STX11 and empty vector. (I) Fura-2 calcium imaging assay measuring anti-CD3 induced SOCE in HD (black) and FHLH4 (red) T cells. (n=30-40 cells/group). (J) Fura-2 calcium imaging assay measuring SOCE in HD expressing empty vector (EV) (black) and FHLH4 T cells expressing human STX11 (red), respectively. (n=30-40 cells/group). (K-L) Representative Fura-2 calcium imaging assay (K) and quantification of repeats (L) measuring SOCE in Jurkat T cells treated either with Scr shRNA expressing empty vector (EV) (black), STX11 shRNA, expressing EV (red), STX11 shRNA, expressing wildtype STX11 (green), STX11 shRNA, expressing FHLH4 mutant STX11 (blue).

Given that STX11 is highly expressed in most immune cells and is the only known SNARE that, when mutated, causes defects in T and NK cell cytotoxicity in humans and mice [2], we hypothesized that reduced SOCE contributes to the cytotoxicity defects of FHLH4 patient T cells. To test this hypothesis, we isolated PBMCs from an FHLH4 patient with a homozygous deletion frameshift mutation in the STX11 coding sequence, c.752delA:p.Lys251fs. The patient presented with all symptoms of FHLH4 disease by 4 years of age. Given the limited sample amount, we expanded PBMC numbers by periodically stimulating patient and healthy donor cells with PHA and IL-2, prior to analysis. Isolation and Sanger sequencing of DNA from patient PBMCs showed deletion of a single Adenine at position 752 in the STX11 gene which lead to frameshift following Lysine 251 (Figure 1E). The absence of any contaminating wildtype peaks showed that the mutation was present in all blood cells. Figure 1F shows schematic of the predicted mutant STX11 protein with an estimated molecular weight of ∼39.5 KDa compared to ∼33 KDa of wildtype STX11. The frameshift mutation resulted in disruption of the SNARE domain, ablation of the terminal cysteines and elongation of the resulting transcript. We subjected whole cell lysates prepared from the FHLH4 patient and healthy donor (HD) PBMCs to SDS-PAGE and Western blot and found that while the wildtype STX11 band was absent, a faint but distinct band running at ∼37 KDa was present in the FHLH4 patient sample lysate (Figure 1G). Therefore, the predicted and observed molecular weight of the FHLH4 mutant protein was higher but its levels appeared significantly reduced compared to STX11 from HD. This is likely due to protein instability and degradation because ectopic expression of the FHLH4 STX11 mutant in HEK293 cells showed at least two clear bands (Figure 1H), although reduced reactivity to anti-STX11 antibody and premature termination of translation are also possible. Interestingly, all prior FHLH4 STX11 patient mutations that have been biochemically characterized, till date, also showed severe depletion of STX11 protein levels in patient T and NK cells [18–21]. Therefore, the frame shift mutation that has been characterized in our study is representative of most other STX11 human mutations, characterized earlier, in terms of the mechanisms underlying the observed phenotypes. In accordance with our findings in the STX11 depleted cell lines (Figure 1A-B), SOCE was found to be significantly defective in FHLH4 T cells when compared to HD (Figure 1I). Further, expression of wildtype (Figure 1J), but not FHLH4 mutant STX11 (Figure 1K-L), reversed defective SOCE in STX11 deficient patient (Figure 1J) and Jurkat T cells (Figure 1K-L), respectively. These data suggest that, in addition to being unstable, FHLH4 mutant STX11 was functionally compromised and no additional unknown mutations resulted in reduced SOCE.

To determine whether the inhibition of SOCE in Figure 1 resulted from changes in the activity of CRAC channels or their PM expression levels, we quantified the total as well as plasma membrane levels of Orai1 in different cell lines (Figure 2). We generated a stable HEK293 cell line by expressing a previously characterized Orai1 fusion construct containing bungarotoxin binding site (BBS) in the extracellular loop and YFP in the C-terminus (Orai-BBS-YFP) (Figure 2-figure supplement 1A) [9]. Resting and store-depleted Orai-BBS-YFP expressing HEK293 cells were labelled with Alexa-647 conjugated alpha-bungarotoxin (BTX-A647), washed, fixed and analyzed using flow-cytometry (Figure 2A-B) as well as microscopy (Figure 2C) to estimate total expression levels and PM localization of Orai1. We found no difference in the total (Figure 2A) or surface Orai1 (Figure 2B-C) expression in any group. Verification of knockdown efficiency in these HEK293 cells is shown in (Figure 2-figure supplement 1(B-C)). Similar observations were made in Jurkat T cells transiently transduced with pMSCV-Orai1-BBS-YFP (Figure 2D) (Figure 2-figure supplement 1D) as well as U2OS cells (Figure 2-figure supplement 2A-E) stably expressing the same plasmid construct as described above for HEK293. Neither total Orai1 expression levels, nor it’s PM localization was found to be altered upon store-depletion or STX11 ablation in these cell types.

STX11 depletion does not affect PM Orai1 levels and global membrane trafficking or secretion.

(A-C) Measurement of Bungarotoxin binding site (BBS) tagged Orai1-YFP in the PM of scramble and STX11 shRNA treated, resting and store-depleted, HEK293 cells. Measurement of Orai1-YFP (A) and Alexa Fluor 647-conjugated Bungarotoxin binding (B) in population of cells using flow-cytometry. Quantification of Orai1-YFP and BTX-A647 in individual cells using confocal imaging (C). (D) Confocal images and quantification of Orai1-YFP and BTX-A647 in scramble and STX11 shRNA treated, resting and store-depleted, Jurkat T cells. (n=30-40, N=3) (E-G) Scr or STX11 shRNA treated HEK293 and Jurkat T cells were loaded with Alexa-647-Transferrin (30 minute pulse), washed, chased for 0, 5, 10 and 20 min and imaged to quantify total Transferrin levels inside each cell. Representative images from HEK293 cells (E) and quantification from 50-60 HEK293 (F) and Jurkat T cells (G) is shown. N=3 (H) Quantification of non-receptor mediated endocytosis in STX11-depleted HEK293 cells. Representative images of FITC-conjugated Dextran uptake in Scr and STX11-shRNA treated HEK293 cells (left panel) and quantification of the mean intensity per cell (right panel). (I) Scr or STX11 shRNA treated HEK293 cells were transfected with secretory Alkaline phosphatase (AP) expressing plasmid and levels of AP secreted into the culture supernatants were measured using p-Nitrophenyl phosphate substrate at 48h post-transfection. (J-K) Immunolabelling of HEK293 cells to assess ER (J) and Golgi (K) health. Scr and STX11 shRNA treated HEK293 cells were stained with anti-KDEL (J) and anti-GM130 antibodies (K) respectively and imaged. (L) Measurement of ER calcium content in Scr and STX11 shRNA treated HEK293 cells using transiently transfected ER-localized CEPIA. Cells were incubated in Ringer’s buffer containing 0 mM calcium followed by 1 mM EGTA and stimulated with 1 μM ionomycin. (N=3).

To determine whether STX11 deficient cells harbored a general defect in vesicle trafficking and fusion, we used the Transferrin receptor (TfR) endocytosis and recycling assay, which internalizes iron bound holo-transferrin via endocytosis of cell surface TfR, followed by exocytosis and extracellular release of iron free apo-Transferrin [22]. RNAi treated HEK293 (Figure 2E-F) and Jurkat T cells (Figure 2G) were incubated with Alexa-647 conjugated Transferrin (A647-TfR) for a 30 minute endocytic pulse. Excess A647-Transferrin was washed off and the emptying of internalized A647-Transferrin via recycling of TfRs over time was quantified at various time points (chase). Levels of A647-Transferrin after endocytosis (30-minute pulse-0-minute chase) as well as it’s exocytosis during 5, 10, 20-minute chase times were found to be identical between STX11 or Scr shRNA treated cells. Non-receptor mediated fluid-phase endocytosis of FITC-conjugated Dextran was also found to be identical between Scr and STX11 shRNA treated HEK293 cells (Figure 2H). In addition, we transfected HEK293 cells with a plasmid containing alkaline phosphatase (AP) enzyme fused to Ig-κ chain secretion signal peptide at N-terminal, which allowed us to measure AP secreted into the culture supernatant via the biosynthetic-exocytosis pathway. Scramble and STX11 depleted cells showed comparable levels of AP secretion (Figure 2I). Overall, these experiments did not indicate any generalized defects in the membrane recycling or secretory pathways upon depletion of STX11 in multiple cell types.

Using a dominant negative mutant of N-ethylmaleimide sensitive factor (NSF), which blocked TfR recycling but not SOCE, we have previously shown that CRAC channels are not dependent on vesicle trafficking for their activation [9]. Yet, Q-SNAREs are often found in complex with other SNAREs such as SNAP23/25 in target membranes and some previous studies have implicated the complex of SNAP23/25 and STX1A in the regulation of specific ion channels [23, 24]. Each of the three SNAPs, SNAP23/SNAP25/SNAP29 are capable of contributing two SNARE domains to the trans- SNARE complex. RNAi mediated depletion of SNAP23/SNAP25/SNAP29 did not show a significant reduction in SOCE of HEK 293 cells (Figure 2, figure supplement 3A-C). To determine whether SNAP23/SNAP25/SNAP29 might still form a part of the STX11:Orai complex to regulate SOCE, we co-expressed myc-tagged Orai1 with untagged SNAPs and performed respective co-IPs. We did not find any interaction between either of the three SNAPs and Orai1 (Figure 2-figure supplement 3D). Similar studies were performed by co-expressing YFP-Stim1 and the three myc-tagged SNAPs. Again, no interaction was found between any of the SNAPs and Stim1 (Figure 2-figure supplement 3E). Collectively, these data established that Orai1 is constitutively expressed in the PM, and not vesicles, of diverse cell types and STX11 depletion does not lead to a general defect in vesicle fusion or PM targeting of Orai1.

Labelling of STX11 depleted cells with ER, Golgi organelle specific markers also showed no abnormalities (Figure 2J-K) suggesting that the overall health and function of ER and Golgi were not adversely affected by STX11 depletion. To assess ER calcium content, we ectopically expressed ER-targeted CEPIA in control and STX11 depleted cells and measured the mean intensity. Both groups showed comparable ER calcium content (Figure 2L).

Previous studies have shown that SOCE in T cells is mediated by CRAC channels formed by Orai proteins. To test whether CRAC current (ICRAC) was affected by knockdown of STX11 in T cells, we performed whole cell patch clamp recordings on scramble and STX11 shRNA treated Jurkat T cells (Figure 3A-B). Leak-subtracted ICRAC was recorded in 20 mM Ca2+ and a Na+-based divalent cation free (DVF) solution. These recordings showed that both the Ca2+ and DVF CRAC currents were significantly decreased in the STX11 shRNA treated cells with the normalized currents shown in Figure 3C. The electrophysiological properties of the residual current in STX11 shRNA treated cells was indistinguishable from ICRAC in scramble shRNA treated cells in terms of blockade of the Ca2+ current by La3+ and depotentiation of the DVF current over tens of seconds. Moreover, the reversal potential of the Ca2+ and DVF currents were similar between scramble and STX11 shRNA treated Jurkat T cells, suggesting that STX11 depletion does not affect the calcium selectivity of CRAC channels (Figure 3D). Together, these results demonstrate that STX11 works by specifically modulating ICRAC in T cells.

STX11 regulates ICRAC.

(A-D) Measurement of ICRAC in STX11 depleted Jurkat T-cells in the whole-cell recording configuration in 20 mM extracellular Ca2+ Ringer’s solution. ICRAC was induced by passive depletion of intracellular Ca2+ stores by dialyzing 8 mM BAPTA into the cell via the patch-pipette. (A) Representative current at −100 mV in Jurkat T cells treated with Scr shRNA construct. The current is blocked by extracellular La3+ (10 µM) and replacing the 20 mM Ca2+ Ringer’s solution with a divalent free solution (DVF) evokes a large Na+ current which depotentiates over tens of seconds. The current-voltage (I-V) relationship of the Ca2+ and DVF currents are shown on the right. (B) ICRAC from a Jurkat T cell treated with STX11 shRNA. Both Ca2+ and Na+ current amplitudes are reduced relative to control cells. The I-V relationships (right plots) show no change in ion selectivity. (C-D) Summary of the current amplitudes of Ca2+ and Na+ currents and current reversal potentials in Scr and STX11 knockdown cells.

CRAC channel mediated sustained calcium influx is crucial for the activation of a key calcium dependent transcription factor known as nuclear factor of activated T cells (NFAT) [4, 25]. We stimulated control and STX11 depleted Jurkat T cells with Thapsigargin (TG)+PMA, prepared nuclear, cytosolic extracts and subjected them to western blot using anti-NFAT1 antibody (Figure 4A). Nuclear translocation of NFAT was severely compromised in STX11 depleted T cells. In addition, immunofluorescence imaging and quantification of total versus nuclear NFAT from anti-CD3 stimulated Jurkat cells showed a defect in nuclear translocation of NFAT1 in STX11 depleted cells (Figure 4B-C). To determine whether NFAT induced gene expression was affected [4, 25], we quantified IL-2 mRNA by qPCR and secreted IL-2 levels from culture supernatants of anti-CD3 stimulated Jurkat T cells. STX11 depleted cells showed a significant decrease in IL-2 gene expression as well as secretion (Figure 4D, 4E). We stimulated HD and FHLH4 patient T cells either through T cell receptor using anti-CD3 or with phorbol myristate acetate (PMA) + Ionomycin and assessed IL-2 gene expression. Importantly, induction of IL-2 expression was significantly defective in FHLH4 patient T cells but could be largely restored with Ionomycin (Figure 4F). We next performed granule release assay on the in vitro cultured HD and FHLH4 patient CD8 T cells in response to anti-CD3 mediated stimulation and observed a significant defect in the FHLH4 patient CD8 T cell degranulation (Figure 4G, 4H). To determine whether the reduced granule release results from a direct defect in vesicle fusion or SOCE, we stimulated cells with Ionomycin along with (PMA). Remarkably, CD8 T cell degranulation was fully restored in FHLH4 T cells suggesting that defective SOCE primarily causes cytolytic defects in FHLH4 T cells. Of note, the patient exhibited all symptoms of FHLH4 disease and we observed strong suppression of SOCE (Figure 1I), IL-2 gene expression and secretion (Figure 4D-F). However, in vitro culture of FHLH4 patient CD8 T cells with IL-2 has been previously shown to partially or fully restore cytolytic defects by other groups [18, 26]. Therefore, considering previous reports, a relatively lower (50%) defect in CD8 T cells degranulation was not surprising [18, 26]. Taken together, the data show that defects in FHLH4 T cell degranulation as well as interleukin-2 gene expression originate from suppressed SOCE and not vesicle trafficking.

STX11 regulates NFAT activation, IL-2 gene expression and degranulation in FHLH4 patient T cells.

(A-C) Estimation of nuclear translocation of NFAT. (A) Western blot showing nuclear translocation of NFAT in Jurkat T cells treated with scr or STX11 shRNA and stimulated with PMA+TG. N=3. (B) Representative confocal images of control and STX11 depleted Jurkat T cells stimulated with anti-CD3, immunolabelled with anti-NFAT1 antibody and counter-stained with DAPI. (N=3) (C) Box and whisker plot showing quantification of nuclear NFAT from cells populating 10 randomly chosen fields per group in (B). (D) Quantification of IL-2 mRNA in anti-CD3 stimulated Jurkat T cells treated with scr or STX11 shRNA using qPCR. The bars show mean ± SE of relative IL-2 mRNA. N=3. (E) Quantification of IL-2 EIA on supernatants of scr (black) or STX11 (red) shRNA treated Jurkat T cells stimulated with anti-CD3. (F) qPCR analysis of IL-2 mRNA expression in various stimulated HD (black) and FHLH4 T cells (red). (G-H) Granule release assay performed on HD (black) and FHLH4 CD8 T (red) cells (G) and its quantification (H).

Unlike most Q-SNAREs, STX11 lacks a transmembrane domain but harbors a stretch of cysteines close to the C-terminus and several basic residues, spread throughout the sequence (Figure 5-figure supplement 1A) [27–29]. The cellular localization of STX11 is not established because, unlike primary T cells, native STX11 is undetectable in Jurkat T and HEK293 cell lines using available commercial antibodies. Tagging STX11 on either N- or C-terminus result in partial degradation (Figure 5-figure supplement 1B) and complete mis-localization of the protein (Figure 5-figure supplement 1C). Therefore, we inserted an HA tag inside the N-terminal unstructured region of human STX11. The internal HA tagged STX11 as well as un-tagged STX11 localized in plasma membrane of most HEK293 cells (Figure 5A) and co-localized with Orai1-YFP in resting (Figure 5B) as well as store-depleted cells (Figure 5C). Figure 5D shows quantification of Orai1 and STX11 co-localization using Pearson’s correlation coefficient. Notably, despite loss of all terminal cysteines and a short terminal part of the SNARE domain, the FHLH4 mutant STX11 did not lose the ability to localize to PM (Figure 5E), therefore, terminal cysteines do not appear to target STX11 to the PM.

STX11 directly binds resting Orai1 in plasma membrane.

Intracellular localization of STX11 and its association with Orai1. (A) Representative confocal images of HEK293 cells transfected with either HA-tagged STX11 or untagged STX11 and stained using anti-HA antibody or anti-STX11 antibody respectively. N=3 (B-D) Localization of HA-tagged STX11 with respect to Orai1-YFP in the basal as well as nuclear plane of resting (B) and store-depleted (C) HEK293 cells. (D) Pearson’s correlation coefficient showing Orai1 and STX11 colocalization from (B) and (C). N=3. (E) Representative confocal images of FHLH4 mutant STX11 expressing HEK293 cells. (F-G) Co-IP to assess STX11 binding to Orai1. Whole cell lysates of resting and store-depleted HEK293 cells co-expressing either Flag-Orai1 and STX11 (F) or Orai1-Myc and STX11 (G) were subjected to IP and Western blot using anti-Myc, anti-Flag or anti-STX11 antibodies, as indicated. (N=3) (H) Schematic showing key domains of STX11 and Orai1 used for in vitro pull-down assays. (I) Pull-down assay showing in vitro binding of His-tagged STX11 to Orai1. (Top) Ponceau S staining showing the input of MBP alone or MBP-tagged Orai1 fragments. (Bottom) Western blot using anti-STX11 antibody. Input:1/5th of the protein. N=4. (J) Pull-down assay showing in vitro binding of His-tagged Habc domain of STX11 to Orai1 fragments. (Top) Ponceau S staining showing the input of MBP alone or MBP-tagged Orai1 fragments. Input:1/5th of the protein. (Bottom) Western blot using anti-His antibody. N=3. (K) Binding free energy distribution of STX11-Habc and Orai1 C-terminus interactions. (L) Sphere and cartoon representation of the structure of STX11 Habc (green) and Orai1 C-terminus (cyan) complex after MD simulation. The N-termini are highlighted in blue, and C-termini in red. (M) Representative confocal images showing localization of STX11 with respect to Orai1:Stim1 puncta in store-depleted cells. HEK293 cells expressing Orai1-YFP, CFP-Stim1 were transfected with HA-tagged STX11, store-depleted and stained using anti-HA antibody. (N=3) (N) Quantification of the localization of STX11 with respect to Stim1:Orai1 puncta in store-depleted cells.

To test whether STX11 and Orai1 formed a complex, we co-expressed STX11 with either Flag- or Myc-tagged Orai1 in HEK293, prepared whole cell lysates (WCLs) and subjected them to co-immunoprecipitation (co-IP) followed by western blot using either anti-Flag, anti-Myc or anti-STX11 antibodies. Both Orai1 and STX11 could co-IP each other (Figure 5F-G). To further assess whether STX11 directly bound Orai1, we expressed and purified from E. coli, MBP-tagged N- and C-terminal cytosolic tails of Orai1 (Figure 5H). In vitro pull-down assay performed by incubating MBP-tagged Orai1 cytoplasmic tails with full length soluble His-tagged STX11 showed that the C-terminus of Orai1 directly bound STX11 (Figure 5I). SNAREs typically utilize their SNARE domain for associating with other SNAREs [30]. To identify the domain of STX11 involved in the regulation of SOCE, we expressed His-tagged Habc and SNARE domains of STX11 (Figure 5H) in E. coli, purified and assessed their binding to the MBP-tagged Orai1 tails using pull-down assay. We found that the Habc domain of STX11 bound to the C-terminus of Orai1 (Figure 5J) but the SNARE domain of STX11 did not show any binding to Orai1 and only a faint binding to SNAP23 (Figure 5-figure supplement 1D). We next used AlphaFold3 (AF3) to obtain deeper insights into the domain interactions of STX11 and Orai1. Of all the possible combinations, the complex of STX11 Habc and Orai1 C-terminus resulted in a significantly high prediction score (Table 1). We, therefore, executed AF3 predictions with different initial seeds to generate multiple models of STX11 Habc with Orai1 C-terminus. We calculated the contact frequency of interface residues in different models and found that multiple residues showed frequency greater than 80% (Figure 5-figure supplement 2A). For further analysis, we considered the predicted model with the best score (Table 2). We next performed all-atom MD simulation in aqueous environment (Movies 1-3) to assess the conformation, interaction stability of the STX11 Habc and Orai1 C-terminus complex, which was stable throughout the simulation time as suggested by RMSD (Figure 5-figure supplement 2B), and binding free energy (ΔG) values (Figure 5-figure supplement 3) (Figure 5K). The individual trajectories were concatenated and clustered to obtain a centroid structure for this complex. There were 46 clusters, and the largest cluster, with 52 members, showed an elaborate protein-protein interface where STX11 Habc was found to be oriented in an anti-parallel orientation to the Orai1 C-terminus (Figure 5L).

Because STX11 binds the C-terminus of Orai1 and the region largely overlaps with the Stim1 binding domain, we next sought to determine whether STX11 clusters with Orai:Stim puncta. We over-expressed HA-tagged STX11 in a Orai1-YFP and CFP-Stim1 expressing stable HEK293 cell line, store-depleted with TG, and stained using anti-HA antibody. We did not observe co-clustering of STX11 in Orai:Stim puncta even 8 minutes post store-depletion (Figure 5M-N). Accordingly, STX11 did not co-IP with Stim1 in store-depleted cells (Figure 5-figure supplement 4A). These data suggest that STX11 comes off Orai1 C-terminus prior to Stim1 dependent clustering of Orai1. Since co-IP studies suggested sustained interaction between STX11 with Orai1 (Figure 5F-G), we quantified the fraction of Orai1 that resides inside Stim:Orai puncta w.r.t total Orai1 in PM and found that post TG addition, nearly 60% of Orai1 remains outside of Stim1 puncta even in unmanipulated cells (Figure 5-figure supplement 4B). We next used AF3 to generate the complex of Stim-Orai activating region (SOAR) domain of Stim1 and Orai1 (Figure 5-figure supplement 5A). Given that SOAR is a dimer under resting state, we used two copies each of SOAR and Orai1 [31] [32]. The prediction scores for their complex are shown in Table 3. The best scoring model among multiple runs of AF3 was considered for MD simulations on the complex of Orai1 C-terminus (256-301) and SOAR. The ΔG values for the complex are shown in (Figure 5-figure supplement 5B-C). Three independent MD simulations were run for the complex (Movies 4-6) which showed an initial conformational change which stabilized thereafter (Figure 5-figure supplement 6A). All three simulation trajectories were concatenated and clustered to obtain a centroid structure for this complex (Figure 5-figure supplement 6B). There were 43 clusters, and the largest cluster had 77 structures. Collectively, these data suggest that STX11 directly binds Orai1 with an affinity that is physiological but relatively poorer when compared to the binding of SOAR dimers to Orai1. Therefore, presence of high density of SOAR dimers in ER-PM junctions of store-depleted cells could potentially outcompete STX11:Orai1 interactions by progressively allowing formation of SOAR:Orai1 oligomers. It is also possible that another, yet to be characterized, molecular step and/or protein actively catalyzes the removal of STX11 from Orai1 prior to its entrapment in ER-PM puncta. Future studies will be able to systematically assess these possibilities.

The major interactions predicted between STX11 and Orai1 from all-atom MD simulations are highlighted in (Figure 5-figure supplement 3B-C). To assess the significance of these interactions, we performed site-directed mutagenesis of two key residues of STX11 predicted to be involved in binding to Orai1 (Figure 6A). Ectopic expression of untagged N147A_E150A STX11 showed PM localization (Figure 6A) as well as stable expression (Figure 6B) but failed to rescue SOCE in STX11 depleted Jurkat T cells (Figure 6C-D). We next expressed His-tagged wildtype and N147A_E150A mutant STX11 in E. coli., purified (Figure 6E) and assessed their ability to bind MBP-tagged Orai1 C-terminal tail in an in vitro pull-down assay (Figure 6F-G). Compared to wildtype STX11, N147A_E150A mutant STX11 showed reduced binding to Orai1 C-terminus. Further, we introduced four mutations, R289A_E272A_E275A_E278A, in the Orai1 C-terminus (Figure 6H), tagged with YFP and expressed in HEK293 cells. The Orai1 mutant localized to PM (Figure 6H) but failed to rescue SOCE in Orai1 deficient cells (Figure 6I-J). We expressed and purified MBP-tagged R289A_E272A_E275A_E278A mutant Orai1 C-terminus from E. coli and assessed its ability to bind purified wildtype as well as N147A_E150A mutant STX11 in a pull-down assay (Figure 6K-L). Binding of Orai1 mutant to both was significantly compromised. Because Stim1 binds an overlapping region of the Orai1 C-terminus, we next tested whether mutation of Orai1 residues affected Stim1 binding. We co-expressed YFP-tagged soluble CRAC activation domain (CAD) of Stim1 (YFP-CAD) (Figure 6M) with either CFP-tagged wildtype or mutant Orai1 in HEK293 cell line. YFP-CAD localizes to the cytosol in HEK293 cells but in cells overexpressing Orai1, a significant fraction of it localizes to plasma membrane due to its association with Orai1 [10]. We quantified relative YFP-CAD levels in the plasma membrane of both groups but found no difference in the PM recruitment of CAD (Figure 6N) suggesting that the inability of R289A_E272A_E275A_E278A mutant Orai1 to support SOCE does not result from defective binding to CAD/SOAR. However, CAD was unable to stimulate constitutive calcium influx from R289A_E272A_E275A_E278A mutant Orai1 in the above experiment (Figure 6O-P). Collectively, these data demonstrate a direct role of STX11 in binding and regulation of Orai1.

Residues involved in Orai1:STX11 complex formation are crucial for SOCE.

Analysis of STX11 and Orai1 mutants. (A) Schematic showing STX11 mutations (Top). STX11 N147A_E150A mutant expressing HEK293 co-labelled with Alexa Fluor-488 conjugated wheat germ agglutinin (WGA) or co-transfected with Orai1-YFP and immunolabelled with anti-STX11 to assess co-expression in PM of resting and store-depleted cells (Bottom). Scale bar 10μm. (B) Western blot to compare total expression of mutant and wildtype STX11 using anti-STX11 antibody. (C-D) Representative Fura-2 calcium imaging assay (C) and quantification of repeats (D) measuring rescue of SOCE in Jurkat cells treated either with scr shRNA (black) or STX11 shRNA expressing empty vector (red), wildtype STX11 (green) or N147A_E150A mutant STX11 (blue). (E) Coomassie blue stained SDS-PAGE showing expression and purification of His-tagged wildtype or N147A_E150A mutant STX11 from E. coli. (F-G) Pull-down assay measuring their binding to MBP-tagged Orai1 C-terminus. (Top) Ponceau S staining showing the input of MBP-tagged Orai1 fragments. (Bottom) Western blot using anti-STX11 antibody. N=4 (G) Quantification of STX11 band intensities in (F). (H) Schematic showing Orai1 mutations (Top). Confocal images of HEK293 cells expressing YFP-tagged wildtype or R289A_E272A_E275A_E278A mutant Orai1 (Bottom). (I-J) Representative Fura-2 calcium imaging assay (I) and quantification of repeats (J) measuring rescue of SOCE in HEK293 cells treated either with Scr (black) or Orai1 shRNA expressing empty vector (red), wildtype Orai1 (green) or R289A_E272A_E275A_E278A mutant Orai1 (blue). (K-L) Pull-down assay measuring the binding of MBP-tagged wildtype or R289A_E272A_E275A_E278A mutant Orai1 C-terminus to wildtype or N147A_E150A mutant STX11. (Top) Ponceau S staining showing the input of MBP-tagged Orai1 fragments. (Bottom) Western blot using anti-STX11 antibody. N=4. (L) Quantification of STX11 band intensities in (K). (M-N) Representative confocal images (M) and quantification (N) of PM localized YFP-CAD in CFP-tagged wildtype or R289A_E272A_E275A_E278A mutant Orai1 expressing HEK293 cells. (O-P) Representative Fura-2 calcium imaging assay (O) and quantification of repeats (P) of constitutive calcium flux from cells imaged in panels M-N.

Upon store depletion, Stim1 localizes to ER-PM junctional regions where it traps and co-clusters with Orai1. To further understand the mechanism, using confocal microscope we first imaged HEK293 cells expressing CFP-Orai1, Stim1-YFP under resting conditions and quantified the levels of Orai1 and Stim1 in ER-PM junctions (puncta) of store-depleted control and STX11 depleted cells. The distribution of Orai1 and Stim1 was normal in resting cells (Figure 7A). However, in store-depleted cells, clustering of CFP-Orai1 was visibly defective in STX11 RNAi treated cells (Figure 7B), even though the intensities (Figure 7C) and area of Stim1-YFP puncta were normal (Figure 7D).

Non-functional clustering of Orai1 in STX11 depleted cells.

(A-L) Quantification of Orai1 and Stim1 intensities in the ER-PM puncta of store-depleted cells. (A-B) Representative images of resting (A) and store-depleted (B) shRNA treated HEK293 cells expressing N-terminal CFP tagged Orai1 (CFP-Orai1) and C-terminal YFP tagged Stim1 (Stim1-YFP) from confocal microscopy. (C-F) Mean Stim1-YFP intensity (C) and area (D) of Stim1 clusters. (E) Fraction of CFP-Orai1 intensities inside Stim1-YFP clusters of shRNA treated cells, post store-depletion. (F) Pearson’s correlation coefficient of Stim1 and Orai1 co-localization inside Stim1 puncta. (G-J) Quantification of Orai1 and Stim1 intensities in the ER-PM puncta of C-terminal YFP-tagged Orai1 (Orai1-YFP) and N-terminal CFP-tagged Stim1 (CFP-Stim1) expressed in HEK293 cells and imaged using TIRF microscopy. (G) Representative TIRF images of resting and store-depleted shRNA treated HEK293 cells. (H-K) Mean CFP-Stim1 intensity (H) and area (I) of Stim1 clusters. Fraction of Orai1-YFP intensities (J) and area (K) of Orai1 clusters in control and STX11 depleted cells. (L) Pearson’s correlation coefficient of Stim1 and Orai1 co-localization inside Stim1 puncta (M-N) Representative Fura-2 calcium imaging assay (M) and quantification of repeats (N) measuring TG induced SOCE in Orai1-YFP and CFP-Stim1 expressing HEK293 cells treated with scr (black) or STX11 (red) shRNA. (O-Q) Representative TIRF images of GFP-Mapper in store-depleted scramble and STX11 shRNA treated HEK293 cells (O) and quantification of intensity (P) and area of GFP-MAPPER puncta (Q). n=15.

Quantification of Orai1 intensities confirmed a partial defect in entrapment with relatively lower fraction of Orai1 present inside Stim1 puncta in STX11 depleted cells (Figure 7E-F). To determine whether reduced clustering of Orai1 resulted from structural defects within ER-PM junctions, we reversed the direction of fluorescent tags on Orai1 as well as Stim1 and repeated the experiment from Figure 7A-F in Figure 7G-L, using TIRF microscopy. Interestingly, the defect in Stim1 mediated entrapment of Orai1 could be rescued (Figure 7G-L) but SOCE remained significantly inhibited despite co-overexpression of Orai and Stim (Figure 7M-N) [33] [34] [35], suggesting ‘nonfunctional’ co-clustering of Orai1 and Stim1. We further assessed binding of full length YFP-Stim1 to mutant R289A_E272A_E275A_E278A Orai1-CFP and found that Stim1 was able to co-cluster with mutant Orai1 (Figure 7-figure supplement 1A-D).

Stim1 can itself serve as a marker of ER-PM junctions in store-depleted cells and quantification of area and mean intensity of Stim1-YFP or CFP-Stim1 puncta showed no defect in STX11 depleted cells imaged using confocal (Figure 7C,D) or TIRF microscopy (Figure 7H,I). To still assess whether STX11 depletion altered the structure and/ or proximity of junctional ER to PM, we expressed an alternate marker of ER-PM junctions, GFP-MAPPER, in STX11 depleted cells [36]. Quantification of the area and intensity of GFP-MAPPER puncta imaged using TIRF microscopy also did not show any significant difference in store-depleted, STX11 shRNA treated HEK293 cells (Figure 7O-Q). Collectively, these data show that STX11 binding to resting Orai1 has implications for CRAC channel function beyond mere co-entrapment by Stim1.

Defects in ‘functional’ entrapment of Orai1 suggest absence of a necessary molecular transition within Orai1 which precedes its interaction and gating by Stim1. To test whether STX11 induced structural changes within CRAC channel pore subunit, Orai1, we co-expressed Orai1-CFP along with Orai1-YFP in scramble control or STX11 depleted HEK293 cells and measured Orai:Orai FRET. We found significantly higher Orai1:Orai1 FRET in STX11 depleted resting cells in the absence of ectopically expressed Stim1 (Figure 8A). We have previously shown that store-depletion induces on-site assembly of Orai1 dimers into oligomers [14]. Notably, store-depletion induced a similar level of increase in Orai1:Orai1 FRET in scramble control cells. To confirm these findings, we performed acceptor photobleaching analysis on Orai1-CFP (donor), Orai1-YFP (acceptor) co-expressing HEK293 resting cells where we had stably knocked down the expression of endogenous Orai1 and Stim1. Although the trends were largely similar, the differences in Orai1:Orai1 FRET were much more significant in these cells (Figure 8B-C). These data demonstrate a previously unsuspected, novel molecular transition in Orai1 necessary for CRAC channel activation, which we define as ‘priming’. Further, we show that priming precedes the interaction of Orai1 with Stim1 in ER-PM junctions. To determine whether ‘priming’ simply induced a structural change in Orai1 protomers or facilitated correct assembly of Orai1 multimers, we expressed Flag-Orai1 in control or STX11 depleted resting cells, mechanically lysed to isolate membranes, crosslinked Orai1 oligomers using BS3 and resolved using SDS-PAGE (Figure 8D) (Figure 8-figure supplement 1). We found a greater percentage of higher order oligomers in STX11 depleted cells suggesting stoichiometric changes (Figure 8E).

STX11 primes Orai1 for optimal on-site assembly.

Oligomerization of Orai1. (A) N-FRET analysis of scramble and STX11 shRNA treated HEK293 cells co-expressing Orai1-CFP and Orai1-YFP before and after store-depletion. (B-C) Representative images (B) and quantification (C) of FRET analysis using acceptor photobleaching of Scr and STX11 shRNA treated Orai1, Stim1 double knockdown resting HEK293 cells co-expressing Orai1-CFP (donor) and Orai1-YFP (acceptor). (D-E) Representative western blot (D) and quantification of band intensities from repeats (E) showing distribution of BS3 crosslinked Flag-Orai1 oligomeric bands in control and STX11 depleted HEK293 cells. N=3. (F) Representative confocal image and quantification (right) of YFP-CAD localization in Scr and STX11 shRNA treated HEK293 cells expressing Orai1-CFP. (G) Fura-2 calcium imaging assay (G) and quantification of repeats (H) to measure constitutive calcium influx in Orai1-CFP and YFP-CAD expressing, control (black) or STX11 depleted (red) HEK293 cells. n=80-90, N=3 (I-J) Fura-2 calcium imaging assay (I) and quantification of repeats (J) to measure constitutive calcium influx in Orai1-H134S mutant expressing control (black) and STX11 depleted (red) HEK293 cells. n=60-70, N=3 (K-L) Fura-2 calcium imaging assay (K) and quantification of repeats (L) to measure constitutive calcium influx in Orai1-ANSGA mutant expressing control (black) and STX11 depleted (red) HEK293 cells. n=50-60, N=3

Increasing the density of Stim1 has been earlier proposed to bypass the requirements for accessory factors in SOCE. To determine whether increasing the levels of Stim1 could compensate for STX11 depletion, we expressed Orai1 tethered to two Stim-Orai activating region (SOAR) domains of Stim1 and GFP (Orai1-S-S-GFP) (Figure 8-figure supplement 2A-B), which has been previously shown to constitutively activate Orai1 in HEK293 cells [31]. In STX11-depleted Orai1-S-S-GFP expressing cells, the magnitude of calcium entry was significantly smaller (Figure 8-figure supplement 2C-D). To determine whether reduced calcium entry resulted from a defect in the ability of SOAR to bind Orai1, we expressed YFP-CAD, in control or STX11 depleted Orai1-CFP expressing HEK293 cell line (Figure 8F). Quantification of PM localized YFP-CAD as a fraction of total CAD showed no defect in STX11 depleted cells. Yet, as seen with Orai1-S-S-GFP, calcium entry was significantly reduced (Figure 8G-H). We next expressed a constitutively active H134S mutant of Orai1 (Figure 8-figure supplement 3A) as it harbors an open pore and the C-terminal cytosolic tails of Orai1 are unlatched, pointing towards the cytosol [37]. Remarkably, expression of H134S Orai1 completely restored constitutive calcium influx in STX11 depleted cells (Figure 8I-J) and showed similar PM expression when compared to wildtype Orai1 in all the groups (Figure 8-figure supplement 3B). To further establish whether binding with STX11 induces straightening of Orai1 tails or molecular transitions within transmembrane helices, we expressed in STX11 depleted cells the constitutively active ANSGA (261-265) mutant of Orai1 (Figure 8-figure supplement 3C), which harbors 4 consecutive mutations in the nexus region close to the Orai1 C-terminus resulting in the straightening of the C-terminal tail along with constitutive activation of CRAC currents [38]. The constitutive calcium entry in STX11 deficient Orai1 ANSGA mutant expressing cells was significantly suppressed (Figure 8K-L) even though the PM expression of the ANSGA Orai1 mutant was unchanged in STX11 depleted cells (Figure 8-figure supplement 3D). Taken together, these data show that STX11 induces a molecular switch in the pore-forming subunit, Orai1, which precedes and primes it for optimal on-site assembly into multimers and subsequent gating by Stim1. Such SNARE dependent priming of Orai1 is crucial and cannot be compensated by merely increasing the density of Stim proteins bound to Orai or by introducing mutations that alter the orientation of Orai1 tails. Figure 9A and Figure 9B show the proposed models depicting unprimed and primed Orai1 and its role in T cell function.

Models depicting STX11 mediated priming of Orai1 and its role in T cell signaling and function.

(A) STX11 unbound (unprimed) and bound (primed) Orai1 in resting and store-depleted cells. For simplicity, STX11 interaction with only two subunits of Orai1 has been shown. (B) Regulation of T cell SOCE, NFAT activation and effector function by STX11.

FHLH is a heterogeneous immune disorder where diverse genes such as Prf1, STX11, UNC13D, Munc18-2, MAGT1, XIAP, LYST, AP3B1, RAB27A, ITK, amongst several others, are broadly grouped together based on disease symptoms, although the age of onset and severity varies across the spectrum [39, 40]. Congenital immunodeficiency involving defects in the CTL and NK cell cytotoxicity appear to be the common factors underlying the development of FHLH, although the individual pathways involved are highly diverse. Using shRNA, we screened 6 additional genes, Perforin1, RAB27A, UNC13D and Munc18-2, Synaptotagmin11 and 13, that have been either associated with vesicle fusion or routinely grouped together with STX11 in diagnosing FHLH. SOCE was not affected by depletion of most genes (Figure 1-figure supplement 3). Future studies will evaluate the role of others.

Discussion

CRAC channels conduct a small, highly specific calcium current in response to the depletion of intracellular calcium stores. According to the prevailing view, Orai multimers reside in the PM and depend on the ER resident store sensor, Stim, for all structural transitions leading up to their activation [10]. Unlike other ion channels, the process of activation of CRAC currents is exceptionally slow. For instance, following break-in or store-depletion, on average, it takes ∼5 minutes for Stims to cluster in the ER-PM junctions and for measurable CRAC currents to flow from Orai1. During this time, Stims undergo intramolecular transitions and slowly trap and gate freely diffusing Orai. It is, therefore, believed that no molecular transitions take place within the pore forming subunit, Orai, until its entrapment by Stim in the ER-PM junctions.

We have shown that in STX11 deficient cells, the basal Orai:Orai FRET is significantly higher even in the absence of Stim1 or store-depletion. These data establish that Orai1 undergoes a, hitherto unrecognized, SNARE dependent molecular transition which precedes trapping and gating by Stim1. We have earlier shown that store-depletion induces dynamic on-site assembly of Orai1 dimers into oligomers [14]. In STX11 depleted resting cells, the ∼25-50% increase over basal Orai:Orai FRET largely results from a change in the oligomerization of Orai1 because the proportion of BS3 crosslinked Orai1 oligomers was also found to be shifted towards higher molecular weight. It is very likely that a change in the overall conformation of Orai1 protomers facilitated this change in store-dependent Orai assembly and contributed to the overall change in Orai:Orai FRET. This conclusion is also supported by the observation that STX11 depletion induced change in Orai:Orai FRET which not only persisted but increased in Orai1, Stim1 double deficient cells.

Furthermore, STX11 bound a region of Orai1 C-terminus that has been previously shown to bind Stim1. Therefore, STX11 induced molecular transitions in Orai are independent of and precede its association and clustering by Stims. Because tagging of STX11 with fluorescent tags on either end results in mis-localization and degradation, our study could not demonstrate the real-time removal of STX11 from the Orai1-C-terminus to allow for Stim1 binding in live cells. It is also unclear whether binding or removal of STX11 induced the proposed molecular transition defined as ‘priming’ of Orai1. Further, it remains to be determined whether STX11 removal is actively induced by an additional chaperone or passive displacement due to formation of relatively higher avidity oligomeric associations between Orai1:Stim1 in ER-PM junctions. Future studies will be able to systematically evaluate each of these possibilities.

We have shown that the C-terminal cytosolic tail of Orai1 directly couples with STX11 in resting cells. Although binding appeared reasonably strong in co-IPs, STX11:Orai1 interaction appeared somewhat weak in pull-downs. This is likely due, in part, to the instability of purified His-STX11 in solution which necessitated shorter co-incubation times during pull-down assays to prevent precipitation of STX11. Instability could arise from the absence of a stabilizing co-chaperone, such as one of the known binding partners of SNAREs [41] [42] or membranes to which STX11 has been proposed to attach in live cells. Together, these factors appear crucial for the stabilization of purified STX11 in vitro.

The individual relevance of Orai1 N-versus C-terminus in the trapping versus gating of Orai1 remains unclear [43]. Though crucial for trapping and gating, Orai1 tails were truncated in the early structures of Drosophila Orai [37]. A previous NMR structure of truncated Orai1 C-terminal tails suggested that the tails of two adjacent Orai1 subunits, bend, pair with each other in an antiparallel fashion and sit closely apposed to PM [44]. However, in recent structures of constitutively active H134 mutant Orai, the C-terminal tails were found to orient away from the membrane [37]. Although C-term tails in ANSGA Orai1 mutant are also proposed to be unlatched, STX11 depletion inhibited ANSGA Orai1 induced constitutive calcium influx [38]. In the absence of direct structural evidence that establishes conformational similarity between constitutively active H134 versus ANSGA mutant Orai1, Orai1+CAD and Orai1-S-S, it is likely that Orai1 can assume several different open conformations/ assemblies [14, 37, 45–47]. Since most mutants showed sensitivity to STX11 depletion, it reinforces our conclusions that STX11 induced conformational shifts precede gating and involve Orai1 TMs.

Using Fura-2 imaging assays in STX11 depleted cells, we have conclusively established that STX11 is necessary as well as limiting not only in live and unperturbed HEK, U2OS, Jurkat and primary T cells but also HEK293 cells co-overexpressing Orai1 and Stim1. Previous whole cell patch clamp analysis of HEK293 cells co-overexpressing Orai1 and Stim1 showed up to 50-100-fold amplification of CRAC currents whereas Fura-2 based imaging of live cells routinely shows only 2-fold amplification [33] [34] [35]. The patch clamp studies performed on Stim, Orai over-expressing cells are often extrapolated to conclude that all other accessory factors that have since been identified are unnecessary for native CRAC currents. A major problem with this interpretation is that even though amplification of CRAC currents was shown, none of the previous patch clamp studies established whether higher currents resulted from a greater number of active channels or unchecked conductance per channel by performing single channel recordings. Direct single-channel recordings of CRAC are challenging. As a result, estimates of open probability and channel number often rely on fluctuation (noise) analysis rather than direct measurements of single-channel events [48]. The fact that recording of ICRAC in the whole cell configuration involves use of 20mM extracellular calcium and dialysis of cells with high concentrations of BAPTA, a fast calcium chelator, the calcium dependent inactivation of CRAC channels, a key mechanism of regulation of ICRAC magnitude in live cells is disrupted. Furthermore, hyperpolarizing potentials and voltage clamping used to record CRAC currents prevents the progressive depolarization of cells due to influx of cations which would otherwise dampen the driving force for CRAC currents in intact cells. Given these limitations, Fura-2 assays offer a more realistic and physiological reflection of SOCE amplitudes in live cells as they allow for these regulatory mechanisms, which are necessary for the regulation of native currents, to be still operational. While patch clamp is a direct method of verification of CRAC currents, exercising logic in the extrapolation of amplitudes of ICRAC using patch clamp is imperative.

Most human mutations in the STX11 coding region result in the complete loss or severe depletion of STX11 protein levels [28]. The FHLH4 disease, therefore, results from reduced STX11 expression in most reported patients. Targeted mutagenesis of the STX11:Orai1 interface lining residues based on the AlphaFold 3 model, however, didn’t destabilize or mis-localize either N147A_E150A STX11 or R289A_E272A_E275A_E278A Orai1 and yet both mutants showed reduced binding to their respective wildtype counterparts. Importantly, the R289A_E272A_E275A_E278A mutant of Orai1 retained the ability to bind and cluster with Stim1.

Most importantly, we have shown that deficiency of a Q-SNARE, STX11, causes a, vesicle trafficking independent, severe defect in SOCE, IL-2 gene expression and degranulation in FHLH4 T cells. Mechanistically, STX11 directly binds and induces a molecular switch in Orai1 necessary for its assembly into ‘functional’ multimers which we call a ‘primed’ state. We suggest that ion channel priming is a novel, direct and primary role of specific SNAREs, independent of vesicle trafficking. In accordance with this hypothesis, early SNARE-like proteins have been found in bacteria, which lack an endomembrane system but express a variety of ion channels [49–51]. SNAREs are now thought to have evolved from a common archaeal precursor found in the genomes of Asgard and members of Legionella [52].

Analyses of FHLH4 patient T cells suggest that SOCE defects significantly contribute to the pathogenesis of FHLH4 disease. We, therefore, propose that pharmacological induction or promotion of primed intermediate state of Orai1 could potentially provide an alternative to immunosuppressants and bone marrow transplants for FHLH4 as well as SCID patients, with minimal risk of autoimmunity.

Materials and Methods

Plasmid constructs and transfection

Stim1 CAD was sub-cloned in eYFPC1 vector after amplification from YFP-Stim1 plasmid to generate YFP-CAD. pLKO.1, psPAX2 and pMD2.G were purchased from Addgene. pLKO.1 cloned shRNA sequences targeting the genes of interest, were either purchased from Horizon Discovery, UK or designed and cloned in-house using the Broad Institute portal (https://portals.broadinstitute.org/gpp/public/gene/search). Full-length human STX11 was cloned from human cDNA prepared from HEK293 cells and subcloned into pcDNA3.1(+), pcDNA/4TO/Myc-HisA (Invitrogen, Grand Island, NY), pMSCV-IRES-mcherry (Addgene), pEF1alpha-IRES (Clontech) and pET28b vector [16] with an N-terminal 6XHis-tag. The fragments of STX11 were cloned in pET28b vector with an N-terminal 6xHis tag for expression in E. coli. HA tag was inserted between H30 and G31 in STX11 cloned in pMSCV-IRES-mCherry construct using PCR. H134S and ANSGA mutants of Orai1 were generated using site directed mutagenesis and PCR, respectively. The fragments of Orai1 (1-87, 1-47, 48-103, 256-301 and 272-292) were amplified from full length constructs and cloned into pMAL-c5X vector (New England Biolabs), in-frame with MBP protein coding sequence, as described previously [16]. SNAP23, SNAP25 and SNAP29 cDNA cloned in pCMV-Sport6.1 vectors were purchased from Dharmacon and subcloned into pcDNA4/TO/Myc-HisA. SNAP23 was subcloned in pMAL-c5X vector in-frame with MBP. All plasmid DNA transfections in human cell lines and primary cells were done using Lipofectamine 2000 (Invitrogen)/ Lipofectamine 3000 (Invitrogen) or Amaxa nucleofection kit (Lonza, Basel, Switzerland) respectively, as per manufacturer’s protocol.

Cell Lines

HEK293-FT cell line was cultured in high glucose DMEM with 10% FBS and 10mM HEPES, 1X Penicillin Streptomycin, 1X GlutaMax and 1X non-essential amino acid and transfected with the appropriate plasmids to generate viral supernatants. Lentiviral shRNA transduction experiments were performed in HEK293, U2OS or Jurkat (ATCC, Manassas, VA) cell lines cultured in low glucose DMEM (Hyclone, Logan, UT) or RPMI (Hyclone) respectively, with 5-10% fetal bovine serum (Hyclone), 1X Penicillin Streptomycin, and GlutaMax (Gibco, Grand Island, NY). Stable cell lines generated using HEK293 and U2OS parent lines have been described previously [9]. To generate HEK293 stable cell line expressing Orai1-BBS-YFP construct, the cells were nucleofected (SE Cell Line 4D-Nucleofector X kit_Amaxa _V4XC) and YFP positive cells were sorted. All cell lines were tested for mycoplasma contamination twice every year and found to be negative.

Lentiviral transductions

For the generation of lentiviral supernatants, shRNAs cloned in pLKO.1 were co-transfected with psPAX2 packaging and pMD2.G envelope plasmids into HEK293-FT cells using the calcium phosphate method of transfection. 48- and 72-hours post-transfection, viral supernatants were collected, pooled and stored at −80°C till further use. For transduction of HEK293 and Jurkat cells, 0.1 million cells were plated in 6- or 24-well plates either the day before (HEK293) or the same day (Jurkat). Viral supernatants were added to the cells along with 8μg/ml polybrene and cells were spun at 2500 RPM, 30°C for 90 mins. 24 hours or 48 hours post-spinfection, Puromycin was added at a final concentration of 1μg/ml to HEK293 and Jurkat cells, respectively. Cells were analyzed 3-5 days post transduction.

Immunocytochemistry and confocal imaging

For STX11 localization, resting or store-depleted (1μM Thapsigargin) WT HEK293 or HEK293 cells stably expressing Orai1-YFP were either plated in 6 well plates and spinfected with viral supernatants generated from pMSCV-STX11(HA)-IRES-mCherry, pMSCV-IRES-mCherry (empty vector control) or plated in carbon coated, glow discharged 35mm glass bottom dishes (IBDI) and transfected with pEF1alpha-STX11-IRES-mCherry. 24 hr post-spinfection, cells were plated in 35mm glass bottom dishes. For immunolabelling, cells were washed with Ringer’s buffer, fixed with 4% PFA and blocked with 3% BSA containing 0.1% NP40 for 1.5 hrs. Post-fixation and permeabilization, cells were incubated either with anti-HA (CST) or rabbit anti-Stx11 (SySy/ Proteintech) primary antibodies O/N at 4°C, washed and incubated with either Alpaca VHH anti-rabbit AF488, donkey anti-rabbit A647 or goat anti-rabbit A647 secondary antibodies for 1 hour at room temperature (RT). Cells were counter stained with DAPI for 10 min at RT and imaged using Olympus FV3000 laser scanning confocal microscope. Images were acquired sequentially using the following parameters: DAPI (DM405/488 dichroic, 405nm excitation, 430-470nm emission); anti-HA Alexa488 (DM405/488, 488nm excitation, 500-590nm emission); STX11 (DM405/488/561/640, 640nm excitation, 650-750nm emission); Orai1-YFP (DM405/488/561/640, 488nm excitation, 521-591nm emission); CFP-Stim1 (DM405-445/514/594, 405nm excitation, 448-510nm emission). For STX11-N147A-E150A mutant, HEK293 cells were spinfected with viral supernatants and imaged as described above where WGA (Wheat germ agglutinin) conjugated with Alexa 488 was used to mark plasma membrane. For immunolabelling of ER and Golgi, HEK293 cells were treated with Scr or STX11 shRNA, washed, fixed, blocked and incubated with anti-KDEL and anti-GM130 primary antibodies overnight at 4°C. Following incubation, the cells were washed with 1X PBS containing 0.1% NP40, incubated with Donkey anti-mouse A647 secondary antibody for 30 mins at RT, washed and counter-stained with DAPI for 10 mins. Images were acquired using Olympus FV3000 laser scanning microscope as mentioned above.

Quantification of Orai1 and Stim1 intensities inside/outside ER-PM puncta

HEK293 cells stably or transiently expressing wildtype Orai1-YFP or E272A-E275A-E278A-E289A Orai1-YFP with CFP-Stim1 or CFP-Orai1 with Stim1-YFP were plated in 6 well plates and transduced with scramble or STX11 shRNA. 48hr post transduction, cells were trypsinized and plated in carbon coated 35mm dishes. Images were acquired using a TIRF microscope setup described before 14. Briefly, resting cells were imaged first and positions were marked. Store-depleted images were acquired after incubation with 1 μM Thapsigargin (TG) and 10mM EGTA for ∼6 mins. Stim1 and Orai1 clusters were defined using Otsu local thresholding and Pearson’s correlation coefficient was measured using colocalization finder in ImageJ software. Stim1 and Orai1 intensity (AU), area in µm2 and Pearson’s correlation coefficient values values obtained from the analysis were plotted using Origin software.

Orai:Orai FRET

Scramble and STX11 shRNA transduced HEK293 cells were plated on freshly carbon coated, glow discharged 35mm glass bottom dishes (IBDI) and transfected with 0.5 μg each of Orai1-CFP and Orai1-YPF plasmid DNA using Lipofectamine 3000. 16 hours post transfection, co-transfected cells were imaged using PlanApoN 60XO/1.42 NA objective with Olympus FV3000 laser scanning microscope. Orai1-CFP, Orai1-YFP and FRET images were acquired with the following excitation laser and emission bandwidth parameters using DM405-445/514 dichroic mirror: CFP Ex-405nm Em-460-630nm; YFP Ex-514nm Em-530-630nm; FRET Ex-405nm, Em-530-630nm. ROIs representing the plasma membrane rim were drawn around each cell, followed by local thresholding using Bernsen’s algorithm in Fiji software to generate an image mask which was applied to CFP, YFP, FRET images to extract mean intensity values. The background subtracted mean intensities of FRET (IFRET), CFP (IDonor) and YFP(Iacceptor) images were measured for each cell. Cells singly expressing Orai1-CFP or Orai1-YFP only were used as CFP donor or YFP acceptor channel bleed-through (BT) controls. The BT was calculated using the following formula:

BTDonor=IFRETIDonor when the Orai1-CFP donor alone is expressed,

BTacceptor=IFRETIacceptor when the Orai1-YFP acceptor alone is expressed.

FRET and NFRET percentage was calculated by subtracting the contributions of CFP and YFP bleed through from the FRET signal using the following: FRET=IFRET−(BTDonor∗IDonor)−(BTAcceptor∗IAcceptor)NFRET%=(FRET(IAcceptor∗IDonor))∗100

Acceptor Photobleaching

Scramble and STX11 shRNA transduced, Orai1, Stim1 double knockdown HEK293 cells where were plated on Poly-D-Lysine coated glass bottom dishes (Cellvis) and transfected with 0.5 ug DNA each of Orai1-CFP and Orai1-YPF using Lipofectamine 2000. 16 hours post transfection, co-transfected cells were imaged using PlanApoN 60XO/1.42 NA objective with Olympus FV3000 laser scanning microscope. Images were acquired with the following excitation laser and emission bandwidth parameters using DM405-445/514 dichroic mirror: CFP Ex-405nm Em-460-508nm; YFP Ex-514nm Em-530-630nm, with 4% offset. The pre bleach images of Orai1-CFP (the donor) and Orai1-YFP (the acceptor) were captured, sequentially. This was followed by bleaching of the acceptor, Orai1-YFP, using 100% 514 laser power for 2 mins and then acquiring the donor and acceptor images post bleach. ROIs were drawn around the cell periphery using the Orai1-CFP image. Local thresholding was performed using Bernsen’s algorithm to generate binary masks, which were then applied to the CFP images to measure mean intensities. The background subtracted mean intensities of CFP images were measured for each cell and the FRET efficiencies were calculated as follows: E%=(1−IbeforeIafter)∗100 Ibefore and Iafter denotes the average fluorescence intensity of the donor in the plasma membrane before and after acceptor photobleaching.

Alpha-bungarotoxin binding assay

U2OS or HEK293 cell lines stably expressing Orai1-BBS-YFP construct were either transduced with scramble or STX11 shRNA and harvested 3 days post-transduction. Jurkat cells treated with Scr and STX11 shRNA were transiently transduced with pMSCV-Orai1-BBS-YFP 48 hrs prior to harvest. For surface labeling, cells were incubated on ice with AF647 conjugated alpha-bungarotoxin (BTX-A647) (1μg/ml) dissolved in 1X HBSS with 2% FBS 30 mins, washed and fixed with 4% PFA (∼20 min at RT). For flow-cytometry, U2OS and HEK293 cells were analyzed on Cytoflex FACS analyzer (Beckman Coulter) and data were analyzed using Flow Jo software. For microscopy, the shRNA treated HEK293 cells were plated in 35mm glass-bottom dishes, one day prior to harvest, whereas labeled Jurkat cells were plated on poly-D-lysine coated coverslips, mounted onto glass slides and sealed. Imaging was performed using Olympus FV3000 confocal microscope.

Line profiles of Orai1-YFP and BTX-A647 were drawn on raw confocal images of HEK293 cells. Quantifications of Orai1-YFP and BTX-A647 were calculated using the following formula: %Orai1−YFPorBTX−A647in thePM=(PMIntensityTotal Intensity)∗100

Alkaline phosphatase (AP) assay

HEK293 cells were transduced with Scr and STX11 shRNA and 48 hrs post-transduction the cells were transfected with pAP tag plasmid (Genhunter) in low glucose DMEM without phenol red. Supernatants from control and transfected cells were collected at 48hour post transfection and triplicates of undiluted supernatant or those diluted at 1:1, 1:2,1:4 using DMEM without phenol red were added to 96-well plates and mixed in 1:1 ratio with 1 mg/ml pNPP (Disodium-4-Nitrophenyl Phosphate Hexahydrate) substrate in AP Assay Buffer (100 mM Tris-Cl pH-9.0, 100 mM NaCl and 5 mM MgCl2) to obtain final AP:pNPP ratios of 1:1, 1:2, 1:4 and 1:8. Supernatants from untransfected cells were used as blank controls. After 25 mins of incubation at 37°C in the dark, the assay was terminated by adding 0.4M NaOH and absorbance at 405 nm was measured using Softmax Pro version 5.4 in FlexStation 3.

Measurement of GFP-Mapper cluster size and intensity

HEK293 cells treated with Scr and STX11 shRNA were transfected with GFP-Mapper plasmid, 48 hrs post-transduction. The cells were store-depleted with 1 μM Thapsigargin (TG) and 10mM EGTA for ∼6 mins, fixed with 4% PFA, washed and imaged using TIRF microscopy. To analyze the images, local thresholding was performed in ImageJ using the Bernsen algorithm, and the resulting mean intensity and area (μm2) of the GFP-Mapper clusters were plotted using Origin software.

Validation of STX11 knockdown using qPCR

Scramble and STX11 shRNA treated HEK293 and Jurkat T cells were counted and equal number of cells were used for lysis. The lysates were homogenized using QIA shredder columns (Qiagen) and RNA isolation was done using RNeasy Plus Mini kit (Qiagen). Total RNA isolated from both the groups was used for cDNA synthesis using oligo dT primers and Superscript IV as per manufacturer’s guidelines. The concentration of synthesized cDNA was estimated using Qubit ssDNA assay kit and STX11 TaqMan probes were used to perform the qPCR using LightCycler 96 (Roche). Beta-actin and RPL30 were used as housekeeping control.

Single cell Ca2+ imaging

Cells were plated in carbon-coated glass bottom dishes (one day prior or 30 mins before the assay for HEK293, U2OS and Jurkat T cells respectively) and loaded with 1μM Fura-2 AM dye in HBSS (CaCl2 1.8mM, KCl 5.36mM, MgSO4 0.81mM, NaCl 136.89 mM, Na2HPO4 0.335mM, D-Glucose 5.55mM) for 30 mins, at 37°C in the dark. After incubation, cells were washed and incubated in Ringer’s buffer (135mM NaCl, 5mM KCl, 1.8mM CaCl2, 1mM MgCl2, 5.6mM Glucose, 10mM HEPES) for an additional 10 mins, washed and imaged in Ringer’s buffer or Calcium-free Ringer’s buffer (135mM NaCl, 5mM KCl, 1mM MgCl2, 5.6mM Glucose, 10mM HEPES at pH7.5), as indicated. Olympus IX-71 inverted microscope equipped with a Lamda-LS illuminator (Sutter Instrument), Fura-2 (340/380) filter set (Chroma, Bellows Falls, VT), a 10X 0.3NA objective lens (Olympus, UPLFLN), and a Photometrics Coolsnap HQ2 CCD camera was used to capture images at a frequency of ∼1 image pair every 2 or 4 seconds interval. Data were acquired, analyzed and plotted using MetaFluor (Molecular Devices), Microsoft Excel, and Origin softwares. Approximately 30-40 cells were imaged per group in each experiment unless otherwise stated. In-vitro calibration of Fura-2 was performed using calibration kit (Molecular Probes_F6774) according to the manufacturer’s protocol. Briefly, calcium calibration buffers spanning a range of free calcium concentrations (0, 0.017, 0.038, 0.065, 0.1, 0.15, 0.225, 0.351, 0.602, 1.35, and 39 µM) and containing 15 µM microsphere beads were loaded onto cleaned glass slides, sealed with a coverslip and imaged. A standard linear calibration curve was constructed by plotting the 340/380 ratios (y-axis) against the known free calcium concentrations (x-axis) and used to quantify the absolute intracellular free calcium concentrations using simple linear regression equation y=mx+c.

Quantification of SOCE

Quantification of mean SOCE ± SE from three independent experiments was performed as follows. The mean of 5 peak amplitude values after the addition of 2mM calcium from one of the scramble shRNA groups, amongst the three experiments, was normalized to the baseline values (5 values just before the addition of 2mM calcium) and was set at 100%. The relative percent responses of all other Scr and test shRNA groups were normalized with respect to this group.

Constitutive calcium influx assay

For constitutive calcium influx assay, shRNA treated HEK293 cells were either transfected or nucleofected with Orai1-S-S-EGFP, Orai1(H134S)-YFP or Orai1(ANSGA)-YFP and ∼12-14 hour post transfection (∼4-5 hours post nucleofection), cells were loaded with Fura-2 AM, washed and imaged in Ringer’s buffer with 0 mM calcium to acquire baselines and 2 mM calcium thereafter. To measure CAD induced constitutive calcium influx, Orai1-CFP expressing stable HEK293 cells were nucleofected with YFP-CAD and analyzed ∼4 hour later as described above. To measure CAD induced calcium influx in E272A-E275A-E278A-R289A mutant Orai1-YFP expressing cells, Orai1-shRNA treated HEK293 cells were transfected with wildtype or mutant Orai1-YFP constructs along with YFP-CAD. To identify cells expressing Orai1 mutants and fusion proteins in the plasma membrane, cells were imaged using a 20X 0.7NA water objective lens (Olympus, UApoN340, Japan). Images were acquired at a frequency of ∼1 image pair every 10 seconds interval to avoid photobleaching and analyzed as described above. Quantification of constitutive calcium influx using Orai1 mutants was done by plotting the average of five consecutive data points immediately following addition of 2mM calcium across three independent repeats. Fura 2 calibration was done separately for these experiments as described above.

Transferrin recycling assay

HEK293 and Jurkat T cells treated with Scr or STX11 shRNA were plated on 35-mm glass-bottom dishes and washed three times with serum-free DMEM to remove residual FBS. The cells were then incubated with Alexa Fluor 647-conjugated human holo transferrin (20ug/ml) in serum-free DMEM, for 30 min at 37°C (pulse), placed on ice for 5 min and subjected to three rapid washes with acidified ice-cold serum-free DMEM (pH 3.5-5.5) to remove transferrin bound to cell surface receptors. Cells were then chased for indicated time points at 37°C, washed with HBSS, fixed with 4% PFA, washed again with PBS and imaged using an IX83 widefield microscope equipped with a 60X/1.25 NA objective lens (Olympus UPFLN). Fluorescence images were collected on a Hamamatsu Orca-Fusion camera using the same settings across all conditions. Brightfield images were used to define cell boundaries, and per-cell mean fluorescence intensities were quantified using ImageJ software.

Dextran-FITC uptake assay

HEK293 cells treated with Scr or STX11 shRNA were plated on 35-mm glass-bottom dishes, washed with serum-free DMEM and incubated with FITC-conjugated dextran (1mg/ml) in serum-free DMEM, for 30 min at 37°C. The cells were then washed with HBSS, fixed with 4% PFA and imaged. To preserve FITC fluorescence post-fixation, the cells were treated with 20mM ammonium chloride prior to imaging. Images were acquired as described above for Transferrin recycling assay.

Measurement of NFAT nuclear localization by Western blot

Jurkat T cells were transduced with scramble or STX11 shRNA to knock-down STX11 as described above. On day 4 post-transduction, cells were collected, spun down and resuspended in plain RPMI media and rested for 1hr at 37°C. Following this, cells were counted and divided into two equal groups. One group was resuspended in RPMI (unstimulated) and the other in RPMI media containing 1 μM TG + 10 ng/ml PMA (Phorbol Myristate Acetate) (stimulated) and incubated at 37°C for 30 mins. Following incubation, cells were pelleted and the nuclear and cytosolic protein fractions were separated using NE-PER kit according to the manufacturer’s guidelines and subjected to SDS-PAGE and western blot using the mouse anti-NFATc2 primary antibody (Santa Cruz) followed by Donkey anti-mouse secondary antibody.

Estimation of nuclear translocation of NFAT by immunolabeling

Control and STX11 depleted Jurkat T cells were rested in plain RPMI for ∼1 hour prior to the assay. ∼100,000 cells per group were plated on freshly carbon-coated coverslips for 40 minutes and stimulated for 1 hour at 37°C with 5 μg/ml anti-CD3 antibody, diluted in plain RPMI. Following this, cells were washed and fixed using 4% PFA diluted in 1X PBS for 20 minutes at RT, washed twice and incubated with 30 mM Glycine for 10 minutes at RT. For permeabilization and blocking, cells were incubated with 0.1% Saponin, 3% BSA diluted in 1X PBS for 1 hour at RT, washed and incubated with anti-NFAT primary antibody (anti-NFAT1, CST) at 4°C, overnight. Following primary antibody application, cells were washed with 1X PBS and incubated with anti-Rabbit AF647 secondary antibody for 1 hour, washed and stained with DAPI for 5 minutes followed by additional washes. Images were acquired in the FV3000 confocal microscope using a 100X objective lens. DAPI was used to identify the nuclear area. Nuclear versus whole cell (total) NFAT mean intensity ratio was plotted across different groups.

Whole cell lysates (WCLs), Western blot and Co-immunoprecipitation (Co-IP)

HEK293 cells transfected with the desired plasmids were lysed using buffer containing 50 mM Tris-Cl (pH 8.0), 150 mM NaCl, 1% NP-40, 1 mM PMSF, and protease inhibitor cocktail. The whole cell lysates were centrifuged at 21000 g for 15 minutes and supernatants were subjected to SDS-PAGE, and western blot. For immunoprecipitations, lysates were divided into two equal parts. To the first tube, the appropriate anti-mouse or anti-rabbit primary antibody was added and to the second, same amount of the respective IgG control antibody was added. The antibody-lysate mixtures were incubated overnight at 4°C. Following this, Protein A/G Mag Sepharose beads were added and incubated with the antibody-lysate mixtures for 4 hours, washed with the lysis buffer containing 0.1% NP-40, boiled with 1X Laemmli buffer + 120 mM DTT and subjected to SDS-PAGE and western blot analysis. Typically, 1/10th of the whole cell lysate (WCL) was loaded in the input lane of the co-IP blots. The full-length blots pertaining to each figure have been deposited in Zenodo https://doi.org/10.5281/zenodo.21663510

E. coli expression and in-vitro binding assays

Full-length His6-tagged Stx11 and truncation mutants were cloned in pET28b, expressed in Lemo21 (DE3) E. coli cells and induced with 1 mM IPTG (Isopropyl β-d-1-thiogalactopyranoside) for 14-18 hours at 18°C. The cell pellets were lysed in buffer containing 50 mM Tris-Cl (pH 8.0), 150 mM NaCl, 10% Glycerol, 1 mM PMSF, 1% Sarkosyl, 0.1 mg/ml Lysozyme, protease inhibitor cocktail and sonicated on ice. DNase I was added after sonication and the lysates were further incubated for ∼60 minutes before centrifugation at 21000 g for 40 mins. The supernatants were subjected to SDS-PAGE to confirm expression by Coomassie staining and subsequently used for pull-down assays. MBP-tagged Orai1 constructs were expressed in Lemo21 (DE3) E. coli cells, induced with 0.3mM IPTG (Isopropyl β-d-1-thiogalactopyranoside) for 14-18 hours at 18°C. The cell pellets were lysed in buffer containing 50 mM Tris-Cl (pH 8.0), 150 mM NaCl, 5% Glycerol, 1 mM PMSF, 0.1 mg/mL Lysozyme and protease inhibitor cocktail (1:200) and sonicated on ice. DNase I was added after sonication and the lysates were further incubated for ∼60 minutes before centrifugation at 21000 g for 40 mins. The supernatants were collected and subjected to SDS-PAGE to confirm expression by Coomassie staining and used for pull-down assays. Following lysis, the His6-tagged Stx11 proteins were purified using Talon Beads and MBP-tagged Orai1 proteins using Dextrin Sepharose/Amylose resin according to the manufacturer’s guidelines. For in-vitro binding assays, lysates prepared from cells expressing MBP or MBP-tagged Orai1 fragments were incubated with 25 μL Dextrin Sepharose Beads and incubated for 2 hours at 4°C. After incubation, beads were washed thrice and ∼50-125 ng of purified His-tagged full length Stx11, His-tagged SNARE or His-tagged Habc domains were diluted in buffer containing 50mM Tris-Cl (pH8), 150 mM NaCl, 5% Glycerol and 0.1% NP-40 and added to the beads. Following 1 hour of incubation at 4°C, the beads were washed thrice, re-suspended and boiled in the binding buffer containing 1X Laemmli and 120 mM DTT. The eluate was subjected to SDS-PAGE and Western blot. MBP and MBP-tagged proteins were detected with Ponceau staining, Full length STX11 using Rabbit anti-STX11 primary antibody (Thermo) followed by Donkey anti-rabbit HRP and His-tagged STX11 fragments were detected or mouse anti-6XHis primary antibody (Invitrogen) followed by Donkey anti-mouse HRP secondary antibody. Densitometric quantification of the relative band intensities was performed using ImageJ software and plotted in Origin software. To control for variation in experimental repeats, the band intensity of the control complex for wild-type (MBP-Orai1-WT:His-STX11-WT) for individual repeats was set to 100 percent and relative intensities of bands from other test groups were calculated with respect to this.

BS3 crosslinking of Flag-Orai oligomers

Scramble and STX11 shRNA transduced HEK 293 cells were transfected with Flag-Orai1, harvested, washed, pelleted and subjected to one cycle of snap freeze-thaw. The pellet was resuspended in ice cold lysis buffer (1X HBSS, 1mM PMSF and protease inhibitor cocktail), sonicated for 30 secs (Branson 2510) and mechanically homogenized using dounce homogenizer. The homogenized supernatant was spun at 1500 x g for 10 mins at 4° C, to pellet the cell debris. The supernatant was ultracentrifuged at 105,000 x g for 1 hour at 4° C (Optima MAX-XP ultracentrifuge, Beckman Coulter). The membrane fraction pellet was resuspended in lysis buffer and re-homogenized using dounce homogenizer. The protein concentration was estimated using BCA assay and BS3 cross-linker was added to the resuspended membranes, at a final concentration of 1mM, and incubated for 30 mins at room temperature. BS3 was quenched using 50mM Tris-HCl pH 7.5 for 15 mins at room temperature, followed by addition of 1X Laemmli and 100 mM DTT and incubation at 37° C for 10 mins. The samples were resolved using 4-15% gradient SDS-PAGE and subjected to Western blot using anti-Flag antibody.

Quantification of Flag-Orai oligomers using Western blot

Equal size ROIs were drawn over corresponding protein bands of same size, using ImageJ. The total intensity of each band was measured by calculating the area under the intensity curve and its relative proportion was calculated by dividing the raw intensity of each band by the sum of intensities of all the bands in the respective lanes.

qPCR to estimate gene expression in Jurkat T cells

Scramble (scr) and STX11 shRNA treated Jurkat T cells were rested for ∼1 hour prior to the assay, stimulated with soluble anti-CD3 (5-10 μg/ml) for 3 hours and washed with HBSS. Total RNA was isolated using RNeasy Mini kit as per manufacturer’s instructions. cDNA was synthesized using random hexamers and Invitrogen Superscript IV kit and quantified using Qubit ssDNA assay. Taqman probes for IL-2 and beta-actin housekeeping control were used for performing the qPCR in triplicates. Data analysis was done by calculating the double delta Ct values.

EIA to measure IL-2 levels in supernatants of Jurkat T cells

Scramble (scr) and STX11 shRNA treated Jurkat T cells were stimulated with plate-coated anti-CD3 (10 μg/ml) and the culture supernatants were collected at 12 and 24 hour post-stimulation. EIA plates were coated with purified mouse anti-human IL-2 capture antibody (1 μg/ml) overnight at 4°C, washed with 0.05% PBST and blocked with 2% BSA for 2 hours. After blocking, the plates were incubated with the culture supernatants for 2 hours at RT, washed and incubated with the detection antibody (Biotin mouse anti-human IL-2, 1 μg/ml) for 2 hours. After washing, Streptavidin HRP (0.1 μg/ml) followed by TMB reagent were added. 250 mM HCl was added to stop the reaction and absorbance was measured at 450 nm.

Isolation and culture of human PBMCs

Whole blood freshly collected in Heparin or Citrate Phosphate Dextrose Adenine (CPDA) solution was subjected to density gradient centrifugation using Ficoll-Paque PLUS media. The buffy coat was separated, washed twice with HBSS and cultured with IL-2 (50 ng/ml). Cells were stimulated once per week with PHA (2 μg/ml) for 72 hours and rested in IL-2 for the remaining 72-96 hours. Unless specified, all the assays were performed following 24-48 hours of rest, post stimulation.

Isolation of genomic DNA and sequencing

Genomic DNA was extracted from PBMCs using Phenol-Chloroform-Isoamyl alcohol. STX11 genomic region flanking the mutation was PCR amplified using primers: 5’ Forward -CATGCACGACTACAACCAGGC and 3’ Reverse - GGGACAGCAGAAGCAGCAGAGGG. The resulting PCR products were separated on 2% agarose gel, excised and extracted using the Macherey-Nagel Nucleospin columns and subjected to Sanger sequencing using the 5’ Forward PCR primer to confirm the mutation.

Measurement of SOCE in human PBMCs

PBMCs in culture were washed, rested and allowed to adhere to freshly carbon-coated glass bottom (IBDI) dishes for 1 hour in plain RPMI at 37°C. The cells were washed with HBSS and incubated with 1 ml of 1 μM Fura-2-AM diluted in Ringer’s buffer for 40 minutes, washed and incubated for an additional 10 minutes to allow de-esterification of the dye. The assay was started with 1 ml of Calcium free Ringer’s buffer in the imaging dish and images were captured every 4 seconds. After capturing the baseline for ∼60 seconds, stores were depleted by the addition of 1 μM Thapsigargin. In other assays, 10 μg/ml of anti-CD3 and 5 μg/ml of the secondary antibody were used to cross-link the TCRs and thereby induce store depletion. ∼5 minutes post-store-depletion, Calcium Chloride (CaCl2) was added back to the dish at a final concentration of 2 mM to estimate the magnitude of store operated calcium entry.

Degranulation assay

PBMCs were cultured in RPMI containing 10% FBS and 50 ng/ml IL-2 and stimulated with PHA (2 μg /ml) 48hrs before the assay. 24hrs prior to the assay, IL-2 was washed off but PHA was re-added. On the day of the assay cells were washed twice to remove PHA and any growth factors. To measure degranulation, cells were either left unstimulated, stimulated with a combination of anti-CD3 (10 μg/ml plate-coated), anti-CD28 (2 μg/ml soluble) and anti-CD49d (2 μg/ml soluble) or a combination of Ionomycin (1 μM) and PMA (50 ng/ml) for 3.5 hours. CD107a-PE antibody (1:50 dilution) was added to each of the three groups at the start of stimulation. Following stimulations, cells were transferred to ice, washed with cold HBSS containing 2% FBS and incubated with anti-CD8 APC for 20 minutes, washed twice with cold HBSS and analyzed using Cytoflex (Beckman Coulter) flow cytometer.

qPCR to estimate gene expression in human PBMCs

PBMCs were taken off IL-2 48 hours prior to the assay. On the day of the assay, cells were washed twice, left unstimulated and either stimulated with anti-CD3 (10 μg/ml, plate-coated), anti-CD28 (2 μg/ml) and anti-CD49d (2 μg/ml) or with Ionomycin (1 μM) and PMA (50 ng/ml) for 6 hours. To end the stimulation, cells were washed with cold HBSS, pelleted and used for RNA extraction. Total RNA was isolated using Qiagen RNeasy mini kit and cDNA was synthesized using random hexamers and Superscript IV (Invitrogen). cDNA was quantified using Qubit and subjected to QPCR analysis using Taqman probes for IL-2 and beta-actin housekeeping gene control in triplicates. Analysis was performed by calculating the double delta Ct values.

Transduction of human PBMCs with pMSCV-STX11

For viral transduction of PBMCs, polystyrene non-TC treated 24-well plates were coated with retronection (20 μg/ml) overnight at 4°C, blocked with 2% BSA for 30 minutes at room temperature and washed twice with HBSS. The retroviral supernatants were added onto the coated wells and the plates were spun at 1800g for 2 hours at 30°C. Following spin, the wells were washed with the blocking solution. PBMCs cultured in RPMI containing 10% FBS were stimulated with PHA (2 μg/ml) and IL-2 (50 ng/ml) for 48 hours prior to transduction. Stimulated PBMCs were transferred to coated plates at a density of 0.25 million cells/well centrifuged at 400g for 40 minutes at 30°C. The cells were analyzed 48-72 hours post transduction.

Patch clamp measurements

Patch clamp recordings were performed by using an Axopatch 200B amplifier (Axon Instruments, Foster City, CA) interfaced to an ITC-18 input/output board (Instrutech, Port Washington, NY) and an iMac G5 computer. Currents were filtered at 1 kHz with a four-pole Bessel filter and sampled at 5 kHz. Recording electrodes were pulled from 100-µl pipettes (VWR), coated with Sylgard, and fire-polished to a final resistance of 2 to 5 MΩ. Stimulation and data acquisition and analysis were performed by using in-house routines developed by RS Lewis (Stanford University) on the Igor Pro platform (Wavemetrics, Lake Oswego, OR). All data were corrected for the liquid junction potential of the pipette solution relative to Ringer’s in the bath (−10 mV) and for leak currents collected in 20 mM extracellular free calcium plus 10µM La3+. The holding potential was +30 mV. The standard voltage stimulus consisted of a 100-ms step to –100 mV followed by a 100-ms ramp from –100 to +100 mV applied at 1 s intervals. ICRAC was typically activated by passive depletion of ER Ca2+ stores by intracellular dialysis of 8 mM BAPTA.

Solutions and chemicals for patch clamp

The standard extracellular Ringer solution contained 130 mM NaCl, 4.5 mM KCl, 20 mM CaCl2, 10 mM tetraethylammoniumchloride (TEA-Cl), 10 mM D -glucose and 5 mM HEPES (pH 7.4 with NaOH). The standard divalent-free (DVF) Ringer solutions contained (in mM) 150 NaCl, 10 HEDTA, 1 EDTA, and 10 HEPES (pH 7.4). The internal solution contained: 135 mM Cs aspartate, 8 mM MgCl2, 8 mM Cs-BAPTA and 10 mM HEPES (pH 7.2 with CsOH).

Patch clamp data analysis

Analysis of current amplitudes was typically performed by measuring the peak currents during the −100 mV pulse. Averaged results are presented as the mean value standard error of the mean (SEM). Reversal potentials were measured from the average of several leak-subtracted sweeps (4–6) in each cell. For datasets with two groups, statistical analysis was performed with two-tailed t test to compare between control and test conditions with a confidence level of 95%, and results with p<0.05 were considered statistically significant.

STX11:Orai1 and SOAR:Orai complex prediction

Full length STX11 and its two domains, (Habc 41-167 and SNARE 183-267) and Orai1 (N-terminus 1-87 and C-terminus 256-301) were used to generate the complex of STX11 and Orai1 using AlphaFold3 (AF3) in all combinations. For Stim1, SOAR, residue 344-444 were used along with Orai C-terminus 256-301. The resultant models were assessed using ipTM and pTM scores. The best scoring combinations were further considered to generate more models by changing the seed values. The best model in terms of highest ipTM and pTM values was considered for further analysis. Custom script was used to analyze the contact frequency of interacting residues of STX11-Habc and Orai1 C-terminus complex as well as SOAR and Orai1 C-terminus, across all predicted models.

Molecular dynamics (MD) simulation

The STX11 Habc:Orai1 C-terminus and SOAR:Orai1 C-terminus complexes were subjected to all-atom MD simulations. Initially, the complexes were prepared using protein-preparation module of Schrodinger which involves H-bond network optimization and restrained minimization of the initial structure. The prepared structures were solvated using TIP3P water model in an orthorhombic water box and neutralized by adding counter ions. OPLS4 force field was used, and simulation system was generated by specifying 150 mM salt (NaCl). A buffer distance of 10 Å beyond the solute in each direction was used. The solvated system was subjected to default relaxation protocol of Desmond followed by production run for 500 ns at 300K and 1 atm pressure in NPT ensemble. The default relaxation protocol includes several short simulation steps. (1) Brownian dynamics simulation for 100 ps at 10 K temperature in NVT with restraints on solute heavy atoms (2) Simulation in NVT ensemble at 10 K for 12 ps with restraints on solute heavy atoms (3) 12 ps simulation in NPT ensemble at 10 K with restraints on solute heavy atoms (4) Simulation in NPT ensemble for 12 ps with restraints on solute heavy atoms (5) Simulation in NPT ensemble for 24 ps without restraints. Three independent runs were executed with different initial seeds for both structures.

Analysis of MD simulation

The MD runs were analyzed for the stability of the complex and interactions between the subunits. Simulation stability was assessed based on the protein backbone RMSD over simulation time. The RMSD was calculated using simulation interaction diagram (SID) module of Schrodinger. The first frame was used as reference frame for RMSD calculation. Script “analyze_trajectory_ppi.py” was used to calculate the interactions between STX11:Orai1 and Stim1:Orai1. The binding energy (ΔG) at each nanosecond across simulation was calculated using “thermal_mmgbsa.py” script of Schrodinger. The trajectories of the three runs of STX11:Orai1 and Stim1:Orai1 interactions have been deposited in Zenodo (https://doi.org/10.5281/zenodo.21663510). The representative structure was generated through trajectory clustering using the trajectory clustering tool in Schrödinger, which applies affinity propagation to the pairwise backbone RMSD-based similarity matrix. Within each affinity propagation run, convergence was defined as no change in the set of exemplar frames for 15 consecutive iterations, with a maximum of 400 iterations per run. If convergence was not reached, the damping factor was increased from 0.5 in increments of 0.01 until convergence. The last 200 ns frames from each run were combined and clustered using scripts “trj_merge.py” and “trajectory cluster” of Schrodinger. The cluster representative of the largest cluster was used to analyze the protein-protein interactions. PDB Sum was used to calculate the interactions in the cluster representative. Figures were generated using Pymol and MD movies using Maestro.

Statistical Analysis

Statistical analysis was done using unpaired two-tailed Student’s t test where only two groups were being compared and ANOVA where more than two groups were present. Normality of distribution was checked in Origin. Mann Whitney U Test was used for datasets with non-normal (or skewed) distribution. P values for all significance tests were *P<0.05; **P<0.01; ***P<0.001.